---
title: "Microplastic fingerprinting"
output: 
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
# Set root directory to the project root (Project folder)
# This assumes the .Rmd is in the root of the project.
knitr::opts_knit$set(root.dir = getwd())
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(out.width='750px', dpi=200)
```

## Documentation

This repo is accompanying the publication: "Chemical Fingerprints of New vs. Weathered Microplastics: A Machine Learning Approach"

## Loading Packages and Functions
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# if packages are not install then install
required_packages <- c("tidyverse", "readxl", "writexl", "data.table", "crayon", "grid", "gridExtra", "stats", "vegan", "cluster", "factoextra", "FactoMineR", "ggforce", "MASS", "caret", "randomForest", "ranger", "pheatmap", "viridis", "missForest", "pROC", "VIM", "softImpute", "ggpubr", "RANN", "foreach", "doParallel", "parallel")

new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) {
  install.packages(new_packages)
}

# #########################
##   Load Libraries    ##
#########################
library(tidyverse) # Loads dplyr, tidyr, purrr, ggplot2, readr, tibble
library(readxl)
library(writexl)
library(data.table)
library(crayon)
library(grid)
library(gridExtra)
library(stats)
library(vegan)
library(cluster)
library(factoextra)
library(FactoMineR)
library(ggforce)
library(MASS)
library(caret)
library(randomForest)
library(ranger)
library(pheatmap)
library(viridis)
library(missForest)
library(pROC)
library(VIM)
library(softImpute)
library(ggpubr)
library(RANN)
library(foreach)
library(doParallel)
library(parallel)

# Load Helper Functions
# Ensure the helper script is in the same folder as this Rmd
helper_file <- "Helper Functions for Computational Workflow.R"
if(file.exists(helper_file)) {
  source(helper_file)
} else {
  warning(paste("CRITICAL WARNING: Helper file '", helper_file, "' not found in working directory:", getwd()))
}

# --- OPTIMIZATION: Helper function for Stats Reporting in Step 6 ---
print_filtering_stats <- function(results, meta_cols = c("Plastic_type", "technique", "Source", "Polymer")) {
  
  df_final <- results$filtered_data
  df_orig  <- results$original_data
  
  # Get list of feature names only
  feats_final <- setdiff(names(df_final), meta_cols)
  feats_orig  <- setdiff(names(df_orig), meta_cols)
  
  # 1. Percentage of chemical features retained (%)
  pct_features <- (length(feats_final) / length(feats_orig)) * 100
  cat(paste0("1. Percentage of chemical features retained: ", round(pct_features, 2), "%\n"))
  
  # 2. Percentage of missing values after feature reduction (%)
  if(length(feats_final) > 0) {
    vals_final <- df_final[, feats_final, drop = FALSE]
    pct_missing <- (sum(is.na(vals_final)) / prod(dim(vals_final))) * 100
    cat(paste0("2. Percentage of missing values after feature reduction: ", round(pct_missing, 2), "%\n"))
  } else {
    cat("2. Percentage of missing values after feature reduction: 0% (No features left)\n")
    vals_final <- data.frame(0) # Placeholder for signal calc
  }
  
  # 3. Percentage of signal response after feature reduction (%)
  signal_final <- sum(vals_final, na.rm = TRUE)
  signal_orig  <- sum(df_orig[, feats_orig, drop = FALSE], na.rm = TRUE)
  pct_signal   <- (signal_final / signal_orig) * 100
  cat(paste0("3. Percentage of signal response after feature reduction: ", round(pct_signal, 2), "%\n"))
}
```

## Step 1.1: Data import
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# ATDGCMS -----------------------------------------------------------------
# Using relative path
path_atdgcms <- "./Raw data/ATDGCMS"

if(!dir.exists(path_atdgcms)) warning(paste("Directory not found:", path_atdgcms))

atdgcms_list <- list.files(path = path_atdgcms, pattern = '*.csv', full.names = TRUE) %>%
  .[!str_detect(., "2022")] %>%
  # remove USE samples with only 1 replication or specific exclusions
  .[!str_detect(., "USE_(01|02|05|06|09|10|11|12|13|15|16|17|18|19|20)")] %>%
  .[!str_detect(., "USSB_(01|08)")]

# Blank samples 
blank_list <- list.files(path = path_atdgcms, pattern = '*.csv', full.names = TRUE) %>%
  .[str_detect(., "2022-")]

# Import samples to list
read_and_add_filename <- function(file) {
  data.table::fread(
    file,
    showProgress = FALSE,
    data.table = FALSE,
    check.names = TRUE
  ) %>%
    mutate(File = basename(file))
}

atdgcms_step1.1 <- data.table::rbindlist(
  lapply(atdgcms_list, read_and_add_filename),
  use.names = TRUE,
  fill = TRUE
) %>%
  as.data.frame() %>%
  dplyr::select(-any_of(c("Start", "End", "Width", "Base.Peak", "Cpd", "Label", "Height", "Ions"))) %>%
  dplyr::mutate(File = gsub("_", "-", File)) %>%
  dplyr::mutate(File = gsub(".csv", "", File)) %>%
  mutate(type = "Sample")

atdgcms_blank <- data.table::rbindlist(
  lapply(blank_list, function(path) {
    data.table::fread(path, showProgress = FALSE, check.names = TRUE)
  }),
  use.names = TRUE,
  fill = TRUE
) %>%
  as.data.frame() %>%
  dplyr::select(any_of(c("Area", "File", "m.z", "RT"))) %>%
  mutate(type = "Blanks")

atdgcms_step1.1 <- bind_rows(atdgcms_step1.1, atdgcms_blank) %>% 
  arrange(RT) %>%
  mutate(simplified_file = map_chr(File, ~ paste(str_split_1(.x, pattern = "-")[c(4, 5)], collapse = "-")))

# HPLCTOFMS ---------------------------------------------------------------

# Define relative paths for batches
path_batch0 <- "./Raw data/HPLCTOFMS/EF_Non-target data_Batch 0"
path_batch1 <- "./Raw data/HPLCTOFMS/EF_Non-target data_Batch 1"
path_batch2 <- "./Raw data/HPLCTOFMS/EF_Non-target data_Batch 2"

# Function to safely list files and read them
read_hplc_batch <- function(dir_path, batch_num) {
  
  if(!dir.exists(dir_path)) {
    warning(paste("Directory not found:", dir_path))
    return(data.frame())
  }
  
  cache_dir <- "./data_cache"
  if (!dir.exists(cache_dir)) dir.create(cache_dir, recursive = TRUE)
  cache_file <- file.path(cache_dir, paste0("hplc_batch_", batch_num, "_skip1.rds"))
  
  files_all <- list.files(path = dir_path, pattern = '*.xls', full.names = TRUE)
  if (length(files_all) == 0) return(data.frame())
  
  file_info <- file.info(files_all)
  cur_sig <- data.frame(
    path = normalizePath(files_all, winslash = "/", mustWork = FALSE),
    mtime = as.numeric(file_info$mtime),
    size = as.numeric(file_info$size),
    stringsAsFactors = FALSE
  )
  
  if (file.exists(cache_file)) {
    cached <- readRDS(cache_file)
    if (is.list(cached) && is.data.frame(cached$files) && identical(cached$files, cur_sig)) {
      return(cached$data)
    }
  }
  
  # Get Sample Files
  files_sample <- files_all[!str_detect(files_all, "Blank")]
  
  # Get Blank Files
  files_blank <- files_all[str_detect(files_all, "Blank")]
  
  # Read and combine
  read_xls_fast <- function(path) {
    readxl::read_xls(path, skip = 1) %>% as.data.frame()
  }
  
  combined <- data.table::rbindlist(
    c(lapply(files_sample, read_xls_fast),
      lapply(files_blank, read_xls_fast)),
    use.names = TRUE,
    fill = TRUE
  ) %>%
    as.data.frame() %>%
  dplyr::select(m.z = `m/z`, RT, Height, File) %>%
  mutate(File = gsub("_", "-", File),
         type = ifelse(str_detect(File, "Blank"), "Blanks", "Sample"),
         Day = case_when(str_detect(File, "Day0") ~ "0",
                         str_detect(File, "Day2") ~ "2",
                         str_detect(File, "Day15") ~ "15",
                         TRUE ~ "34"),
         Replicate = ifelse(str_detect(File, "R1"), "R1", "R2"),
         Batch_number = as.character(batch_num))
  
  saveRDS(list(files = cur_sig, data = combined), cache_file)
  return(combined)
}

# Import Batches
hplc_batch0_step1.1 <- read_hplc_batch(path_batch0, 0) %>%
   mutate(simplified_file = map_chr(File, ~ paste(str_split_1(.x, "-")[1:2], collapse = "-")))

hplc_batch1_step1.1 <- read_hplc_batch(path_batch1, 1) %>%
   mutate(simplified_file = map_chr(File, ~ paste(str_split_1(.x, "-")[2:3], collapse = "-")))

hplc_batch2_step1.1 <- read_hplc_batch(path_batch2, 2) %>%
   mutate(simplified_file = map_chr(File, ~ paste(str_split_1(.x, "-")[2:3], collapse = "-")))

# Combine All using data.table::rbindlist() (much faster than bind_rows)
all_hplc <- data.table::rbindlist(list(hplc_batch0_step1.1,
                                        hplc_batch1_step1.1,
                                        hplc_batch2_step1.1), use.names = TRUE, fill = TRUE) %>% arrange(RT)
```

## Step 1.2: Collapsing compounds based on RT1, RT2, Ion1 threshold

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# ATDGCMS
atdgcms_grouped <- grouping_comp(atdgcms_step1.1,
                                 rtthres1 = 0.005,
                                 mzthres = 0.05,
                                 type = "ATDGCMS")

# HPLCTOFMS
# Apply different alignment to 2 parts of chromatogram via rtthres1 (RT <= 6 min) and rtthres2 (RT > 6 min)
hplc_batch_grouped <- grouping_comp(all_hplc,
                                    split_time = 6,
                                    rtthres1 = 0.4,
                                    rtthres2 = 0.28,
                                    mzthres = 0.0005,
                                    type = "HPLCTOFMS")
```

## Step 1.3: Remove all the benchmark compounds for HPLC (4) and GC (1 - Toluene)

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# ATDGCMS
idx <- which(atdgcms_grouped$m.z >= 98 & 
               atdgcms_grouped$m.z <= 98.5 & 
               atdgcms_grouped$RT > 13.4 & 
               atdgcms_grouped$RT < 13.5)
if(length(idx) > 0) atdgcms_grouped <- atdgcms_grouped[-idx,]

# HPLC
hplc_benchmark_mz <- c(156.0957, 198.1723, 246.2568, 305.1956, 342.1722)
hplc_benchmark_idx <- c()
for (mz in hplc_benchmark_mz) {
  # Vectorized check for range
  idx <- which(abs(hplc_batch_grouped$m.z - mz) <= 0.0005)
  hplc_benchmark_idx <- c(hplc_benchmark_idx, idx)
}
if(length(hplc_benchmark_idx) > 0) hplc_batch_grouped <- hplc_batch_grouped[-hplc_benchmark_idx,]
```


## Step 2: Blank subtraction & Remove all detections with negative peak area

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# ATDGCMS Optimization
atdgcms_adjusted_step3 <- atdgcms_grouped %>%
  dplyr::group_by(Feature) %>%
  dplyr::mutate(
    # Check if this feature has any blanks
    has_blank = any(type == "Blanks"),
    # Calculate blank stats (returns NA if no blanks, handled by check)
    blank_mean = mean(Area[type == "Blanks"], na.rm = TRUE),
    blank_sd   = sd(Area[type == "Blanks"], na.rm = TRUE)
  ) %>%
  ungroup() %>%
  # Filter: Keep samples (discard blanks)
  dplyr::filter(type != "Blanks") %>%
  # Apply subtraction logic
  dplyr::mutate(
    Area = if_else(has_blank, Area - blank_mean, Area),
    # If no blank exists, treat threshold as 0 (always keep positive)
    threshold = if_else(has_blank, 3 * blank_sd, 0)
  ) %>%
  # Keep only positive areas above threshold
  dplyr::filter(Area > 0, Area > threshold) %>%
  dplyr::select(-has_blank, -blank_mean, -blank_sd, -threshold)

# HPLCTOFMS Optimization (Group by Batch AND Feature)
hplc_adjusted_step3 <- hplc_batch_grouped %>%
  dplyr::group_by(Batch_number, Feature) %>%
  dplyr::mutate(
    has_blank = any(type == "Blanks"),
    blank_mean = mean(Height[type == "Blanks"], na.rm = TRUE),
    blank_sd   = sd(Height[type == "Blanks"], na.rm = TRUE)
  ) %>%
  ungroup() %>%
  dplyr::filter(type != "Blanks") %>%
  dplyr::mutate(
    Height = if_else(has_blank, Height - blank_mean, Height),
    threshold = if_else(has_blank, 3 * blank_sd, 0)
  ) %>%
  dplyr::filter(Height > 0, Height > threshold) %>%
  dplyr::select(-has_blank, -blank_mean, -blank_sd, -threshold)
```

# Step 3: Add info on Source and rename columns
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# ATDGCMS
gc <- atdgcms_adjusted_step3 %>%
  dplyr::filter(!is.na(Feature)) %>%
  dplyr::select(File, Feature, m.z, RT, Area) %>%
  mutate(Source = ifelse(str_detect(File, "USE"), "Environmental", "Store-Bought")) %>%
  dplyr::rename(Values = Area) %>%
  mutate(technique = "GC")

# HPLCTOFMS
hplc <- hplc_adjusted_step3 %>% 
  dplyr::filter(!is.na(Feature)) %>%
  dplyr::select(File, Feature, m.z, RT, Height) %>%
  dplyr::mutate(Source = ifelse(str_detect(File, "USE"), "Environmental", "Store-Bought")) %>%
  dplyr::rename(Values = Height) %>%
  mutate(technique = "HPLC")
```

## Step 4: Importing Trace metal data

#### Data prepping

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# Helper to read excel safely
read_trace_metal <- function(path) {
  if(!file.exists(path)) stop(paste("File missing:", path))
  readxl::read_excel(path) %>% column_to_rownames(var = "File")
}

# Trace metal data
icpms_round1 <- read_trace_metal("./Raw data/ICPMS_Trace metal/icpms_round1_rawdata_removal USE-01-rep1 and USE-03 (only2observations).xlsx")
icpms_round2_batch1 <- read_trace_metal("./Raw data/ICPMS_Trace metal/icpms_round2_batch1_rawdata.xlsx")
icpms_round2_batch2 <- read_trace_metal("./Raw data/ICPMS_Trace metal/icpms_round2_batch2_rawdata.xlsx")

# 1) Select [H2] instead of [H2 HMI]
remove_H2_HMI_modus <- function(icpms_df) {
  strings <- colnames(icpms_df)
  pattern <- "^(.*\\[ )([A-Za-z0-9 ]+)( \\])$"
  
  # Filter out H2 HMI
  keep_cols <- strings[!str_detect(str_match(strings, pattern)[,3], "H2 HMI")]
  icpms_df %>% dplyr::select(all_of(keep_cols))
}

icpms_round1 <- remove_H2_HMI_modus(icpms_round1)
icpms_round2_batch1 <- remove_H2_HMI_modus(icpms_round2_batch1)
icpms_round2_batch2 <- remove_H2_HMI_modus(icpms_round2_batch2)

# 2) Process Uncertainty/Recovery
# Load auxiliary files
rec_unc_1 <- readxl::read_excel("./Raw data/ICPMS_Trace metal/Trace metal data_Recovery_Uncertainty_round1.xlsx")
rec_unc_2_1 <- readxl::read_excel("./Raw data/ICPMS_Trace metal/Trace metal data_Recovery_Uncertainty_round2_batch1.xlsx")
rec_unc_2_2 <- readxl::read_excel("./Raw data/ICPMS_Trace metal/Trace metal data_Recovery_Uncertainty_round2_batch2.xlsx")

# Function to select best method per element
get_final_metal_names <- function(recovery_uncertainty, target_cols) {
  # Filter to columns actually present in the data
  recovery_uncertainty <- recovery_uncertainty %>% filter(`Element+Modus` %in% target_cols)
  
  first_numbers <- sub(" .*", "", recovery_uncertainty$`Element+Modus`)
  
  # Identify duplicates
  dupes <- first_numbers[duplicated(first_numbers)] %>% unique()
  
  kept_names <- c()
  
  # Process all unique elements
  for(num in unique(first_numbers)) {
    rows <- recovery_uncertainty[first_numbers == num, ]
    if(nrow(rows) == 1) {
      kept_names <- c(kept_names, rows$`Element+Modus`)
    } else {
      # Selection logic: Max Recovery, else Min Uncertainty
      best_row <- rows[which.max(rows$`Recovery [%]`), ]
      if(nrow(best_row) == 0) best_row <- rows[which.min(rows$`U(k=2)`), ]
      kept_names <- c(kept_names, best_row$`Element+Modus`)
    }
  }
  return(kept_names)
}

final_names_1 <- get_final_metal_names(rec_unc_1, colnames(icpms_round1))
final_names_2_1 <- get_final_metal_names(rec_unc_2_1, colnames(icpms_round2_batch1))
final_names_2_2 <- get_final_metal_names(rec_unc_2_2, colnames(icpms_round2_batch2))

# Subset
final_icpms_round1 <- icpms_round1 %>% dplyr::select(all_of(final_names_1))
final_icpms_round2_batch1 <- icpms_round2_batch1 %>% dplyr::select(all_of(final_names_2_1))
final_icpms_round2_batch2 <- icpms_round2_batch2 %>% dplyr::select(all_of(final_names_2_2))

# Combine
common_cols <- Reduce(intersect, list(names(final_icpms_round1), names(final_icpms_round2_batch1), names(final_icpms_round2_batch2)))
combined_icpms <- rbind(final_icpms_round1[common_cols], final_icpms_round2_batch1[common_cols], final_icpms_round2_batch2[common_cols])

# 4) Discard Seawater Metals
strings_to_remove <- c("Na", "Mg", "K", "Ca")
cols_to_remove <- unique(unlist(sapply(strings_to_remove, function(x) grep(x, colnames(combined_icpms)))))
if(length(cols_to_remove) > 0) combined_icpms <- combined_icpms[, -cols_to_remove]

icp <- combined_icpms %>%
  rownames_to_column(., var = "File")  %>% 
  pivot_longer(., cols = 2:ncol(.), names_to = "Feature", values_to = "Values") %>%
  mutate(
    technique = "ICP",
    Source = ifelse(str_detect(File, "USE"), "Environmental", "Store-Bought"),
    Values = as.numeric(Values)
  )
```

## Step 5: Combine Label Info with raw data

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
p <- "./Raw data/Sample Labelling_all data_GC+HPLC+ICP_30Sept2025.xlsx"
if(!file.exists(p)) stop(paste("Label file not found:", p))

cache_dir <- "./data_cache"
cache_file <- file.path(cache_dir, "step5_combined_labeled.rds")
use_step5_cache <- TRUE
loaded_from_cache <- FALSE

if (use_step5_cache && file.exists(cache_file)) {
  cached <- readRDS(cache_file)
  cached_mtime <- cached$meta$label_mtime
  current_mtime <- as.numeric(file.info(p)$mtime)
  
  if (is.list(cached) &&
      is.list(cached$data) &&
      isTRUE(all.equal(cached_mtime, current_mtime))) {
    gc <- cached$data$gc
    hplc <- cached$data$hplc
    icp <- cached$data$icp
    gc_hplc <- cached$data$gc_hplc
    gc_icp <- cached$data$gc_icp
    hplc_icp <- cached$data$hplc_icp
    gc_hplc_icp <- cached$data$gc_hplc_icp
    sampinfo <- cached$data$sampinfo
    loaded_from_cache <- TRUE
  }
  
  rm(cached)
}

if (!loaded_from_cache) {
  sampinfo <- readxl::read_excel(path = p) %>% 
    dplyr::select(c(File, Plastic_type, Subcategory, Polymer))
  
  # Join Info
  icp <- left_join(icp, sampinfo, by = 'File')
  gc <- left_join(gc, sampinfo, by = 'File')
  hplc <- left_join(hplc, sampinfo, by = 'File')
  
  # Create Combinations
  gc_hplc <- bind_rows(gc, hplc)
  gc_icp <- bind_rows(icp, gc %>% dplyr::select(-c("m.z", "RT"))) 
  hplc_icp <- bind_rows(icp, hplc %>% dplyr::select(-c("m.z", "RT"))) 
  gc_hplc_icp <- bind_rows(icp, gc_hplc %>% dplyr::select(-c("m.z", "RT")))
  
  if (!dir.exists(cache_dir)) dir.create(cache_dir, recursive = TRUE)
  cache_payload <- list(
    meta = list(label_file = p, label_mtime = as.numeric(file.info(p)$mtime)),
    data = list(
      gc = gc,
      hplc = hplc,
      icp = icp,
      gc_hplc = gc_hplc,
      gc_icp = gc_icp,
      hplc_icp = hplc_icp,
      gc_hplc_icp = gc_hplc_icp,
      sampinfo = sampinfo
    )
  )
  saveRDS(cache_payload, cache_file)
}
```


## Step 6: Feature Reduction

### Summary report of data metrics before applying feature reduction

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
df_in <- gc

# 1. Process Data ONCE (Consolidated Pivot)
df_wide <- df_in %>% 
  dplyr::select(File, Plastic_type, Source, Polymer, Feature, technique, Values) %>%
  dplyr::group_by(File, Source, Polymer, Plastic_type, technique, Feature) %>%
  dplyr::summarise(Values = mean(Values, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = Feature, values_from = Values) %>%
  tibble::column_to_rownames(var = "File") 

df_numeric <- df_wide %>% dplyr::select(where(is.numeric))

# --- Basic Statistics ---
cat("Number of chemical features: ", ncol(df_numeric), "\n")
missing_pct <- sum(is.na(df_numeric)) / prod(dim(df_numeric)) * 100
cat("Percentage of missing values (%):", round(missing_pct, 2), "\n")

# --- Correlations ---
stats_df <- tibble::tibble(
  Feature = colnames(df_numeric),
  num_missing_values = colSums(is.na(df_numeric)),
  mean_response = colMeans(df_numeric, na.rm = TRUE)
)

if (all(c("m.z", "RT") %in% colnames(df_in))) {
  feature_meta <- df_in %>%
    dplyr::group_by(Feature) %>%
    dplyr::summarise(
      m_z = mean(m.z, na.rm = TRUE),
      retention_time = mean(RT, na.rm = TRUE),
      .groups = "drop"
    )
  analysis_df <- stats_df %>% dplyr::left_join(feature_meta, by = "Feature")
  cat("Pearson (Mean Response vs Missing):", round(cor(analysis_df$mean_response, analysis_df$num_missing_values, use = "pairwise"), 2), "\n")
  cat("Pearson (m/z vs Missing):          ", round(cor(analysis_df$m_z, analysis_df$num_missing_values, use = "pairwise"), 2), "\n")
  cat("Pearson (RT vs Missing):           ", round(cor(analysis_df$retention_time, analysis_df$num_missing_values, use = "pairwise"), 2), "\n")
} else {
  cat("Pearson (Mean Response vs Missing):", round(cor(stats_df$mean_response, stats_df$num_missing_values, use = "pairwise.complete.obs"), 2), "\n")
  message("NOTE: 'm.z' and 'RT' columns missing. Skipping specific correlations.")
}
```

### Step 6.1 - Feature reduction Option 1: Regular 80% rule - Remove feature if there are more than 80% MVs

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
results <- shared_feature_filtering(gc, feat_reduc_mode = "regular80")
print_filtering_stats(results)
```

### Step 6.2 - Feature reduction Option 2: Modified 80% rule
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
results <- shared_feature_filtering(gc, feat_reduc_mode = "modified80")
print_filtering_stats(results)
```


### Step 6.3 - Feature reduction Option 3: In-house Comprehensive feature filtering 
```{r , echo=FALSE, warning = FALSE, message=FALSE}
results <- shared_feature_filtering(gc, feat_reduc_mode = "inhouse")
print_filtering_stats(results)
```

### Step 6.4 - Data fusion to create multi-instrumental datasets
```{r, echo = FALSE, message = FALSE, warning = FALSE}
# 1. Apply feature reduction to single analytical datasets
gc_clean   <- shared_feature_filtering(gc, feat_reduc_mode = "inhouse")$filtered_data
hplc_clean <- shared_feature_filtering(hplc, feat_reduc_mode = "inhouse")$filtered_data
icp_clean  <- shared_feature_filtering(icp, feat_reduc_mode = "inhouse")$filtered_data

# 2. Create data fusion combos (Fused + Filtered Data)
gc_hplc_fused     <- data_fusion(gc_hplc, source_feats = list(gc_clean, hplc_clean))
gc_icp_fused      <- data_fusion(gc_icp, source_feats = list(gc_clean, icp_clean))
hplc_icp_fused    <- data_fusion(hplc_icp, source_feats = list(hplc_clean, icp_clean))
gc_hplc_icp_fused <- data_fusion(gc_hplc_icp, source_feats = list(gc_clean, hplc_clean, icp_clean))

# 3. Quality Assurance Report (Example: GC-ICP)
# We need the Original Data (Unfiltered, Wide Format) to compare against Fused Data
# Helper 'process_and_filter' can generate this structure cleanly.
class_col <- "Plastic_type" # Define class column locally to avoid errors

# Use helper to get raw wide data (threshold=0 returns standard object with original_data)
# We use GC-ICP as the example for the report
temp_res <- process_and_filter(gc_icp, class_col = class_col, threshold = 0.8, filter_mode = "none")

# Create a results object for the stats printer
# filtered_data = The fused result (Step 2)
# original_data = The raw wide result (Step 3 temp_res)
fusion_results <- list(
    filtered_data = gc_icp_fused,
    original_data = temp_res$original_data
)

# Print Stats
cat("--- Quality Assurance Report for GC-ICP Data Fusion ---\n")
print_filtering_stats(fusion_results)
```


# Step 7: Random Forest

### Option 1: Use SB to predict ENV - Only select shared features between SB and ENV 
```{r class.source = 'fold-hide', echo=FALSE, warning = FALSE, message=FALSE}
# 1. Parallel Setup
num_cores <- max(1, parallel::detectCores(logical = FALSE) - 1)
cl <- parallel::makePSOCKcluster(num_cores)
doParallel::registerDoParallel(cl)
cat("Parallel processing registered with", num_cores, "cores.\n")

t0 <- Sys.time()
data_combinations <- list(
  gc = gc_clean,
  gc_icp = gc_icp_fused
)

# Store results
rf_results_SB_train <- list()

for(name in names(data_combinations)) {
  cat("Processing dataset:", name, "\n")
  
  rf_results_SB_train[[name]] <- run_rf_analysis_manuscript1(
    data = data_combinations[[name]],
    type_col = "Plastic_type", 
    remove_cols = c("Source", "Polymer", "technique"),
    train_proportion = 0.8,
    use_store_vs_environmental_split = TRUE,
    data_name = name, 
    group_for_significance = "Plastic_type", 
    do_pairwise_test = TRUE,
    do_top_importance_selection = TRUE,
    do_rfa = TRUE,
    do_impute_norm_screen = TRUE,
    seed = 123,
    min_sample_number = 1
  )
  gc() # Memory cleanup
}

t1 <- Sys.time()
print(paste0("Total run time: ", round(as.numeric(difftime(t1, t0, units = "secs")), 2), " seconds"))
stopCluster(cl)
registerDoSEQ()
```

### Option 2: Use Both SB and ENV to predict ENV

```{r class.source = 'fold-hide', echo=FALSE, warning = FALSE, message=FALSE}
# Re-init Parallel for safety
num_cores <- max(1, parallel::detectCores(logical = FALSE) - 1)
cl <- parallel::makePSOCKcluster(num_cores)
doParallel::registerDoParallel(cl)

t0 <- Sys.time()
data_combinations <- list(
  gc = gc_clean,
  gc_icp = gc_icp_fused
)

rf_results_SB_ENV_train <- list()

for(name in names(data_combinations)) {
  cat("Processing dataset:", name, "\n")
  
  rf_results_SB_ENV_train[[name]] <- run_rf_analysis_manuscript1(
    data = data_combinations[[name]],
    data_name = name, 
    use_source_split = TRUE,
    type_col = "Plastic_type", 
    remove_cols = c("Source", "Polymer", "technique"),
    group_for_significance = "Plastic_type",
    do_pairwise_test = TRUE,
    do_top_importance_selection = TRUE,
    do_rfa = TRUE,
    train_proportion = 0.8,
    do_impute_norm_screen = TRUE,
    seed = 123,
    min_sample_number = 1
  )
  gc()
}

t1 <- Sys.time()
print(paste0("Total run time: ", round(as.numeric(difftime(t1, t0, units = "secs")), 2), " seconds"))
stopCluster(cl)
registerDoSEQ()
```

## Step 8: HCA
```{r echo=FALSE, warning=FALSE, message=FALSE}
# Data prep for HCA
# Use result from Step 7 Option 2
target_obj <- rf_results_SB_ENV_train$gc
if(is.null(target_obj)) target_obj <- rf_results_SB_ENV_train[[1]]

# target_obj is a list of folds; use the first fold
fold_obj <- if (is.list(target_obj) && !is.null(target_obj$final_imp_norm_train)) {
  target_obj
} else {
  target_obj[[1]]
}
if (is.null(fold_obj)) stop("No folds found in rf_results_SB_ENV_train.")

df <- dplyr::bind_rows(
  fold_obj$final_imp_norm_train,
  fold_obj$final_imp_norm_test
) %>%
  dplyr::mutate(Plastic_type = gsub("..", " ", Plastic_type, fixed = TRUE)) %>%
  dplyr::mutate(Plastic_type = gsub(".", " ", Plastic_type, fixed = TRUE)) %>%
  dplyr::mutate(Plastic_type_Source = paste0(Plastic_type, "-", Source))

hc_df <- df %>% 
  dplyr::select(-any_of(c("Subcategory", "Source", "Plastic_type", "Polymer", "technique"))) %>% 
  dplyr::select(where(is.numeric)) %>%
  as.data.frame()

# --- Clean: remove bad rows & handle non-finite values ---
hc_df[!is.finite(as.matrix(hc_df))] <- 0
keep <- rowSums(hc_df, na.rm = TRUE) > 0
hc_df_clean <- hc_df[keep, , drop = FALSE] 
labels_clean <- df$Plastic_type_Source[keep] 

# --- Distance Calculation ---
use_bray <- !any(hc_df_clean < 0, na.rm = TRUE)

if (use_bray) {
  d <- vegan::vegdist(hc_df_clean, method = "manhattan")
  link <- "average"
} else {
  d <- dist(hc_df_clean, method = "euclidean")
  link <- "average"
  message("Detected negative values; using Euclidean instead of Bray-Curtis.")
}

if (any(!is.finite(d))) stop("Distance matrix still has non-finite entries.")

# --- Cluster & plot ---
hca_samp <- hclust(d, method = link)
cat("cophenetic correlation coefficient (CCC) from HCA is:", cor(d, stats::cophenetic(hca_samp)), "\n")

par(cex = 1)
plot(hca_samp, labels = labels_clean, hang = -1, main = "")
```
