---
title: "Microplastic fingerprinting"
output: 
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/huyng/OneDrive - Toronto Metropolitan University/Microplastic/Microplastic-Fingerprinting")
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(out.width='750px', dpi=200)
```

## Documentation

This repo is accompanying the publication: "Computational fingerprinting workflow for environmental source tracking of microplastic based on representative additives"

Demo of each ML algorithms is shown below.

## Loading Packages and Functions
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# #########################
##   Load Libraries    ##
#########################
library(ggplot2)
library(ranger)
library(foreach)
library(doParallel)
library(parallel)
library(vegan)
library(readxl)
library(tidyverse)
library(dplyr)
library(data.table)
library(writexl)
library(tidyr)
library(grid)
library(gridExtra)
library(stats)
library(FactoMineR)
library(factoextra)
library(ggforce)
library(cluster)
library(MASS)  # For mvrnorm
library(caret)
library(randomForest)
library(pheatmap)
library(viridis)
library(missForest)
library(pROC)
library(VIM)
library(softImpute)
library(purrr)
library(ggpubr)
library(missForest)
library(pROC)
library(VIM)
library(softImpute)
library(purrr)
library(ggpubr)
library(writexl)

# Functions -------------------------------------------------------------------------------------------------------
'%notin%' <- Negate('%in%')

# Grouping compounds based on RT1, RT2, and Ion1 - Version 1
grouping_comp <- function(data, split_time, rtthres1, rtthres2, mzthres, type) {
  
  # create empty list, each sub-list is a compound group with following criteria:
  # rtthres: RT threshold window
  # mzthres: mz threshold window
  dat <- data[ , , drop = FALSE]  # Shallow copy to avoid modification of original data
  
  # Initialize the compound column filled with NA values
  dat$Feature <- NA
  i <- 1
  j <- 1 # Counter for HPLC data only
  
  # Checkpoint for 'type' parameter
  if (!(type %in% c("ATDGCMS", "HPLCTOFMS"))) {
    stop("Invalid type specified. Must be either 'ATDGCMS' or 'HPLCTOFMS'")
  }

  # Loop through each row to group compounds
  if (type %in% "HPLCTOFMS") {
      # Split HPLC data into RT <= 6 min and RT > 6 min
      dat1 <- dat %>% filter(RT <= split_time)
      dat2 <- dat %>% filter(RT > split_time)
      
      for (row in 1:nrow(dat1)) {
        # filter data by index, ALWAYS DO THIS INSTEAD OF CREATE SUBSET DATAFRAME
        rt <- base::as.numeric(dat1[row, "RT"])
        mz <- base::as.numeric(dat1[row, "m.z"])
        
        idx1 <- which(
          dat1$RT <= (rt + rtthres1) & dat1$RT >= (rt - rtthres1) &
            dat1$m.z <= (mz + mzthres) & dat1$m.z >= (mz - mzthres) &
            is.na(dat1$Feature)
        )
        
        if (length(idx1) > 0) {
          dat1[idx1, "Feature"] <- paste0("Compound_RT1_", i, ".", type)
          i <- i + 1
        }  
      }
      
      for (row in 1:nrow(dat2)) {
        # filter data by index, ALWAYS DO THIS INSTEAD OF CREATE SUBSET DATAFRAME
        rt <- base::as.numeric(dat2[row, "RT"])
        mz <- base::as.numeric(dat2[row, "m.z"])
        
        idx2 <- which(
          dat2$RT <= (rt + rtthres2) & dat2$RT >= (rt - rtthres2) &
            dat2$m.z <= (mz + mzthres) & dat2$m.z >= (mz - mzthres) &
            is.na(dat2$Feature)
        )
        
        if (length(idx2) > 0) {
          dat2[idx2, "Feature"] <- paste0("Compound_RT2_", j, ".", type)
          j <- j + 1
        }  
      }
      
      dat <- rbind(dat1, dat2)
      
  } else {
    for (row in 1:nrow(dat)) {
      
      # filter data by index, ALWAYS DO THIS INSTEAD OF CREATE SUBSET DATAFRAME
      rt <- base::as.numeric(dat[row, "RT"])
      mz <- base::as.numeric(dat[row, "m.z"])
      
      idx <- which(
        dat$RT <= (rt + rtthres1) & dat$RT >= (rt - rtthres1) &
          dat$m.z <= (mz + mzthres) & dat$m.z >= (mz - mzthres) &
          is.na(dat$Feature)
      )
      
      if (length(idx) > 0) {
        dat[idx, "Feature"] <- paste0("Compound_", i, ".", type)
        i <- i + 1
      }  
    }
  }
  
  return(dat)
}

# Filtering similar and unique compound 
# comp_filter <- function(data) {
#   # Get all unique feature groups once
#   unique_comp_groups <- unique(data$Feature)
#   
#   # Preallocate lists for indices
#   all_other_compounds_idx <- vector("list", length(unique_comp_groups))
#   all_unique_compounds_idx <- vector("list", length(unique_comp_groups))
# 
#   # Loop over unique compound groups, but avoid appending to vectors
#   for (i in seq_along(unique_comp_groups)) {
#     comp_grp <- unique_comp_groups[i]
#     
#     # Use vectorized comparison instead of grepl with fixed=TRUE
#     idx <- which(data$Feature == comp_grp)
#     if (length(unique(data[idx,]$File)) < 2) {
#       all_unique_compounds_idx[[i]] <- idx
#     } else {
#       all_other_compounds_idx[[i]] <- idx
#     }
#   }
#   
#   # Unlist and remove empty elements (NULLs)
#   all_other_compounds_idx <- unlist(all_other_compounds_idx, use.names = FALSE)
#   all_unique_compounds_idx <- unlist(all_unique_compounds_idx, use.names = FALSE)
#   
#   return(list(all_other_compounds_idx, all_unique_compounds_idx))
# }

# comp_filter_opt <- function(data) {
#   # Convert the data frame to a data.table
#   dt <- as.data.table(data)
#   
#   # For each feature, compute the number of unique File values
#   dt[, n_files := uniqueN(File), by = Feature]
#   
#   # Extract the indices based on the unique file count
#   all_unique_compounds_idx <- which(dt$n_files < 2)
#   all_other_compounds_idx  <- which(dt$n_files >= 2)
#   
#   return(list(all_other_compounds_idx = all_other_compounds_idx, 
#               all_unique_compounds_idx  = all_unique_compounds_idx))
# }
```

## STEP 1.1: Data import

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# ATDGCMS
setwd("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/ATDGCMS")

atdgcms_list <- list.files(pattern = '*.csv', full.names = TRUE) %>%
  .[!str_detect(., "2022")] %>%
  # remove USE samples with only 1 replication
  .[!str_detect(., "USE_01")] %>%
  .[!str_detect(., "USE_02")] %>%
  .[!str_detect(., "USE_05")] %>%
  .[!str_detect(., "USE_06")] %>%
  .[!str_detect(., "USE_09")] %>%
  .[!str_detect(., "USE_10")] %>%
  .[!str_detect(., "USE_11")] %>%
  .[!str_detect(., "USE_12")] %>%
  .[!str_detect(., "USE_13")] %>%
  .[!str_detect(., "USE_15")] %>%
  .[!str_detect(., "USE_16")] %>%
  .[!str_detect(., "USE_17")] %>%
  .[!str_detect(., "USE_18")] %>%
  .[!str_detect(., "USE_19")] %>%
  .[!str_detect(., "USE_20")] %>%
  .[!str_detect(., "USSB_01")] %>%
  .[!str_detect(., "USSB_08")] # %>%
  # .[!str_detect(., "USSB_09")] # %>%
  # .[str_detect(., "_USSB")] #  environmental samples

# Blank samples 
blank_list <- list.files(pattern = '*.csv', full.names = TRUE) %>%
  .[str_detect(., "2022-")]

# Import samples to list
# Function to read each file and add a "File" column
read_and_add_filename <- function(file) {
  df <- read.csv(file) %>%
    mutate(File = basename(file))  # Add a "File" column with the file name
  return(df)
}

# Apply the function to all files and combine them into a single dataframe
atdgcms_step1.1 <- lapply(atdgcms_list, read_and_add_filename) %>%
  bind_rows() %>%  # Combine all dataframes into one
  # dplyr::bind_rows(purrr::map(atdgcms_list, read.csv)) %>%
  dplyr::select(-c("Start", "End", "Width", "Base.Peak", "Cpd", "Label", "Height", "Ions")) %>%
  # # Change all ATDGCMS file name with underscore-separated to hyphen-separated
  dplyr::mutate(File = gsub("_", "-", File)) %>%
  dplyr::mutate(File = gsub(".csv", "", File)) %>%
  mutate(type = "Sample")

atdgcms_blank <- dplyr::bind_rows(purrr::map(blank_list, read.csv)) %>%
  dplyr::select(c("Area", "File", "m.z", "RT")) %>%
  mutate(type = "Blanks")

atdgcms_step1.1 <- rbind(atdgcms_step1.1, atdgcms_blank) %>% 
  arrange(RT) %>%
  # mutate(Replicate = ifelse(str_detect(File, "R1"), "R1",
  #                           ifelse(str_detect(File, "R2"), "R2", "R3"))) %>%
  mutate(simplified_file = map_chr(File, ~ paste(str_split_1(.x, pattern = "-")[c(4, 5)], collapse = "-")))

# HPLCTOFMS

###### Import Batch 0
setwd("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/HPLCTOFMS/EF_Non-target data_Batch 0")

hplc_batch0 <- list.files(pattern = '*.xls') %>%
  .[!str_detect(., "Blank")]

# Blank samples 
blank_hplc_batch0 <- list.files(pattern = '*.xls') %>%
  .[str_detect(., "Blank")]

# Import samples to list
hplc_batch0_step1.1 <- dplyr::bind_rows(purrr::map(hplc_batch0, read_xls, skip = 1), 
                                        purrr::map(blank_hplc_batch0, read_xls, skip = 1)) %>%
  dplyr::select(m.z = `m/z`, RT, Height, File) %>%
  mutate(File = gsub("_", "-", File),
         type = ifelse(str_detect(File, "Blank"), "Blanks", "Sample"),
         Day = case_when(str_detect(File, "Day0") ~ "0",
                         str_detect(File, "Day2") ~ "2",
                         str_detect(File, "Day15") ~ "15",
                         TRUE ~ "34"),
         Replicate = ifelse(str_detect(File, "R1"), "R1", "R2"),
         Batch_number = "0",
         simplified_file = map_chr(File, ~ paste(str_split_1(.x, "-")[1:2], collapse = "-")))

colnames(hplc_batch0_step1.1)[1] <- "m.z"

###### Import Batch 1
setwd("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/HPLCTOFMS/EF_Non-target data_Batch 1")

# Import sample data
hplc_batch1 <- list.files(pattern = '*.xls') %>%
  .[!str_detect(., "Blank")]
# # Import Blank data 
blank_hplc_batch1 <- list.files(pattern = '*.xls') %>%
  .[str_detect(., "Blank")]

# Put sample and blank data together and create additional description columns
hplc_batch1_step1.1 <- dplyr::bind_rows(purrr::map(hplc_batch1, read_xls, skip = 1), 
                                        purrr::map(blank_hplc_batch1, read_xls, skip = 1)) %>%
  dplyr::select(m.z = `m/z`, RT, Height, File) %>%
  mutate(File = gsub("_", "-", File),
         type = ifelse(str_detect(File, "Blank"), "Blanks", "Sample"),
         Day = case_when(str_detect(File, "Day0") ~ "0",
                         str_detect(File, "Day2") ~ "2",
                         str_detect(File, "Day15") ~ "15",
                         TRUE ~ "34"),
         Replicate = ifelse(str_detect(File, "R1"), "R1", "R2"),
         Batch_number = "1",
         simplified_file = map_chr(File, ~ paste(str_split_1(.x, "-")[2:3], collapse = "-")))

# Change column name
colnames(hplc_batch1_step1.1)[1] <- "m.z"

###### Import Batch 2
setwd("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/HPLCTOFMS/EF_Non-target data_Batch 2")

hplc_batch2 <- list.files(pattern = '*.xls') %>%
  .[!str_detect(., "Blank")]
# Blank samples 
blank_hplc_batch2 <- list.files(pattern = '*.xls') %>%
  .[str_detect(., "Blank")]

hplc_batch2_step1.1 <- dplyr::bind_rows(purrr::map(hplc_batch2, read_xls, skip = 1), 
                                        purrr::map(blank_hplc_batch2, read_xls, skip = 1)) %>%
  dplyr::select(m.z = `m/z`, RT, Height, File) %>%
  mutate(File = gsub("_", "-", File),
         type = ifelse(str_detect(File, "Blank"), "Blanks", "Sample"),
         Day = case_when(str_detect(File, "Day0") ~ "0",
                         str_detect(File, "Day2") ~ "2",
                         str_detect(File, "Day15") ~ "15",
                         TRUE ~ "34"),
         Replicate = ifelse(str_detect(File, "R1"), "R1", "R2"),
         Batch_number = "2",
         simplified_file = map_chr(File, ~ paste(str_split_1(.x, "-")[2:3], collapse = "-")))

colnames(hplc_batch2_step1.1)[1] <- "m.z"

all_hplc <- rbind(hplc_batch0_step1.1,
                  hplc_batch1_step1.1,
                  hplc_batch2_step1.1) %>% arrange(RT)

## Stable Isotope data
stable_isotope <- read_xlsx(path = paste0(getwd(), "/data/Isotope/Mikroplastik_Results_2ndRound_KCN-1.xlsx")) %>% 
  tidyr::pivot_longer(cols = 4:5, names_to = "replicate", values_to = "Value") %>% 
  dplyr::mutate(Value = abs(Value))
```

## STEP 1.2: Collapsing compounds based on RT1, RT2, Ion1 threshold --------------------------

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# ATDGCMS
atdgcms_grouped <- grouping_comp(atdgcms_step1.1,
                                 rtthres1 = 0.005,
                                 mzthres = 0.05,
                                 type = "ATDGCMS")

# HPLCTOFMS
# Apply different alignment to 2 parts of chromatogram via rtthres1 (RT <= 6 min) and rtthres2 (RT > 6 min)
hplc_batch_grouped <- grouping_comp(all_hplc,
                                    split_time = 6,
                                    rtthres1 = 0.4,
                                    rtthres2 = 0.28,
                                    mzthres = 0.0005,
                                    type = "HPLCTOFMS")
```

## Step 1.3: Remove all the benchmark compounds for HPLC (4) and GC (1 - Toluene)
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
#ATDGCMS
idx <- which(atdgcms_grouped$m.z >= 98 & 
               atdgcms_grouped$m.z <= 98.5 & 
               atdgcms_grouped$RT > 13.4 & 
               atdgcms_grouped$RT < 13.5)
atdgcms_grouped <- atdgcms_grouped[-idx,]

# HPLC
hplc_benchmark_mz <- c(156.0957, 198.1723, 246.2568, 305.1956, 342.1722)
# Search for compound in combined_df2_grouped that fall in the window of benchmark' molecular ions
# Batch 0
hplc_benchmark_idx <- c()
for (mz in hplc_benchmark_mz) {
  idx <- which(hplc_batch_grouped$m.z <= (mz + 0.0005) & hplc_batch_grouped$m.z >= (mz - 0.0005))
  # collect all these indices of benchmark compounds
  hplc_benchmark_idx <- c(hplc_benchmark_idx, idx)
}
hplc_batch_grouped <- hplc_batch_grouped[-hplc_benchmark_idx,]
```


## Step 2: Blank subtraction & Remove all detections with negative peak area

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# ATDGCMS --------------
# Create list to store temp dfs
process_atdgcms <- function(atdgcms_grouped) {
  result_list <- list()
  idx <- 1
  
  # Iterate through each unique feature
  for (feature in unique(atdgcms_grouped$Feature)) {
    temp <- atdgcms_grouped[atdgcms_grouped$Feature == feature, ]
    
    # If no blanks exist for the current feature, add it as is
    if (identical(which(temp$type == "Blanks"), integer(0))) {
      result_list[[idx]] <- temp
      idx <- idx + 1
      next
    } else {
      # Calculate the mean and sd for blanks
      blank_idx <- which(temp$type == "Blanks")
      blank_mean <- base::mean(temp[blank_idx,]$Area, na.rm = TRUE)
      blank_sd   <- stats::sd(temp[blank_idx,]$Area, na.rm = TRUE)
      
      # Exclude blanks from further processing
      temp <- temp[temp$type != "Blanks", ]
      
      # Adjust the 'Area' for each sample within the feature
      for (sample in unique(temp$File)) {
        sample_idx <- which(temp$File == sample)
        temp[sample_idx,]$Area <- temp[sample_idx,]$Area - blank_mean
        
        # Keep only those entries with adjusted Area > 0 and > 3 * blank_sd
        positive_idx <- which(temp$Area > 0 & temp$Area > 3 * blank_sd)
        temp <- temp[positive_idx, ]
      }
    }
    
    # Append processed data frame for this feature
    result_list[[idx]] <- temp
    idx <- idx + 1
  }
   
  output <- dplyr::bind_rows(result_list)
  return(output)
}

# For positive values => combine data again to 1 grand data frame
atdgcms_adjusted_step3 <- process_atdgcms(atdgcms_grouped)

# HPLCTOFMS  --------------
hplc_result_list <- list()
idx <- 1

# Iterate through each batch
for (batch in unique(hplc_batch_grouped$Batch_number)) {
  # Iterate through each unique feature within each batch
  for (feature in unique(hplc_batch_grouped$Feature)) {
    temp <- hplc_batch_grouped[which(hplc_batch_grouped$Feature == feature & hplc_batch_grouped$Batch_number == batch), ]
    
    # If no blanks exist for the current feature, add it as is
    if (identical(which(temp$type == "Blanks"), integer(0))) {
      hplc_result_list[[idx]] <- temp
      idx <- idx + 1
      next
    } else { # if blanks exist for the current feature in the current batch number
      # Calculate the mean and sd for blanks
      blank_idx <- which(temp$type == "Blanks")
      blank_mean <- base::mean(temp[blank_idx,]$Height, na.rm = TRUE)
      blank_sd   <- stats::sd(temp[blank_idx,]$Height, na.rm = TRUE)
      
      # Exclude blanks from further processing
      temp <- temp[temp$type != "Blanks", ]
      
      # Adjust the 'Height' for each sample within the feature
      for (sample in unique(temp$File)) {
        sample_idx <- which(temp$File == sample)
        temp[sample_idx,]$Height <- temp[sample_idx,]$Height - blank_mean
        
        # Keep only those entries with adjusted Height > 0 and > 3 * blank_sd
        positive_idx <- which(temp$Height > 0 & temp$Height > (3 * blank_sd))
        temp <- temp[positive_idx, ]
      }
    }
    
    # Append processed data frame for this feature
    hplc_result_list[[idx]] <- temp
    idx <- idx + 1
  }
}

hplc_adjusted_step3 <- dplyr::bind_rows(hplc_result_list)
```

# Add info on Source and rename columns
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# ATDGCMS
# idx_list_filter1 <- comp_filter_opt(atdgcms_adjusted_step3)
gc <- atdgcms_adjusted_step3 %>%
  dplyr::filter(!is.na(Feature)) %>%
  dplyr::select(File, Feature, m.z, RT, Area) %>%
  mutate(Source = ifelse(str_detect(File, "USE"), "Environmental", "Store-Bought")) %>%
  dplyr::rename(Values = Area)

gc$technique <- "GC"

# HPLCTOFMS
# idx_list_filter2 <- comp_filter_opt(hplc_adjusted_step3)
hplc <- hplc_adjusted_step3 %>% # [idx_list_filter2[[1]],] %>%
  # dplyr::filter(Day %notin% c("0", "2")) %>%
  dplyr::filter(!is.na(Feature)) %>%
  dplyr::select(File, Feature, 
                # Day, 
                m.z, RT, Height) %>%
  dplyr::mutate(Source = ifelse(str_detect(File, "USE"), "Environmental", "Store-Bought")) %>%
  dplyr::rename(Values = Height)

hplc$technique <- "HPLC"
```

## STEP 4: Trace metal data

#### Data prepping
- Problem with icpms_round1 versus. icpms_round2: round1 has Li [No gas] that round2 does not.
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# Trace metal data
icpms_round1 <- readxl::read_excel("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/ICPMS_Trace metal/icpms_round1_rawdata_removal USE-01-rep1 and USE-03 (only2observations).xlsx") %>% 
  column_to_rownames(., var = "File")

icpms_round2_batch1 <- readxl::read_excel("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/ICPMS_Trace metal/icpms_round2_batch1_rawdata.xlsx") %>% 
  column_to_rownames(., var = "File")

icpms_round2_batch2 <- readxl::read_excel("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/ICPMS_Trace metal/icpms_round2_batch2_rawdata.xlsx") %>% 
  column_to_rownames(., var = "File")

# 1) Select [H2] instead of [H2 HMI]----------------------
library(stringr)

remove_H2_HMI_modus <- function(icpms_df) {
  strings <- colnames(icpms_df)
  pattern <- "^(.*\\[ )([A-Za-z0-9 ]+)( \\])$"
  metal_namedata <- data.frame(
    original = strings,
    prefix = str_extract(strings, "^.*\\["),
    content = str_match(strings, pattern)[,3],
    stringsAsFactors = FALSE
  )
  
  metal_namedata <- metal_namedata %>% dplyr::filter(!str_detect(content, pattern = "H2 HMI"))
  df <- icpms_df %>% dplyr::select(metal_namedata$original)
  return(df)
}

icpms_round1 <- remove_H2_HMI_modus(icpms_round1)
icpms_round2_batch1 <- remove_H2_HMI_modus(icpms_round2_batch1)
icpms_round2_batch2 <- remove_H2_HMI_modus(icpms_round2_batch2)

# 2) Between the ones that use [N2O] as analyzing modus -> figure out within the same metal element, which one has the highest "Recovery [%]" or if Recovery [%] is missing then lowest "U(k=2)" ---------------------------------------

recovery_uncertainty_round1 <- readxl::read_excel("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/ICPMS_Trace metal/Trace metal data_Recovery_Uncertainty_round1.xlsx") %>%
  filter(`Element+Modus` %in% colnames(icpms_round1))

recovery_uncertainty_round2_batch1 <- readxl::read_excel("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/ICPMS_Trace metal/Trace metal data_Recovery_Uncertainty_round2_batch1.xlsx") %>%
  filter(`Element+Modus` %in% colnames(icpms_round2_batch1))

recovery_uncertainty_round2_batch2 <- readxl::read_excel("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/ICPMS_Trace metal/Trace metal data_Recovery_Uncertainty_round2_batch2.xlsx") %>%
  filter(`Element+Modus` %in% colnames(icpms_round2_batch2))

get_final_metal_names <- function(recovery_uncertainty) {
  
  # 1) Extract the first number
  first_numbers <- sub(" .*", "", recovery_uncertainty$`Element+Modus`)
  
  # 2) Define a function to find duplicate indices
  duplicate_indices <- function(vector) {
    unique_elements <- unique(vector)
    result <- list()
    
    for (element in unique_elements) {
      indices <- which(vector == element)
      if (length(indices) > 1) {
        result[[element]] <- indices
      }
    }
    
    return(result)
  }
  
  # Apply duplicate_indices to first_numbers
  res <- duplicate_indices(first_numbers)
  
  # 3) For duplicates, pick either the row with the highest "Recovery [%]"
  #    or if that is empty, the row with the lowest "U(k=2)"
  recovery_uncertainty_checked <- c()
  for (ele in seq_along(res)) {
    temp <- recovery_uncertainty[res[[ele]], ]
    idx1 <- temp[which.max(temp$`Recovery [%]`), ]
    
    # if no maximum Recovery row is found, pick the minimum U(k=2) row instead
    if (identical(idx1, character(0))) {
      chosen <- temp[which.min(temp$`U(k=2)`), ]$`Element+Modus`
    } else {
      chosen <- temp[which.max(temp$`Recovery [%]`), ]$`Element+Modus`
    }
    recovery_uncertainty_checked <- c(recovery_uncertainty_checked, chosen)
  }
  
  # 4) Combine the unique rows (non-duplicates) plus the chosen duplicates
  final_metal_names <- c(
    recovery_uncertainty[-unlist(res, use.names = FALSE), ]$`Element+Modus`,
    recovery_uncertainty_checked
  )
  
  return(final_metal_names)
}

final_metal_names_round1 <- get_final_metal_names(recovery_uncertainty_round1)
final_metal_names_round2_batch1 <- get_final_metal_names(recovery_uncertainty_round2_batch1)
final_metal_names_round2_batch2 <- get_final_metal_names(recovery_uncertainty_round2_batch2)

# subset dataframe for further analyses
final_icpms_round1 <- icpms_round1 %>% 
  dplyr::select(all_of(final_metal_names_round1))

final_icpms_round2_batch1 <- icpms_round2_batch1 %>% 
  dplyr::select(all_of(final_metal_names_round2_batch1))

final_icpms_round2_batch2 <- icpms_round2_batch2 %>% 
  dplyr::select(all_of(final_metal_names_round2_batch2))

common_cols <- Reduce(intersect, list(names(final_icpms_round1), 
                                      names(final_icpms_round2_batch1), 
                                      names(final_icpms_round2_batch2)))
combined_icpms_round1_round2 <- rbind(final_icpms_round1[common_cols], 
                                      final_icpms_round2_batch1[common_cols],
                                      final_icpms_round2_batch2[common_cols])
View(combined_icpms_round1_round2)

# 3) Final cleaning of name of metal elements (OPTIONAL) ----------------------------------
# Function to process the strings
process_string <- function(name_vector) {
  # Remove everything after the square bracket
  cleaned_vector <- str_extract(name_vector, "^[^\\[]+")
  # Trim any trailing whitespace
  cleaned_vector <- str_trim(cleaned_vector)
  # Use a regular expression to remove the numerical parts
  # Explanation of Regular Expression:
  # 
  # \\d+ matches one or more digits.
  # *(-> *\\d+)? * optionally matches the pattern -> digits with any number of spaces around it.
  # gsub function is used to find and replace the matching parts in the string with an empty string ("")
  cleaned_vector <- gsub("\\d+ *(-> *\\d+)? *", "", cleaned_vector)
  return(cleaned_vector)
}

# Apply the function to each element of the vector
# colnames(combined_icpms_round1_round2) <- sapply(colnames(combined_icpms_round1_round2), process_string)

# 4) Discard all metal that may appear in seawater - Sodium (Na), Magnesium (Mg), Potassium (K), Calcium (Ca), Sr (optional) --------------------------
strings_to_remove <- c("Na", "Mg", "K", "Ca") #, "Sr")

cols_to_remove <- unique(unlist(sapply(strings_to_remove, function(x) grep(x, colnames(combined_icpms_round1_round2)))))
# remove the seawater metal from combined rounds ICP data
combined_icpms_round1_round2 <- combined_icpms_round1_round2[, -cols_to_remove]

icp <- combined_icpms_round1_round2 %>%
  rownames_to_column(., var = "File")  %>% 
  pivot_longer(., cols = 2:ncol(.), names_to = "Feature", values_to = "Values")

icp$technique <- "ICP"
icp <- icp %>%
  mutate(Source = ifelse(str_detect(File, "USE"), "Environmental", "Store-Bought")) %>%
  mutate(Values = as.numeric(Values))
```

## Combine Label Info with raw data
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
p <- "C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/Sample Labelling_all data_GC+HPLC+ICP_30Sept2025.xlsx"

sampinfo <- readxl::read_excel(path = p)

# joining data frame with sampinfo df
icp <- left_join(x = icp ,
                 y = sampinfo %>% dplyr::select(c(File, Plastic_type, Subcategory, Polymer)), 
                 by = 'File') # %>% dplyr::filter(!is.na(Similar_type))

gc <- left_join(x = gc ,
                y = sampinfo %>% dplyr::select(c(File, Plastic_type, Subcategory, Polymer)), 
                by = 'File') # %>% dplyr::filter(!is.na(Similar_type))

hplc <- left_join(x = hplc ,
                  y = sampinfo %>% dplyr::select(c(File, Plastic_type, Subcategory, Polymer)), 
                  by = 'File') # %>% dplyr::filter(!is.na(Similar_type))

## Aggregate Day 15 and 34 by mean
# new_hplc <- hplc %>%
#   group_by(Feature, Subcategory, Polymer, technique) %>%
#   summarise(Mean_Values = mean(Values, na.rm = TRUE), .groups = "drop")

gc_hplc <- bind_rows(gc, hplc)

gc_icp <- bind_rows(icp, 
                    gc %>% dplyr::select(-c("m.z", "RT"))) # %>% dplyr::filter(!is.na(Similar_type))

hplc_icp <- bind_rows(icp, 
                      hplc %>% dplyr::select(-c("m.z", "RT"))) # %>% dplyr::filter(!is.na(Similar_type))

gc_hplc_icp <- bind_rows(icp, 
                         gc_hplc %>% dplyr::select(-c("m.z", "RT"))) # %>% dplyr::filter(!is.na(Similar_type))

```


## Data cleaning
### Summary of data before feature reduction
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
df_in <- gc_hplc_icp
# 1. Process Data ONCE (Consolidated Pivot)
# We calculate the mean for duplicates and pivot to wide format immediately
df_wide <- df_in %>% 
  dplyr::select(File, Plastic_type, Source, Polymer, Feature, technique, Values) %>%
  dplyr::group_by(File, Source, Polymer, Plastic_type, technique, Feature) %>%
  dplyr::summarise(Values = mean(Values, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = Feature, values_from = Values) %>%
  tibble::column_to_rownames(var = "File") 

# 2. Separate Numeric Data from Metadata
# Identify metadata columns (non-numeric) to exclude them from calculations
df_numeric <- df_wide %>% dplyr::select(where(is.numeric))

# --- PART 1 & 2: Basic Statistics ---

cat("Number of chemical features: ", ncol(df_numeric), "\n")

# Note: Using df_numeric for dimensions to avoid diluting the % with metadata columns
missing_pct <- sum(is.na(df_numeric)) / prod(dim(df_numeric)) * 100
cat("Percentage of missing values (%):", round(missing_pct, 2), "\n")

# --- PART 3: Correlations ---

# Step 1: Calculate NA counts and Mean Response directly from the numeric matrix
stats_df <- tibble::tibble(
  Feature = colnames(df_numeric),
  num_missing_values = colSums(is.na(df_numeric)),
  mean_response = colMeans(df_numeric, na.rm = TRUE)
)

# Step 2: Check for Metadata Columns (The Fix)
# We only calculate m/z and RT correlations if the columns actually exist
if (all(c("m.z", "RT") %in% colnames(df_in))) {
  
  feature_meta <- df_in %>%
    dplyr::group_by(Feature) %>%
    dplyr::summarise(
      m_z = mean(m.z, na.rm = TRUE),
      retention_time = mean(RT, na.rm = TRUE),
      .groups = "drop"
    )
  
  # Join and calculate all correlations
  analysis_df <- stats_df %>% dplyr::left_join(feature_meta, by = "Feature")
  
  cat("Pearson (Mean Response vs Missing):", round(cor(analysis_df$mean_response, analysis_df$num_missing_values, use = "pairwise"), 2), "\n")
  cat("Pearson (m/z vs Missing):          ", round(cor(analysis_df$m_z, analysis_df$num_missing_values, use = "pairwise"), 2), "\n")
  cat("Pearson (RT vs Missing):           ", round(cor(analysis_df$retention_time, analysis_df$num_missing_values, use = "pairwise"), 2), "\n")
  
} else {
  
  # Fallback if m.z/RT are missing
  cor_response <- cor(stats_df$mean_response, stats_df$num_missing_values, use = "pairwise.complete.obs")
  
  cat("Pearson (Mean Response vs Missing):", round(cor_response, 2), "\n")
  message("\nNOTE: 'm.z' and 'RT' columns were not found in the input file. Those correlations were skipped.")
}
```

### Helper function for feature reduction (regular; modified 80% rule and in-house option)
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# ==============================================================================
# MAIN FUNCTION: Process, Impute Zeros, and Filter (Regular / Class / In-House)
# ==============================================================================
process_and_filter <- function(df_input, 
                               class_col = "Plastic_type", 
                               threshold = 0.8, 
                               filter_mode = "regular") { # "regular", "class", or "inhouse"
  
  # --- Step B: Pivot to Wide Format ---
  df_wide <- df_input %>% 
    dplyr::select(File, Source, Polymer, Feature, technique, Values, all_of(class_col)) %>%
    dplyr::group_by(File, Source, Polymer, technique, Feature, across(all_of(class_col))) %>%
    dplyr::summarise(Values = mean(Values, na.rm = TRUE), .groups = "drop") %>%
    tidyr::pivot_wider(names_from = Feature, values_from = Values) %>%
    tibble::column_to_rownames(var = "File") %>%
    dplyr::relocate(any_of(c(class_col, "technique", "Source", "Polymer")), .before = 1)
  
  # Metadata columns
  meta_cols <- c("Plastic_type", "technique", "Source", "Polymer")
  
  # --- Step D: Apply Feature Filtering Logic ---
  
  feature_data <- df_wide %>% dplyr::select(all_of(setdiff(names(df_wide), meta_cols)))
  kept_features <- c()
  
  if (filter_mode == "regular") {
    # Mode 1: Global 80% Rule
    cat("Applying 'Regular' filtering (Global threshold)...\n")
    presence_rate <- colMeans(!is.na(feature_data))
    kept_features <- names(presence_rate[presence_rate >= threshold])
    
  } else if (filter_mode == "class") {
    # Mode 2: Class-Based 80% Rule
    cat(paste0("Applying 'Class' filtering (Threshold met in at least one '", class_col, "')...\n"))
    
    class_split <- split(df_wide, df_wide[[class_col]])
    class_kept_list <- list()
    
    for (cls in names(class_split)) {
      cls_idx <- which(df_wide[[class_col]] == cls)
      cls_dat <- feature_data[cls_idx, ]
      cls_rate <- colMeans(!is.na(cls_dat))
      class_kept_list[[cls]] <- names(cls_rate[cls_rate >= threshold])
    }
    kept_features <- Reduce(union, class_kept_list)
    
  } else if (filter_mode == "inhouse") {
    # Mode 3: In-House Filtering
    cat("Applying 'In-House' filtering strategy...\n")
    
    # 1. Calculate Statistics per Feature
    classes <- unique(df_wide[[class_col]])
    n_feats <- ncol(feature_data)
    
    # Pre-allocate matrices for speed
    class_presence_mat <- matrix(0, nrow = length(classes), ncol = n_feats)
    class_counts_mat   <- matrix(0, nrow = length(classes), ncol = n_feats)
    
    for (i in seq_along(classes)) {
      cls <- classes[i]
      cls_rows <- which(df_wide[[class_col]] == cls)
      cls_dat  <- feature_data[cls_rows, , drop=FALSE]
      
      # A. % Present in this class (for Rule 1)
      class_presence_mat[i, ] <- colMeans(!is.na(cls_dat))
      
      # B. Exact Count of data points in this class (for Rule 4)
      class_counts_mat[i, ]   <- colSums(!is.na(cls_dat))
    }
    
    # Feature Stats
    global_presence      <- colMeans(!is.na(feature_data))
    max_class_presence   <- apply(class_presence_mat, 2, max)
    
    total_data_points    <- colSums(class_counts_mat)     # Sum of all counts
    n_classes_with_data  <- colSums(class_counts_mat > 0) # How many classes have at least 1 point
    
    # Key Metric for Rule 4: How many classes have >= 3 data points?
    n_classes_with_3plus <- colSums(class_counts_mat >= 3)
    
    # 2. Apply The 4 Rules Logic
    
    # Rule 1: Missingness (Keep if Global > 10% OR Max Class > 90%)
    keep_r1 <- (global_presence > 0.10) | (max_class_presence >= 0.90)
    
    # Rule 2: Singleton Sample (Keep if > 1 data point total)
    keep_r2 <- total_data_points > 1
    
    # Rule 3: Singleton Category (Keep if > 1 category... UNLESS diagnostic)
    is_diagnostic <- max_class_presence >= 0.90
    keep_r3 <- (n_classes_with_data > 1) | is_diagnostic
    
    # Rule 4: Sparse Distribution 
    # Logic: If present in >= 2 categories, we REQUIRE at least 2 categories to have >= 3 points.
    # If present in < 2 categories (i.e. 1), this rule is skipped (True).
    keep_r4 <- (n_classes_with_data < 2) | (n_classes_with_3plus >= 2)
    
    # FINAL DECISION: Feature must pass ALL checks
    final_mask <- keep_r1 & keep_r2 & keep_r3 & keep_r4
    kept_features <- colnames(feature_data)[final_mask]
    
  } else {
    stop("Invalid filter_mode. Choose 'regular', 'class', or 'inhouse'.")
  }
  
  # --- Step E: Build Final Dataframe ---
  final_cols <- unique(c(meta_cols, kept_features))
  df_filtered <- df_wide %>% dplyr::select(any_of(final_cols))
  
  return(list(
    original_data = df_wide,
    filtered_data = df_filtered,
    kept_features = kept_features
  ))
}
```

### Feature reduction 1: Regular 80% rule - Remove feature if there are more than 80% MVs

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# Replace 'gc_hplc_icp' with your actual dataframe name
results <- process_and_filter(hplc, 
                              class_col = "Plastic_type", 
                              threshold = 0.8, 
                              filter_mode = "regular")

# B. Extract Data for Calculation
df_final <- results$filtered_data
df_orig  <- results$original_data

# Define metadata columns to exclude from stats
meta_cols <- c("Plastic_type", "technique", "Source", "Polymer")

# Get list of feature names only
feats_final <- setdiff(names(df_final), meta_cols)
feats_orig  <- setdiff(names(df_orig), meta_cols)

# --- 1. Percentage of chemical features retained (%) ---
pct_features <- (length(feats_final) / length(feats_orig)) * 100
cat(paste0("1. Percentage of chemical features retained: ", round(pct_features, 2), "%\n"))

# --- 2. Percentage of missing values after feature reduction (%) ---
# We calculate this ONLY on the feature columns of the filtered dataset
if(length(results$kept_features) > 0) {
  vals_final <- df_final[, feats_final]
  pct_missing <- (sum(is.na(vals_final)) / prod(dim(vals_final))) * 100
  cat(paste0("2. Percentage of missing values after feature reduction: ", round(pct_missing, 2), "%\n"))
} else {
  vals_final <- df_final[, feats_final]
  cat(paste0("2. Percentage of missing values after feature reduction: ", 0, "%\n"))
}


# --- 3. Percentage of signal response after feature reduction (%) ---
# Sum of intensities in filtered data / Sum of intensities in original (zero-filled) data
signal_final <- sum(vals_final, na.rm = TRUE)
signal_orig  <- sum(df_orig[, feats_orig], na.rm = TRUE)
pct_signal   <- (signal_final / signal_orig) * 100

cat(paste0("3. Percentage of signal response after feature reduction: ", round(pct_signal, 2), "%\n"))
```

### Feature reduction 2: Modified 80% rule
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# Replace 'gc_hplc_icp' with your actual dataframe name
results <- process_and_filter(gc_hplc_icp, 
                              class_col = "Plastic_type", 
                              threshold = 0.8, 
                              filter_mode = "class")

# B. Extract Data for Calculation
df_final <- results$filtered_data
df_orig  <- results$original_data

# Define metadata columns to exclude from stats
meta_cols <- c("Plastic_type", "technique", "Source", "Polymer")

# Get list of feature names only
feats_final <- setdiff(names(df_final), meta_cols)
feats_orig  <- setdiff(names(df_orig), meta_cols)

# --- 1. Percentage of chemical features retained (%) ---
pct_features <- (length(feats_final) / length(feats_orig)) * 100
cat(paste0("1. Percentage of chemical features retained: ", round(pct_features, 2), "%\n"))

# --- 2. Percentage of missing values after feature reduction (%) ---
# We calculate this ONLY on the feature columns of the filtered dataset
if(length(results$kept_features) > 0) {
  vals_final <- df_final[, feats_final]
  pct_missing <- (sum(is.na(vals_final)) / prod(dim(vals_final))) * 100
  cat(paste0("2. Percentage of missing values after feature reduction: ", round(pct_missing, 2), "%\n"))
} else {
  vals_final <- df_final[, feats_final]
  cat(paste0("2. Percentage of missing values after feature reduction: ", 0, "%\n"))
}

# --- 3. Percentage of signal response after feature reduction (%) ---
# Sum of intensities in filtered data / Sum of intensities in original (zero-filled) data
signal_final <- sum(vals_final, na.rm = TRUE)
signal_orig  <- sum(df_orig[, feats_orig], na.rm = TRUE)
pct_signal   <- (signal_final / signal_orig) * 100

cat(paste0("3. Percentage of signal response after feature reduction: ", round(pct_signal, 2), "%\n"))
```


### Feature reduction 3: In-house Comprehensive feature filtering 
```{r , echo=FALSE, warning = FALSE, message=FALSE}
results <- process_and_filter(gc_hplc_icp, 
                              class_col = "Plastic_type", 
                              threshold = 0.8, 
                              filter_mode = "inhouse")

# B. Extract Data for Calculation
df_final <- results$filtered_data
df_orig  <- results$original_data

# Define metadata columns to exclude from stats
meta_cols <- c("Plastic_type", "technique", "Source", "Polymer")

# Get list of feature names only
feats_final <- setdiff(names(df_final), meta_cols)
feats_orig  <- setdiff(names(df_orig), meta_cols)

# --- 1. Percentage of chemical features retained (%) ---
pct_features <- (length(feats_final) / length(feats_orig)) * 100
cat(paste0("1. Percentage of chemical features retained: ", round(pct_features, 2), "%\n"))

# --- 2. Percentage of missing values after feature reduction (%) ---
# We calculate this ONLY on the feature columns of the filtered dataset
if(length(results$kept_features) > 0) {
  vals_final <- df_final[, feats_final]
  pct_missing <- (sum(is.na(vals_final)) / prod(dim(vals_final))) * 100
  cat(paste0("2. Percentage of missing values after feature reduction: ", round(pct_missing, 2), "%\n"))
} else {
  vals_final <- df_final[, feats_final]
  cat(paste0("2. Percentage of missing values after feature reduction: ", 0, "%\n"))
}

# --- 3. Percentage of signal response after feature reduction (%) ---
# Sum of intensities in filtered data / Sum of intensities in original (zero-filled) data
signal_final <- sum(vals_final, na.rm = TRUE)
signal_orig  <- sum(df_orig[, feats_orig], na.rm = TRUE)
pct_signal   <- (signal_final / signal_orig) * 100

cat(paste0("3. Percentage of signal response after feature reduction: ", round(pct_signal, 2), "%\n"))
```

### **Select shared features between store-bought and environmental**

**UPDATE 13 Dec 2025: I edited the feature reduction to work more correct than before and i do feature reduction and then finding shared features between SB and ENV. **

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Find shared features between SB and ENV
shared_feature_filtering <- function(dataset, feat_reduc_mode) {
  SB <- dataset %>% dplyr::filter(Source == "Store-Bought")
  ENV <- dataset %>% dplyr::filter(Source == "Environmental")
  
  shared_cols_SB_ENV <- Reduce(intersect, list(unique(SB$Feature), 
                                               unique(ENV$Feature)))
  
  combined_SB_ENV_shared_cols_ntr <- rbind(SB %>% 
                                             dplyr::filter(Feature %in% shared_cols_SB_ENV),
                                           ENV %>% 
                                             dplyr::filter(Feature %in% shared_cols_SB_ENV))
  output <- process_and_filter(combined_SB_ENV_shared_cols_ntr, 
                               class_col = "Plastic_type", 
                               threshold = 0.8, 
                               filter_mode = feat_reduc_mode)$filtered_data
  return(output)
}

# Then do data cleaning - change variable name accordingly to data combinations
gc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025 <- shared_feature_filtering(gc, 
                                                                               feat_reduc_mode = "inhouse")
hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025 <- shared_feature_filtering(hplc, 
                                                                                 feat_reduc_mode = "inhouse")
icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025 <- shared_feature_filtering(icp, 
                                                                                feat_reduc_mode = "inhouse")
```

**Update 13 Dec 2025: For data fusion, if i run it like this: gc_hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025 <- shared_feature_filtering(gc_hplc, feat_reduc_mode = "inhouse") => The amount of feature is too large (18759 features) which is not feasible at all for RFA feature selection approach or pretty much for any feature selection approach in RF classification later on.**

    **==> SOLUTION:  I create data fusion by combining the reduced feature , FOR EXAMPLE, to create gc_hplc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025 data, I first do Zero Replacement for all the features that only native to an analytical technique but not the other. Afterwards, I only select features from individual dataset that makeup the data fusion, for example to create gc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025; hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025 and icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025 to create this data fusion combinations.**
                                                                                    

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# --- Step A: Map Features to Techniques (Fast Lookup) ---
data_fusion <- function(df_input, 
                        source_feats,
                        class_col = "Plastic_type") {
  tech_map <- df_input %>% 
    dplyr::distinct(Feature, technique)
  
  all_techs <- unique(tech_map$technique)
  
  # --- Step B: Pivot to Wide Format ---
  df_wide <- df_input %>% 
    dplyr::select(File, Source, Polymer, Feature, technique, Values, all_of(class_col)) %>%
    dplyr::group_by(File, Source, Polymer, technique, Feature, across(all_of(class_col))) %>%
    dplyr::summarise(Values = mean(Values, na.rm = TRUE), .groups = "drop") %>%
    tidyr::pivot_wider(names_from = Feature, values_from = Values) %>%
    tibble::column_to_rownames(var = "File") %>%
    dplyr::relocate(any_of(c(class_col, "technique", "Source", "Polymer")), .before = 1)
  
  # Metadata columns
  meta_cols <- c("Plastic_type", "technique", "Source", "Polymer")
  
  # --- Step C: Fast Zero Replacement (Matrix Method) ---
  if (length(all_techs) > 1) {
    cat("Detected multiple techniques (", paste(all_techs, collapse=", "), "). Applying Fast Zero Replacement...\n")
    
    feat_cols <- setdiff(names(df_wide), meta_cols)
    mat <- as.matrix(df_wide[, feat_cols])
    row_techs <- df_wide$technique
    
    for (tech in all_techs) {
      tech_feats <- tech_map$Feature[tech_map$technique == tech]
      target_cols <- intersect(colnames(mat), tech_feats)
      
      if (length(target_cols) > 0) {
        target_rows <- which(row_techs != tech)
        if (length(target_rows) > 0) {
          sub_mat <- mat[target_rows, target_cols, drop = FALSE]
          sub_mat[is.na(sub_mat)] <- 0
          mat[target_rows, target_cols] <- sub_mat
        }
      }
    }
    df_wide[, feat_cols] <- mat
  } else {
    cat("Detected single technique. Skipping Zero Replacement.\n")
  }
  
  final_feat_cols <- c()
  for (source in source_feats) {
    f <- setdiff(names(source), meta_cols)
    final_feat_cols <- c(final_feat_cols, f)
  }
  output <- df_wide[, c(meta_cols, unique(final_feat_cols))]
  return(output)
}

gc_hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025 <- data_fusion(gc_hplc,
                                                                       source_feats = list(gc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
                                                                                           hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025))
gc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025 <- data_fusion(gc_icp,
                                                                      source_feats = list(gc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
                                                                                          icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025))
hplc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025 <- data_fusion(hplc_icp,
                                                                        source_feats = list(hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
                                                                                            icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025))
gc_hplc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025 <- data_fusion(gc_hplc_icp,
                                                                           source_feats = list(gc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
                                                                                               hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
                                                                                               icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025))

save(gc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025, 
     hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
     icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
     gc_hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
     gc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
     hplc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
     gc_hplc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025, file = "my_exported_data_13Dec2025_new 7 data combo for testing imp x norm.RData")
```

# Imputation and Normalization methods
```{r , echo=FALSE, warning = FALSE, message=FALSE}
impute_train_apply <- function(X_train, X_test, imp_name, k = 5){
  
  if(imp_name == "mean_imputation"){
    mu <- sapply(X_train, function(x) mean(x, na.rm = TRUE))
    fill <- function(X){
      X2 <- X
      for(nm in names(X2)){
        idx <- is.na(X2[[nm]])
        if(any(idx)) X2[[nm]][idx] <- mu[[nm]]
      }
      X2
    }
    return(list(train = fill(X_train), test = fill(X_test)))
  }
  
  if(imp_name == "median_imputation"){
    md <- sapply(X_train, function(x) median(x, na.rm = TRUE))
    fill <- function(X){
      X2 <- X
      for(nm in names(X2)){
        idx <- is.na(X2[[nm]])
        if(any(idx)) X2[[nm]][idx] <- md[[nm]]
      }
      X2
    }
    return(list(train = fill(X_train), test = fill(X_test)))
  }
  
  if(imp_name == "zero_imputation"){
    fill <- function(X){
      X2 <- X
      for(nm in names(X2)){
        idx <- is.na(X2[[nm]])
        if(any(idx)) X2[[nm]][idx] <- 0
      }
      X2
    }
    return(list(train = fill(X_train), test = fill(X_test)))
  }
  
  if(imp_name == "half_min_imputation"){
    halfmins <- sapply(X_train, function(x){
      nz <- x[!is.na(x) & x > 0]
      if(length(nz) > 0) min(nz) / 2 else 0
    })
    fill <- function(X){
      X2 <- X
      for(nm in names(X2)){
        idx <- is.na(X2[[nm]])
        if(any(idx)) X2[[nm]][idx] <- halfmins[[nm]]
      }
      X2
    }
    return(list(train = fill(X_train), test = fill(X_test)))
  }
  
  # Practical trainable replacements for leakage-hard cases
  # Update 8 Dec 2025 - from Discussion with Scott's:  avoid impute with multiple values because it will increase computational time and generate lots of uneven replacement values
  if(imp_name == "random_forest_imputation"){
    pp <- caret::preProcess(X_train, method = "bagImpute")
    return(list(train = predict(pp, X_train),
                test  = predict(pp, X_test)))
  }
  
  if(imp_name == "knn_imputation"){
    pp <- caret::preProcess(X_train, method = c("center","scale","knnImpute"), k = k)
    return(list(train = predict(pp, X_train),
                test  = predict(pp, X_test)))
  }
  
  stop("Unsupported imputer in train-apply mode: ", imp_name)
}

no_normalization <- function(x) x

normalize_train_apply <- function(X_train, X_test, norm_name){
  
  if(norm_name == "No_normalization"){
    return(list(train = X_train, test = X_test))
  }
  
  if(norm_name == "TSN_Normalization"){
    tsn <- function(X){
      rs <- rowSums(X, na.rm = TRUE)
      rs[rs == 0] <- 1
      as.data.frame(sweep(X, 1, rs, "/"))
    }
    return(list(train = tsn(X_train), test = tsn(X_test)))
  }
  
  if(norm_name == "Min_Max_Scaler"){
    mins <- apply(X_train, 2, min, na.rm = TRUE)
    maxs <- apply(X_train, 2, max, na.rm = TRUE)
    rng  <- maxs - mins
    rng[rng == 0] <- 1
    
    scale_mm <- function(X){
      Xmat <- as.matrix(X)
      Xmat <- sweep(Xmat, 2, mins, "-")
      Xmat <- sweep(Xmat, 2, rng, "/")
      as.data.frame(Xmat)
    }
    return(list(train = scale_mm(X_train), test = scale_mm(X_test)))
  }
  
  if(norm_name == "Z_Score_Normalization"){
    mu  <- colMeans(X_train, na.rm = TRUE)
    sdv <- apply(X_train, 2, sd, na.rm = TRUE)
    sdv[sdv == 0] <- 1
    
    zfun <- function(X){
      Xmat <- as.matrix(X)
      Xmat <- sweep(Xmat, 2, mu, "-")
      Xmat <- sweep(Xmat, 2, sdv, "/")
      as.data.frame(Xmat)
    }
    return(list(train = zfun(X_train), test = zfun(X_test)))
  }
  
  stop("Unsupported normalizer in train-apply mode: ", norm_name)
}

impute_norm_train_apply <- function(X_train, X_test, imp_name, norm_name){
  imp <- impute_train_apply(X_train, X_test, imp_name)
  normalize_train_apply(imp$train, imp$test, norm_name)
}

no_normalization <- function(x) x

###############################
##  4. Cluster resolution  ##
###############################

## Calculate cluster resolution using the first two PCA components
calculate_cluster_resolution <- function(data1, data2) {
  centroid1 <- colMeans(data1)
  centroid2 <- colMeans(data2)
  
  # Computes the Euclidean distance With Dim.1 as x-axis and Dim.2 as y-axis
  centroid_distance <-  sqrt((centroid1[1][[1]] - centroid2[1][[1]])^2 + (centroid1[2][[1]] - centroid2[2][[1]])^2)
  
  avg_distance1 <- mean(sqrt(rowSums((as.matrix(data1) - 
                       matrix(centroid1, nrow = nrow(data1), ncol = ncol(data1), byrow = TRUE))^2)))
  avg_distance2 <- mean(sqrt(rowSums((as.matrix(data2) - 
                       matrix(centroid2, nrow = nrow(data2), ncol = ncol(data2), byrow = TRUE))^2)))
  
  resolution <- centroid_distance / (avg_distance1 + avg_distance2)
  return(resolution)
}

## Compute average cluster resolution over all pairs of groups
calculate_average_cluster_resolution <- function(X, group_vector) {
  pca_model <- res.pca <- FactoMineR::PCA(
    X,
    scale.unit = FALSE,
    graph = FALSE)
  scores <- as.data.frame(pca_model$ind$coord[, 1:2])
  
  group <- as.factor(group_vector)
  group_levels <- levels(group)
  
  pair_resolutions <- c()
  
  for (i in 1:(length(group_levels) - 1)) {
    for (j in (i + 1):length(group_levels)) {
      data1 <- scores[group == group_levels[i], , drop = FALSE]
      data2 <- scores[group == group_levels[j], , drop = FALSE]
      
      if (nrow(data1) > 1 && nrow(data2) > 1) {
        res <- calculate_cluster_resolution(data1, data2)
        pair_resolutions <- c(pair_resolutions, res)
      }
    }
  }
  
  avg_resolution <- mean(pair_resolutions, na.rm = TRUE)
  return(avg_resolution)
}

mcc_multiclass <- function(true, pred) {
  true <- factor(true)
  pred <- factor(pred, levels = levels(true))
  cm <- table(true, pred)
  
  # Gorodkin multi-class MCC
  t_sum <- rowSums(cm)
  p_sum <- colSums(cm)
  n <- sum(cm)
  c <- sum(diag(cm))
  
  sum_pk_tk <- sum(p_sum * t_sum)
  sum_pk2   <- sum(p_sum^2)
  sum_tk2   <- sum(t_sum^2)
  
  numerator <- (c * n) - sum_pk_tk
  denom <- sqrt((n^2 - sum_pk2) * (n^2 - sum_tk2))
  
  if (denom == 0) return(NA_real_)
  numerator / denom
}

evaluate_combo_oob <- function(df_train, type_col, remove_cols,
                               imp_name, norm_name,
                               ntree = 500, seed = 123,
                               num_threads = 1) {
  
  X <- df_train %>% dplyr::select(-all_of(c(remove_cols, type_col))) %>% as.data.frame()
  y <- df_train[[type_col]]
  
  keep_cols <- names(X)[!sapply(X, function(v) all(is.na(v)))]
  X <- X[, keep_cols, drop = FALSE]
  
  if (ncol(X) < 1) return(list(OOB_MCC = NA_real_))
  if (length(unique(y)) < 2) return(list(OOB_MCC = NA_real_))
  
  pp <- tryCatch(
    impute_norm_train_apply(X, X, imp_name, norm_name),
    error = function(e) NULL
  )
  if (is.null(pp)) return(list(OOB_MCC = NA_real_))
  
  Xp <- pp$train
  if (any(is.na(Xp)) || any(!is.finite(as.matrix(Xp)))) {
    return(list(OOB_MCC = NA_real_))
  }
  
  y_fac <- factor(y)
  p     <- ncol(Xp)
  mtry  <- max(1, floor(sqrt(p)))
  
  train_df <- data.frame(.Class = y_fac, Xp, check.names = FALSE)
  
  set.seed(seed)
  rf <- ranger::ranger(
    dependent.variable.name = ".Class",
    data        = train_df,
    num.trees   = ntree,
    mtry        = mtry,
    splitrule   = "gini",
    importance  = "impurity",
    oob.error   = TRUE,
    probability = FALSE,
    num.threads = num_threads
  )
  
  oob_pred <- rf$predictions
  if (is.null(oob_pred)) {
    oob_pred <- predict(rf, data = train_df)$predictions
  }
  
  oob_pred <- factor(oob_pred, levels = levels(y_fac))
  mcc <- mcc_multiclass(y_fac, oob_pred)
  
  list(OOB_MCC = mcc)
}

##################################################################################
##   6. Final Workflow: Iterate Over Datasets, Imputation, and Normalization    ##
##################################################################################
imputation_methods <- c(
  "mean_imputation",
  "median_imputation",
  "zero_imputation",
  "half_min_imputation"
)

normalization_methods <- c(
  "No_normalization",
  "TSN_Normalization",
  "Min_Max_Scaler",
  "Z_Score_Normalization"
)

find_best_impute_normalize <- function(df_train,
                                       type_col,
                                       remove_cols,
                                       ntree_screen = 500) {
  
  X <- df_train %>% dplyr::select(-all_of(c(remove_cols, type_col))) %>% as.data.frame()
  y <- df_train[[type_col]]
  
  results_list <- list()
  
  for(imp_name in imputation_methods){
    for(norm_name in normalization_methods){
      
      oob_res <- evaluate_combo_oob(
        df_train, type_col, remove_cols,
        imp_name, norm_name,
        ntree = ntree_screen
      )
      
      # Unsupervised metrics on train only
      pp_self <- tryCatch(
        impute_norm_train_apply(X, X, imp_name, norm_name),
        error = function(e) NULL
      )
      if(is.null(pp_self)) next
      
      X_norm <- pp_self$train
      if(any(is.na(X_norm)) || any(!is.finite(as.matrix(X_norm)))) next
      
      cluster_res <- calculate_average_cluster_resolution(X_norm, y)
      
      orig_corr <- tryCatch(cor(X, use = "pairwise.complete.obs"), error = function(e) NULL)
      imp_corr  <- tryCatch(cor(X_norm, use = "pairwise.complete.obs"), error = function(e) NULL)
      
      corr_score <- NA_real_
      if (!is.null(orig_corr) && !is.null(imp_corr) &&
          all(dim(orig_corr) == dim(imp_corr)) && ncol(orig_corr) > 1) {
        corr_score <- suppressWarnings(
          cor(orig_corr[upper.tri(orig_corr)],
              imp_corr [upper.tri(imp_corr)],
              use = "complete.obs")
        )
      }
      
      # KS on train only
      observed_all <- c(); imputed_all <- c()
      for(col in names(X)) {
        if(any(is.na(X[[col]])) && any(!is.na(X[[col]]))){
          observed_all <- c(observed_all, X[[col]][!is.na(X[[col]])])
          imputed_all  <- c(imputed_all, X_norm[[col]])
        }
      }
      ks_p <- if(length(observed_all) < 2 || length(imputed_all) < 2) NA else
        tryCatch(stats::ks.test(observed_all, imputed_all)$p.value, error = function(e) NA)
      
      results_list[[length(results_list)+1]] <- data.frame(
        Imputation     = imp_name,
        Normalization  = norm_name,
        OOB_MCC        = oob_res$OOB_MCC,
        ClusterRes     = cluster_res,
        CorrScore      = corr_score,
        KS_p           = ks_p,
        stringsAsFactors = FALSE
      )
    }
  }
  
  if (length(results_list) == 0) {
    return(list(results_table = data.frame(), best_combo = data.frame()))
  }
  
  df_results <- do.call(rbind, results_list)
  
  ord <- order(
    -df_results$OOB_MCC,
    -df_results$ClusterRes,
    -df_results$CorrScore,
    -df_results$KS_p
  )
  df_results$combined_rank <- NA_integer_
  df_results$combined_rank[ord] <- seq_along(ord)
  
  best_row <- df_results[which.min(df_results$combined_rank), ]
  list(results_table = df_results, best_combo = best_row)
}
```


## Figure 2: PCA
```{r , echo=FALSE, warning = FALSE, message=FALSE}
# select_feats <- c(run_rf_analysis_list_SB_and_ENV_train[[1]]$top_importance_results$top_100$selected_features, 
#                   colnames(run_rf_analysis_list_SB_and_ENV_train[[1]]$final_imp_norm_dat)[1:5])

# df <- run_rf_analysis_list_SB_and_ENV_train[[1]]$final_imp_norm_dat %>% 
#        dplyr::select(all_of(select_feats)) %>%
#        dplyr::relocate(Subcategory, Source, Polymer, technique, Plastic_type, .before = 1)

df <- df_0.001_TSN %>% # run_rf_analysis_list_SB_and_ENV_train_02Aug2025$hplc_SB_and_ENV_train$final_imp_norm_dat %>%
  dplyr::mutate(Plastic_type = gsub("..", " ", Plastic_type, fixed = TRUE)) %>%
  dplyr::mutate(Plastic_type = gsub(".", " ", Plastic_type, fixed = TRUE)) %>%
  dplyr::mutate(Plastic_type_Source = paste0(Plastic_type, " ", Source))

# Must have more than 3 samples per plastic type
# df  <- df [!(df $Plastic_type %in% names(table(df $Plastic_type)[table(df $Plastic_type) < 3])),]

res.pca <- FactoMineR::PCA(ncp =10,
                           df %>% 
                             dplyr::select(-c("Subcategory", "Source", "Plastic_type", "Polymer", "technique",  "Plastic_type_Source")), # %>% # ,  "Plastic_type_Source")),
                             # dplyr::select(run_rf_analysis_list_SB_and_ENV_train_02Aug2025$gc_SB_and_ENV_train$rfa_selected_feats),
                           scale.unit = FALSE,
                           graph = FALSE)

# Scree plot
# fviz_screeplot(res.pca)

# 1) Define a bold, color-blind–safe 9-color palette (Okabe–Ito + gray)
# Biplot
pca_plot <- factoextra::fviz_pca_biplot(res.pca,
                                        axes = c(1,2), # PC1 and PC2
                                        geom = c("point", "text"),
                                        label = "none",
                                        invisible = "var",
                                        repel = TRUE,
                                        pointsize = 5,
                                        habillage = factor(df$Plastic_type),
                                        # addEllipses = TRUE,
                                        # ellipse.level=0.95,
                                        ggtheme = ggplot2::theme_minimal(base_size = 25),
                                        # title = "gc_SB_and_ENV_train - RFA features",
                                        # xlab = "PC 1 (52.6%)",
                                        # ylab = "PC 2 (36.2%)",
                                        # xlim = c(-0.5, 0.8),
                                        # ylim = c(-0.6, 0.45)
                                        ) +
  # if has error "Too few points to calculate an ellipse"
  # ggforce::geom_mark_ellipse(aes(fill = Groups,
                                 # color = Groups),
                             # label.buffer = unit(40, 'mm')) +
  theme(legend.position = 'bottom',
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank()
        )
pca_plot

ggsave(filename = paste0("Figure SI_GC_SB and ENV_0.0001-TSN_Plastic_type_", format(Sys.time(), "%d-%m-%Y"), ".tiff"), 
       plot = pca_plot, 
       device = "tiff",
       compression = "lzw",
       dpi = 600,        # Set DPI to 600
       width = 28,       # Width in inches (adjust as needed)
       height = 12,       # Height in inches (adjust as needed)
       units = "in")
```

## Figure 2: HCA

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Data prep for HCA ---------------------
df <- run_rf_analysis_list_SB_and_ENV_train_02Aug2025$hplc_SB_and_ENV_train$final_imp_norm_dat %>%
  dplyr::mutate(Plastic_type = gsub("..", " ", Plastic_type, fixed = TRUE)) %>%
  dplyr::mutate(Plastic_type = gsub(".", " ", Plastic_type, fixed = TRUE)) %>%
  dplyr::mutate(Plastic_type_Source = paste0(Plastic_type, "-", Source))

hc_df <- df %>% 
  dplyr::select(-c("Subcategory", "Source", "Plastic_type", "Polymer", "technique")) %>% #, "Plastic_type_Source")) %>%
  dplyr::select(where(is.numeric)) %>%
  as.data.frame()

# --- Clean: remove bad rows & handle non-finite values ---
# 1) Replace non-finite with 0 (or choose a more domain-specific imputation if appropriate)
hc_df[!is.finite(as.matrix(hc_df))] <- 0

# 2) Drop rows that are all zero (Bray–Curtis can yield NaN for empty rows)
keep <- rowSums(hc_df, na.rm = TRUE) > 0
hc_df_clean <- hc_df[keep, , drop = FALSE] #%>% 
  # Select the features from best performance feature selection method
  # dplyr::select(run_rf_analysis_list_SB_and_ENV_train_02Aug2025$icp_SB_and_ENV_train$rfa_selected_feats)
labels_clean <- df$Plastic_type_Source[keep] # df$Plastic_type_Source[keep] 

# 3) If any negatives remain, Bray–Curtis is invalid -> fall back to a metric that permits negatives
use_bray <- !any(hc_df_clean < 0, na.rm = TRUE)

if (use_bray) {
  d <- vegan::vegdist(hc_df_clean, method = "manhattan")
  link <- "average"
} else {
  # If your features were centered/scaled, Euclidean or Manhattan are appropriate.
  # (Alternatively, if these are compositions, consider Aitchison with a small pseudocount.)
  d <- dist(hc_df_clean, method = "robust.aitchison")
  link <- "average"
  message("Detected negative values; using Euclidean instead of Bray–Curtis.")
}

# --- Guard: ensure distance has no NA/NaN/Inf ---
if (any(!is.finite(d))) {
  stop("Distance matrix still has non-finite entries after cleaning. Re-check inputs.")
}

# --- Cluster & plot ---
hca_samp <- hclust(d, method = link)

cat("cophenetic correlation coefficient (CCC) from a hierarchical clustering (HCA) is:", cor(d, stats::cophenetic(hca_samp)), "\n")

# tiff(filename = paste0("Figure SI_HCA_HPLC_SB_and_ENV_train_", format(Sys.time(), "%d-%m-%Y"),".tif"), width = 42*600, height = 15*600, res = 600, compression = "lzw")

par(cex = 1)
plot(hca_samp,
     labels = labels_clean,
     hang   = -1,
     main   = "")
# dev.off()
```


A smaller Aitchison distance between two samples would indicate that their compositional structures are similar in terms of the ratios between components.

For selecting Dissimilarity Indices Calculated by vegan::vegdist(), since our data is continous, not normally distributed, and the data is sparse (meaning our data has lots of NA or NULL before data imputation, This basically because there are compounds that occur in some samples and not in the others). Manhattan dissimilarity index separating ATDGCMS and HPLCTOFMS into 2 big clusters, while Canberra//Aitchison//Robust.aitchison grouped ATDGCMS as a subgroup of a part of HPLCTOFMS data.

Since for each observations of each sample, the sum of values of collapsed compounds is sum up to 100%, Aitchison or Robust.Aitchison are currently the best option for dissimilarity index.

***Why Aitchison or Robust.Aitchison for OUR DATA, AKA. Compositional Data?***

Compositional data have particular characteristics and constraints that standard distance metrics like Euclidean don't handle well. Here are some reasons why Aitchison distance is more suitable:

**Closure Constraint:** Because the components in compositional data sum to a constant value (like 1 or 100%), the data is subject to a closure constraint. This constraint violates the assumption of independence between variables that many standard statistical techniques rely on. Aitchison distance accounts for this constraint.

*Implications of Closure Constraint* *Loss of Degrees of Freedom:* Because of the closure constraint, one variable is effectively 'dependent' on the others. For instance, in our example, knowing the proportions of Additive A and Additive B is enough to determine the proportion of Additive C.

*Spurious Correlations:* The closure constraint can lead to spurious or misleading correlations between variables. For example, an increase in the proportion of Additive A will necessarily result in a decrease in the sum of the proportions of Additives B and C, even if there's no biological or chemical reason for them to be inversely related.

*Non-Independence of Components:* In compositional data, the components are not independent of each other, violating the assumptions of independence that many statistical tests and measures (like the Euclidean distance) rely on.

*Comparisons are Relative:* In compositional data, what matters is the relative sizes of the components, not their absolute sizes. Any analysis should, therefore, focus on these relative proportions rather than absolute values.

For these reasons, specialized statistical methods like Aitchison distance are often used for compositional data. These methods take the closure constraint into account and focus on the relative proportions of the different components, providing a more accurate and interpretable analysis.

In this workflow on tracking microplastic additives in aquatic environments, understanding the closure constraint can be pivotal. If you're looking at the relative abundances of different types of plastic additives in water samples, those abundances are compositional data subject to the closure constraint. Using standard statistical techniques without accounting for this constraint could lead to misleading conclusions.

**Relative Importance:** In compositional data, what's generally important is the relative proportion between components, rather than their absolute values. The Aitchison distance is designed to capture the relationships between these relative values effectively.

**Log-Ratio Transformation:** The Aitchison distance is based on log-ratio transformations, which have been proven to be an effective way to analyze compositional data. The log-ratio removes the closure constraint, making the data easier to analyze statistically.

**Handling Zeros:** The robust variant of Aitchison distance often includes techniques for handling zeros in compositional data, which are common and challenging to deal with.

**Scale-Invariance:** Aitchison distance is scale-invariant, meaning that multiplying a component by a constant will not change the distance. This is crucial for compositional data, where only the relative relationships between components matter.

**Interpretable:** The distance measure reflects compositional difference in a way that's meaningful within the context of the data, making it easier to understand and interpret the results.



So with the dimensional reduction of compounds from n = 13046 to 33, we still achieve a somewhat good clustering result in which sample of the same category tends to cluster together.

## Table S6: Correlation and covariance magnitude of features across 7 data combinations
```{r , echo=FALSE, warning = FALSE, message=FALSE}
## Before imp_norm   ===============================================
numeric_df <- gc_hplc_icp_combined_SB_ENV_shared_cols_ntr_clean %>% 
  dplyr::select(where(is.numeric))
cor_mat <- cor(numeric_df,
               use    = "pairwise.complete.obs",
               method = "pearson")
cov_mat <- cov(numeric_df,
               use    = "pairwise.complete.obs")
# extract off‐diagonals
cor_off <- cor_mat[upper.tri(cor_mat)]
cov_off <- cov_mat[upper.tri(cov_mat)]

# drop NAs before computing
cor_off <- cor_off[!is.na(cor_off)]
cov_off <- cov_off[!is.na(cov_off)]

# now summary statistics
avg_abs_cor  <- mean(abs(cor_off))
fro_cov_off  <- sqrt( sum(cov_off^2) )

round(avg_abs_cor,3)
round(fro_cov_off,3)

## After imp_norm  ===============================================
numeric_df <- run_rf_analysis_list_SB_and_ENV_train$gc_hplc_icp_SB_and_ENV_train$final_imp_norm_dat %>% 
  dplyr::select(where(is.numeric))

# Average absolute pairwise correlation
# 0 means no linear association anywhere.
# 1 means every pair is perfectly (positively or negatively) correlated.
# Pearson correlation matrix (pairwise complete observations)
cor_mat <- cor(numeric_df,
               use    = "pairwise.complete.obs",
               method = "pearson")
off_diag <- cor_mat[upper.tri(cor_mat)]
avg_abs_cor <- mean(abs(off_diag))

# Average absolute covariance
# 0 means every pair is (on average) uncorrelated—though note you could have positive and negative covariances that cancel each other out if you didn’t take the absolute value.
# Larger values mean more joint variability.
# Covariance matrix (pairwise complete observations)
cov_mat <- cov(numeric_df,
               use    = "pairwise.complete.obs")
# mean_abs_cor <- mean(abs(cor_mat[upper.tri(cor_mat)]))
# mean_abs_cor

cov_off <- cov_mat[upper.tri(cov_mat)]
fro_cov_off <- sqrt( sum(cov_off^2) )

round(avg_abs_cor,3)
round(fro_cov_off,3)
```

## Table S8. Significant compounds identified via pair-wise tests between plastic categories across seven data combinations
```{r echo=FALSE, warning=FALSE, message=FALSE}
list_pairwise_df <- list()
for (i in 1:length(run_rf_analysis_list_SB_and_ENV_train_02Aug2025)) {
  pw <- pairwise_significance_tests(run_rf_analysis_list_SB_and_ENV_train_02Aug2025[[i]]$final_imp_norm_dat, 
                                    group_col = "Plastic_type", start_col_index = 6)
  list_pairwise_df[[i]] <- pw$uncorrected %>%
    mutate(adjusted_p = p.adjust(pval, method = "bonferroni")) %>%
    filter(adjusted_p < 0.05) %>%
    arrange(adjusted_p) %>%
    dplyr::mutate(Datasets = names(run_rf_analysis_list_SB_and_ENV_train_02Aug2025)[i])
}

pw_df <- bind_rows(list_pairwise_df)
write_xlsx(pw_df, path = "Table S8_Significant compounds identified via pair-wise tests between plastic categories.xlsx")
```

## Figure 3: Bar plot Comparing Feature selection - MCC score - 7 data combinations
```{r echo=FALSE, warning=FALSE, message=FALSE}
df_list <- list()
dataset_names <- c("GC", "HPLC", "ICP", "GC-HPLC", "GC-ICP", "HPLC-ICP", "GC-HPLC-ICP")
for (i in 1:length(names(run_rf_analysis_list_SB_and_ENV_train_02Aug2025))) {
  names <- names(run_rf_analysis_list_SB_and_ENV_train_02Aug2025)[i]
  df_list[[i]] <- data.frame(time = round(c(dim(run_rf_analysis_list_SB_and_ENV_train_02Aug2025[[names]]$final_imp_norm_dat)[2],
                                          length(run_rf_analysis_list_SB_and_ENV_train_02Aug2025[[names]]$sig_selected_feats),
                                          length(run_rf_analysis_list_SB_and_ENV_train_02Aug2025[[names]]$rfa_selected_feats),
                                          length(run_rf_analysis_list_SB_and_ENV_train_02Aug2025[[names]]$top_importance_results$top_100$selected_features),
                                          length(run_rf_analysis_list_SB_and_ENV_train_02Aug2025[[names]]$top_importance_results$top_50$selected_features),
                                          length(run_rf_analysis_list_SB_and_ENV_train_02Aug2025[[names]]$top_importance_results$top_25$selected_features),
                                          length(run_rf_analysis_list_SB_and_ENV_train_02Aug2025[[names]]$top_importance_results$top_10$selected_features)), 3)) %>% dplyr::mutate(Dataset = dataset_names[i]) %>% dplyr::mutate(Selection = c("All features", "Pairwise", "RFA", "Top 100", "Top 50", "Top 25", "Top 10"))
}
View(bind_rows(df_list))
write_xlsx(bind_rows(df_list), path = paste0("Computational fingerprinting paper_feature selection comparison_train SB+ENV_04Aug2025_number selected features_", format(Sys.time(), "%d-%m-%Y"), ".xlsx"))

# PLot =============================
feature_selection <- read_xlsx(path = "C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/Computational fingerprinting paper_feature selection comparison_train SB+ENV_MCC score_time_feature number_04Aug2025.xlsx") %>%  
  dplyr::mutate(Dataset = factor(Dataset, levels = c("GC", "HPLC", "ICP", "GC-HPLC", "GC-ICP", "HPLC-ICP", "GC-HPLC-ICP")))

feature_selection_plot <- ggplot(data = feature_selection, aes(x = Dataset, y = MCC, fill = `Feature selection`)) + 
  geom_bar(position = position_dodge(width = 0.9), stat = "identity") +
  scale_fill_manual(values =  c('#8dd3c7','#ffffb3','#bebada','#fb8072','#80b1d3','#fdb462','#b3de69','#fccde5')) +  
  scale_y_continuous(breaks = seq(from = 0, to = 1, by =.1), expand = c(0,0), limits = c(0, 1)) +
  theme_bw(base_size = 20) +
  theme(axis.text.x = element_text(size =20), axis.text.y = element_text(size =20))
feature_selection_plot

ggsave(filename = paste0("Figure 3_Feature selection - MCC score - 7 data combinations_", format(Sys.time(), "%d-%m-%Y"), ".png"), 
       plot = feature_selection_plot, 
       device = "png",
       dpi = 300,        # Set DPI to 600
       width = 18,       # Width in inches (adjust as needed)
       height = 8,       # Height in inches (adjust as needed)
       units = "in")
```

## Figure 4: Confusion matrices of training and testing on environmental plastics versus. training on store-bought plastics and testing on environmental plastics - 7 data combinations
```{r echo=FALSE, warning=FALSE, message=FALSE}
sb_train_only <- run_rf_analysis_list_SB_train$gc_USSB_train$conf_mat_sig
sb_and_env_train <- run_rf_analysis_list_SB_and_ENV_train_02Aug2025$gc_SB_and_ENV_train$conf_mat_rfa

ggsave(filename = paste0("Figure 4_confusion matrix_SB train only_", "GC", "_", format(Sys.time(), "%d-%m-%Y"), ".png"), 
       plot = sb_train_only, 
       device = "png",
       dpi = 300,        # Set DPI to 600
       width = 15,       # Width in inches (adjust as needed)
       height = 8,       # Height in inches (adjust as needed)
       units = "in")

ggsave(filename = paste0("Figure 4_confusion matrix_SB and ENV train", "GC", "_", format(Sys.time(), "%d-%m-%Y"), ".png"), 
       plot = sb_and_env_train, 
       device = "png",
       dpi = 300,        # Set DPI to 600
       width = 15,       # Width in inches (adjust as needed)
       height = 8,       # Height in inches (adjust as needed)
       units = "in")
```

## Figure SI (For Roxana's paper): Chord diagram of classification - individual & combined data combinations - platic categories with correct prediction probablity > 0.5

(UPDATE 24 July 2025:) Change color to 7 color corresponding to 7 data combinations, the plastic categories will "flow" with 7 colors into 7 data combinations

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(circlize)

# Data frame , 1column is the Analytical dataset, 1 column is the Plastic type, 1 column is the number of correctly classified plastic product 
prob_mat_list <- list(run_rf_analysis_list_SB_and_ENV_train$gc_SB_and_ENV_train$prob_matrix_sig,
                      run_rf_analysis_list_SB_and_ENV_train$hplc_SB_and_ENV_train$prob_matrix_rfa,
                      run_rf_analysis_list_SB_and_ENV_train$icp_SB_and_ENV_train$prob_matrix_rfa,
                      run_rf_analysis_list_SB_and_ENV_train$gc_icp_SB_and_ENV_train$top_importance_results$top_10$prob_matrix_top,
                      run_rf_analysis_list_SB_and_ENV_train$gc_hplc_SB_and_ENV_train$prob_matrix_rfa,
                      run_rf_analysis_list_SB_and_ENV_train$hplc_icp_SB_and_ENV_train$prob_matrix_sig,
                      run_rf_analysis_list_SB_and_ENV_train$gc_hplc_icp_SB_and_ENV_train$prob_matrix,
                      run_rf_analysis_list_SB_only$hplc_SB$prob_matrix_rfa,
                      run_rf_analysis_list_SB_only$icp_SB$prob_matrix_rfa,
                      run_rf_analysis_list_SB_only$hplc_icp_SB$top_importance_results$top_50$prob_matrix_top)

name <- c("GC",
          "HPLC",
          "ICP",
          "GC-HPLC",
          "GC-ICP",
          "HPLC-ICP",
          "GC-HPLC-ICP", 
          "HPLC",
          "ICP",
          "HPLC-ICP")

for (i in 1:length(prob_mat_list)) {
  # 2. Compute, for each row, which column has the max value
  #    `max.col()` is a fast way to get the index of the max in each row
  idx_max <- max.col(prob_mat_list[[i]], ties.method = "first")
  
  # 3. Map those indices back to class names
  predicted_class <- colnames(prob_mat_list[[i]])[idx_max]
  
  # 4. Extract the max probability values
  #    Method A: via matrix indexing
  row_indices     <- seq_len(nrow(prob_mat_list[[i]]))
  predicted_prob  <- prob_mat_list[[i]][cbind(row_indices, idx_max)]
  
  # 4. If you’d like it all in a data.frame:
  df <- as.data.frame(prob_mat_list[[i]])
  df <- df %>% 
    dplyr::mutate(Actual_Class = sub("_.*", "", rownames(prob_mat_list[[i]]))) %>% # bring actual labels in as a column
    dplyr::mutate(Predicted_Class = predicted_class) %>%
    dplyr::mutate(Predicted_Prob = predicted_prob) %>%
    dplyr::mutate(Dataset = name[i]) %>%
    dplyr::relocate(Actual_Class, Predicted_Class, Dataset, Predicted_Prob, .before = 1) %>%
    dplyr::filter(Actual_Class == Predicted_Class)
  
  prob_mat_list[[i]] <- df
}

chorddiagram_df <- bind_rows(prob_mat_list) %>% 
  dplyr::filter(Predicted_Prob > 0.5) %>%
  dplyr::select(c(Dataset, Predicted_Class, Predicted_Prob)) %>%
  dplyr::mutate(Dataset = gsub(pattern = "-", replacement = "_", Dataset))

png(
  filename = paste0("chord_diagram_", format(Sys.time(), "%d-%m-%Y"), ".png"),
  width    = 18,      # in inches
  height   = 18,      # in inches
  units    = "in",
  res      = 600     # 300 dpi for publication‑quality
)

# 1) Label mapping for bottom dataset sectors
label_map <- c(
  GC            = "GC",
  HPLC          = "HPLC",
  ICP           = "ICP",
  GC_HPLC       = "GC-HPLC",
  GC_ICP        = "GC-ICP",
  HPLC_ICP      = "HPLC-ICP",
  GC_HPLC_ICP   = "GC-HPLC-ICP"
)
# Have gaps between sectors
circos.par(gap.after = c(rep(2, length(unique(chorddiagram_df[[1]]))-1), 15, 
                         rep(2, length(unique(chorddiagram_df[[2]]))-1), 15))

# Manually assign color to bottom sectors (Datasets)
grid.col <- c(GC='#8dd3c7',HPLC='#ffffb3',ICP='#bebada',GC_ICP = '#fb8072',HPLC_ICP = '#80b1d3', GC_HPLC = '#fdb462',GC_HPLC_ICP = '#b3de69',
              Balloons="grey", Plastic.Bottles="grey", Plastic.Drinking.Straws="grey", Bottle.Caps="grey", Construction.Material="grey", Mixed.Microplastics="grey", Plastic.Cups="grey", Shotgun.shells="grey", Fishing.Bait.Trays="grey", Food.Containers="grey", Food.Packaging="grey", Food.Wrappers="grey", Textile.Fabric.Face.Masks="grey", Toy.Balls="grey", Cigarillo.Tips="grey", Various.Plastic.Waste="grey", Plastic.Cutlery="grey", Cigarettes="grey", Cigarette.Lighters="grey", Plastic.Cord="grey", Plastic.Cable.Ties="grey", Plastic.Plates="grey")

circlize::chordDiagram(chorddiagram_df,
                       grid.col = grid.col,
                       small.gap = 2, 
                       annotationTrack   = c("grid"),
                       preAllocateTracks = 1)

circos.trackPlotRegion(
  track.index = 1, 
  bg.border = NA,
  panel.fun = function(x, y) {
    sector <- get.cell.meta.data("sector.index")
    xlim   <- get.cell.meta.data("xlim")
    
    # Use mapped label if available, otherwise keep original
    label <- ifelse(sector %in% names(label_map), label_map[[sector]], sector)
    
    circos.text(
      x       = mean(xlim), y = 0,
      labels  = label,
      facing  = "clockwise",     # or "reverse.clockwise"
      niceFacing = TRUE,        # turns off auto-flip
      adj     = c(0, 0.5),
      cex     = 1.5
    )
  }
)
grid.text(
  "Individual and combined datasets",
  x    = unit(0.5, "npc"),        # centered horizontally
  y    = unit(0.075, "npc"),           # 2 mm up from bottom
  just = "center",
  gp   = gpar(fontsize = 20, fontface = "bold")
)

# 2) And the top label for "Dataset"
grid.text(
  "Plastic categories",
  x    = unit(0.5, "npc"),        
  y    = unit(0.98, "npc"),  # 5 mm down from top
  just = "center",
  gp   = gpar(fontsize = 20, fontface = "bold")
)
circos.clear()
dev.off()
```

## Post-hoc Pearson GC-HPLC-ICP data
```{r class.source = 'fold-hide', echo=FALSE, warning = FALSE, message=FALSE}
X_RFA <-  run_rf_analysis_list_SB_and_ENV_train_02Aug2025$gc_hplc_SB_and_ENV_train$final_imp_norm_dat %>% dplyr::select(run_rf_analysis_list_SB_and_ENV_train_02Aug2025$gc_hplc_SB_and_ENV_train$rfa_selected_feats)

X_pairwise <-  run_rf_analysis_list_SB_and_ENV_train_02Aug2025$gc_hplc_SB_and_ENV_train$final_imp_norm_dat %>% dplyr::select(run_rf_analysis_list_SB_and_ENV_train_02Aug2025$gc_hplc_SB_and_ENV_train$sig_selected_feats)

# y <- run_rf_analysis_list_SB_and_ENV_train_02Aug2025$gc_hplc_SB_and_ENV_train$final_imp_norm_dat$Plastic_type
# 
# library(FSelectorRcpp)
# df1 <- data.frame(y = y, X_pairwise)
# df2 <- data.frame(y = y, X_rfa)
# 
# ig1 <- information_gain(y ~ ., df1); ig2 <- information_gain(y ~ ., df2)
# 
# S1 <- cor(df1[-1], use="pairwise.complete.obs"); S2 <- cor(df2[-1], use="pairwise.complete.obs")
# red1 <- rowMeans(abs(S1), na.rm=TRUE); red2 <- rowMeans(abs(S2), na.rm=TRUE)
# 
# mrmr1 <- ig1$importance - red1[ig1$attributes]
# mrmr2 <- ig2$importance - red2[ig2$attributes]
# summary(mrmr1); summary(mrmr2)
# 
# 
# one_vs_rest_mi <- function(y, X){
#   levs <- levels(y)
#   res <- sapply(levs, function(cls){
#     ybin <- factor(ifelse(y==cls, cls, "other"))
#     information_gain(ybin ~ ., data.frame(ybin, X))$importance
#   })
#   rowMeans(res, na.rm=TRUE)           # avg MI across classes per feature
# }
# pc_pairwise <- one_vs_rest_mi(y, X_pairwise)
# pc_rfa      <- one_vs_rest_mi(y, X_rfa)
# summary(pc_pairwise); summary(pc_rfa)


# ============================
# Redundancy Diagnostics (R)
# ============================

# Packages
suppressPackageStartupMessages({
  require(igraph)   # graph metrics
  require(Matrix)   # nearPD (for stabilizing correlation matrices)
})

# -- helpers -------------------------------------------------------------

# Keep only numeric columns, drop zero-variance columns, and return as matrix
.prepare_numeric_matrix <- function(X) {
  Xnum <- X[, vapply(X, is.numeric, logical(1L)), drop = FALSE]
  if (ncol(Xnum) == 0L) stop("No numeric columns in 'X'.")
  # drop zero-variance columns (computed with pairwise NAs removed)
  nzv <- vapply(Xnum, function(col) {
    v <- var(col, na.rm = TRUE)
    is.finite(v) && v > 0
  }, logical(1L))
  Xnum <- Xnum[, nzv, drop = FALSE]
  if (ncol(Xnum) == 0L) stop("All numeric columns have zero variance.")
  as.matrix(Xnum)
}

# Safe correlation matrix (pairwise) with optional method
.safe_cor <- function(Xmat, method = c("pearson","spearman")) {
  method <- match.arg(method)
  S <- suppressWarnings(cor(Xmat, use = "pairwise.complete.obs", method = method))
  # ensure symmetry and unit diagonal (occasionally numeric drift)
  S <- (S + t(S)) / 2
  diag(S) <- 1
  S
}

# Extract absolute upper-triangle values
.upper_abs <- function(S) {
  abs(S[upper.tri(S, diag = FALSE)])
}

# Effective dimensionality from eigenvalues of correlation matrix
.effective_dim <- function(S) {
  e <- eigen(S, symmetric = TRUE, only.values = TRUE)$values
  # Guard against tiny negative eigenvalues from numeric issues
  e <- pmax(e, 0)
  # probabilities
  pe <- e / sum(e)
  # add small epsilon in log to avoid log(0)
  eps <- 1e-12
  erank <- exp(-sum(pe * log(pe + eps)))
  ENF   <- (sum(e)^2) / sum(e^2)  # for a correlation matrix sum(e)=p
  list(erank = as.numeric(erank),
       ENF   = as.numeric(ENF),
       eigenvalues = e)
}

# Make a correlation matrix "near" SPD if solve() fails
.make_spd_corr <- function(S) {
  # nearPD with corr=TRUE projects to nearest correlation matrix (unit diag)
  as.matrix(nearPD(S, corr = TRUE)$mat)
}

# VIF from correlation matrix: VIF_j = (R^{-1})_{jj}
.vif_from_corr <- function(S) {
  S <- (S + t(S)) / 2
  diag(S) <- 1
  inv_ok <- try(solve(S), silent = TRUE)
  if (inherits(inv_ok, "try-error")) {
    # Stabilize via nearest SPD correlation
    S_spd <- .make_spd_corr(S)
    inv_ok <- try(solve(S_spd), silent = TRUE)
    if (inherits(inv_ok, "try-error")) {
      # last-ditch ridge, then renormalize to correlation
      p <- ncol(S)
      S_ridge <- cov2cor(S + diag(1e-4, p))
      inv_ok  <- solve(S_ridge)
    }
  }
  diag(inv_ok)
}

# Graph metrics at one threshold tau (|r| >= tau)
.graph_metrics_once <- function(S, tau) {
  p <- ncol(S)
  A <- (abs(S) >= tau) * 1
  diag(A) <- 0L
  g <- graph_from_adjacency_matrix(A, mode = "undirected", diag = FALSE)
  dens  <- edge_density(g, loops = FALSE)
  deg   <- degree(g)
  avg_d <- if (length(deg)) mean(deg) else 0
  nE    <- gsize(g)
  nC    <- components(g)$no
  # Number of maximal cliques and largest clique size
  n_max_cliques <- count_max_cliques(g)
  max_clique_sz <- clique_num(g)
  trans         <- transitivity(g, type = "global") # clustering coefficient
  data.frame(
    tau = tau,
    p = p,
    edges = nE,
    density = as.numeric(dens),
    avg_degree = as.numeric(avg_d),
    n_components = as.integer(nC),
    n_max_cliques = as.integer(n_max_cliques),
    largest_clique = as.integer(max_clique_sz),
    transitivity = as.numeric(ifelse(is.na(trans), 0, trans))
  )
}

# -- main function -------------------------------------------------------

#' Redundancy diagnostics for a numeric matrix/data frame
#'
#' @param X data.frame or matrix of features (numeric columns will be used)
#' @param method "pearson" (default) or "spearman" for correlation
#' @param tau_grid vector of thresholds for |r| to build graphs (default c(.6,.7,.8,.9))
#' @return a list with: cor_summary, graph_metrics, effective_dim, VIF
redundancy_diagnostics <- function(X, method = c("pearson","spearman"),
                                   tau_grid = c(0.6, 0.7, 0.8, 0.9)) {
  method <- match.arg(method)
  Xmat   <- .prepare_numeric_matrix(X)
  S      <- .safe_cor(Xmat, method = method)
  p      <- ncol(S)

  up_abs <- .upper_abs(S)
  MAC    <- mean(up_abs, na.rm = TRUE)
  q90    <- as.numeric(quantile(up_abs, 0.90, na.rm = TRUE))
  q95    <- as.numeric(quantile(up_abs, 0.95, na.rm = TRUE))
  maxabs <- max(up_abs, na.rm = TRUE)

  cor_summary <- list(
    p = p,
    MAC = MAC,
    q90_abs_r = q90,
    q95_abs_r = q95,
    max_abs_r = maxabs
  )

  # Graph metrics over thresholds
  graph_metrics <- do.call(rbind, lapply(tau_grid, function(tau) .graph_metrics_once(S, tau)))

  # Effective dimensionality
  eff <- .effective_dim(S)

  # VIFs
  vif_vals <- .vif_from_corr(S)
  names(vif_vals) <- colnames(S)
  vif_summary <- list(
    mean  = mean(vif_vals, na.rm = TRUE),
    median= median(vif_vals, na.rm = TRUE),
    p90   = as.numeric(quantile(vif_vals, 0.90, na.rm = TRUE)),
    max   = max(vif_vals, na.rm = TRUE),
    VIF   = vif_vals
  )

  list(
    cor_summary   = cor_summary,
    graph_metrics = graph_metrics,
    effective_dim = eff,
    VIF_summary   = vif_summary,
    corr_matrix   = S # included for optional downstream inspection
  )
}

# -- convenience: compare two feature sets side-by-side ------------------

#' Compare two feature matrices using the diagnostics above
#' @param X1, X2 numeric data frames/matrices (possibly different columns)
#' @param names character vector of length 2 for labeling
compare_redundancy <- function(X1, X2, names = c("Set1","Set2"),
                               method = "pearson",
                               tau_grid = c(0.6,0.7,0.8,0.9)) {
  if (length(names) != 2L) stop("'names' must be length 2.")
  
  d1 <- redundancy_diagnostics(X1, method = method, tau_grid = tau_grid)
  d2 <- redundancy_diagnostics(X2, method = method, tau_grid = tau_grid)

  sum_tbl <- data.frame(
    set = names,
    p   = c(d1$cor_summary$p, d2$cor_summary$p),
    MAC = c(d1$cor_summary$MAC, d2$cor_summary$MAC),
    q90 = c(d1$cor_summary$q90_abs_r, d2$cor_summary$q90_abs_r),
    q95 = c(d1$cor_summary$q95_abs_r, d2$cor_summary$q95_abs_r),
    max_abs_r = c(d1$cor_summary$max_abs_r, d2$cor_summary$max_abs_r),
    ERANK = c(d1$effective_dim$erank, d2$effective_dim$erank),
    ENF   = c(d1$effective_dim$ENF,   d2$effective_dim$ENF),
    VIF_mean   = c(d1$VIF_summary$mean,  d2$VIF_summary$mean),
    VIF_median = c(d1$VIF_summary$median,d2$VIF_summary$median),
    VIF_p90    = c(d1$VIF_summary$p90,   d2$VIF_summary$p90),
    VIF_max    = c(d1$VIF_summary$max,   d2$VIF_summary$max),
    row.names = NULL
  )

  gm1 <- transform(d1$graph_metrics, set = names[1])
  gm2 <- transform(d2$graph_metrics, set = names[2])
  gm  <- rbind(gm1, gm2)

  details <- setNames(list(d1, d2), names)

  list(
    summary_table = sum_tbl,
    graph_metrics = gm,
    details       = details
  )
}


# ============================
# Example usage
# ============================
# Suppose:
#  - X_pairwise: data.frame/matrix of the features selected by your pairwise test
#  - X_RFA     : data.frame/matrix of the features selected by RFA
#
# Run diagnostics for one set:
diag_pairwise <- redundancy_diagnostics(X_pairwise, method = "pearson", tau_grid = c(0.6,0.7,0.8,0.9))
diag_RFA      <- redundancy_diagnostics(X_RFA,      method = "pearson", tau_grid = c(0.6,0.7,0.8,0.9))
#
# Compare sets side-by-side:
cmp <- compare_redundancy(X_pairwise, X_RFA, names = c("Pairwise","RFA"))
cmp$summary_table          # quick numeric comparison
cmp$graph_metrics          # per-threshold graph stats (density, avg_degree, cliques, etc.)
#
# Inspect VIFs (features with high multicollinearity):
sort(diag_pairwise$VIF_summary$VIF, decreasing = TRUE)[1:15]
sort(diag_RFA$VIF_summary$VIF,      decreasing = TRUE)[1:15]
#
# Effective dimensionality:
diag_pairwise$effective_dim$erank
diag_pairwise$effective_dim$ENF

```

# Random Forest
### Helper functions for RF
```{r , echo=FALSE, warning = FALSE, message=FALSE}
pairwise_significance_tests <- function(input_df, 
                                        group_col,
                                        start_col_index){
  # For each feature from start_col_index onward,
  #   for each pair of group categories,
  #     check if both groups have >=3 data points
  #     do Shapiro test for normality
  #     if not normal => Mann-Whitney (wilcox.test)
  #     else => t.test
  # Collect p-values and do multiple corrections
  
  df_results <- data.frame(Feature=character(), comparison_pair=character(), pval=numeric())
  
  # get unique group combos
  group_pairs <- utils::combn(unique(as.character(input_df[[group_col]])), 2)

  for(feature_col in start_col_index:ncol(input_df)){
    for (col in 1:ncol(group_pairs)){
      p_1 <- group_pairs[1,col]
      p_2 <- group_pairs[2,col]

      vec1 <- as.numeric(unlist(input_df[which(input_df[[group_col]] == p_1), feature_col]))
      vec2 <- as.numeric(unlist(input_df[which(input_df[[group_col]] == p_2), feature_col]))
      
      # ensure at least 3 non-NA data points
      if(sum(!is.na(vec1)) >= 3 && sum(!is.na(vec2)) >= 3){
        # skip if all values are the same
        # if(all(vec1 == vec1[1]) || all(vec2 == vec2[1])){
        #   next
        # }
        
        # normality tests
        p_shapiro1 <- tryCatch(shapiro.test(vec1)$p.value, error = function(e) NA)
        p_shapiro2 <- tryCatch(shapiro.test(vec2)$p.value, error = function(e) NA)
        
        # decide parametric vs non-parametric
        if(is.na(p_shapiro1) || is.na(p_shapiro2)){
          # if for some reason test failed, skip
          next
        }
        
        if(p_shapiro1 < 0.05 || p_shapiro2 < 0.05){
          # mann-whitney
          pval_test <- tryCatch(wilcox.test(vec1, vec2)$p.value, error = function(e) NA)
        } else {
          # t-test
          pval_test <- tryCatch(t.test(vec1, vec2, var.equal = FALSE)$p.value, error = function(e) NA)
        }
        
        if(!is.na(pval_test)){
          # append to df
          df_results <- rbind(df_results, data.frame(
            Feature = colnames(input_df)[feature_col],
            Category_1 = p_1,
            Category_2 = p_2,
            pval = as.numeric(pval_test),
            stringsAsFactors = FALSE
          ))
        }
      }
    }
  }
  
  # multiple corrections
  pvalue_correction_methods <- "bonferroni" #, c("holm", "hochberg", "hommel", "BH", "BY")
  
  # For convenience, store all results in a list
  # results_list <- list()
  # j <- 1
  # for(method in pvalue_correction_methods){
    temp <- df_results
    temp$adjusted_pvalue <- p.adjust(temp$pval, method = pvalue_correction_methods)
    significant <- temp %>% dplyr::filter(adjusted_pvalue < 0.05) %>% arrange(adjusted_pvalue)
    # results_list[[j]] <- significant
    
    if(nrow(significant) == 0){
      cat(paste0(pvalue_correction_methods, " did not result in any significant compound\n"))
    } else {
      # cat(paste0("Significant results for method: ", method, "\n"))
      # print(significant)
      cat(paste0(pvalue_correction_methods, " has ", length(unique(significant$Feature)), " significant features\n"))
      # j <- j + 1
    }
  # }
  
  return(list(
    uncorrected = df_results,
    corrected   = unique(significant$Feature)  # results_list
  ))
}

#─────────────────────────────────────────────────────────────────────────────
# Helper: retune mtry (based on subset size) + ntree for any train/test split
#─────────────────────────────────────────────────────────────────────────────
# Helper: retune mtry (based on subset size) + ntree for any train/test split
tune_rf_subset <- function(X_train, y_train, X_test, y_test,
                           ntree_candidates, metric = "Accuracy",
                           parallel = TRUE, n_cores = NULL,
                           num_threads = 1) {
  
  y_train <- as.factor(y_train)
  y_test  <- as.factor(y_test)
  
  k_inner <- min(table(y_train))
  
  p         <- ncol(X_train)
  mtry_vals <- unique(pmax(1, c(floor(sqrt(p)), floor(log2(p)))))
  
  ranger_predict_probs <- function(rf, newdata) {
    pr <- predict(rf, data = newdata)$predictions
    if (is.matrix(pr)) {
      prob_mat <- pr
      pred_chr <- colnames(prob_mat)[max.col(prob_mat, ties.method = "first")]
      pred_fac <- factor(pred_chr, levels = colnames(prob_mat))
      return(list(preds = pred_fac, probs = prob_mat))
    }
    pred_fac <- factor(pr, levels = levels(y_train))
    return(list(preds = pred_fac, probs = NULL))
  }
  
  # Choose cores
  if (is.null(n_cores)) {
    n_cores <- max(1, parallel::detectCores() - 1)
  }
  
  # ---- Branch A: enough data for CV ----
  if (k_inner > 1) {
    
    ctrl <- caret::trainControl(
      method          = "cv",
      number          = k_inner,
      classProbs      = TRUE,
      summaryFunction = defaultSummary,
      savePredictions = "final",
      allowParallel   = parallel
    )
    
    rf_grid <- expand.grid(
      mtry          = mtry_vals,
      splitrule     = "gini",
      min.node.size = 1
    )
    
    best_mod   <- NULL
    best_val   <- -Inf
    best_trees <- NA
    
    # If you parallelize caret resamples, keep ranger internal threads modest
    ranger_threads_for_caret <- if (parallel) 1 else num_threads
    
    for (nt in ntree_candidates) {
      set.seed(123)
      
      # ERROR FIX: 'probability = TRUE' removed below because caret handles it automatically
      tmp <- caret::train(
        x           = X_train,
        y           = y_train,
        method      = "ranger",
        trControl   = ctrl,
        tuneGrid    = rf_grid,
        metric      = metric,
        num.trees   = nt,
        importance  = "impurity",
        # probability = TRUE,  <-- REMOVED THIS LINE
        num.threads = ranger_threads_for_caret
      )
      
      this_val <- suppressWarnings(max(tmp$results[[metric]], na.rm = TRUE))
      if (is.finite(this_val) && this_val > best_val) {
        best_val   <- this_val
        best_mod   <- tmp
        best_trees <- nt
      }
    }
    
    caret_model <- best_mod
    final_rf    <- caret_model$finalModel
    
  } else {
    # ---- Branch B: too few per class -> OOB tuning grid ----
    
    train_df <- data.frame(.Class = y_train, X_train, check.names = FALSE)
    grid <- expand.grid(ntree = ntree_candidates, mtry = mtry_vals)
    
    # Parallel backend
    created_cluster <- FALSE
    if (parallel) {
      if (!foreach::getDoParRegistered() || foreach::getDoParWorkers() < 2) {
        cl <- parallel::makeCluster(n_cores)
        doParallel::registerDoParallel(cl)
        created_cluster <- TRUE
      }
    }
    
    on.exit({
      if (parallel && created_cluster) {
        try(parallel::stopCluster(cl), silent = TRUE)
        foreach::registerDoSEQ()
      }
    }, add = TRUE)
    
    eval_one <- function(nt, m) {
      set.seed(123)
      rf_mod <- ranger::ranger(
        dependent.variable.name = ".Class",
        data         = train_df,
        num.trees    = nt,
        mtry         = m,
        splitrule    = "gini",
        importance   = "impurity",
        probability  = FALSE,
        oob.error    = TRUE,
        num.threads  = 1  # important to avoid nested parallel
      )
      data.frame(
        ntree = nt,
        mtry  = m,
        oob_err = rf_mod$prediction.error
      )
    }
    
    if (parallel) {
      res_grid <- foreach::foreach(i = seq_len(nrow(grid)),
                                   .combine = rbind,
                                   .packages = "ranger") %dopar% {
                                     eval_one(grid$ntree[i], grid$mtry[i])
                                   }
    } else {
      res_grid <- do.call(rbind, lapply(seq_len(nrow(grid)), function(i) {
        eval_one(grid$ntree[i], grid$mtry[i])
      }))
    }
    
    res_grid <- res_grid[is.finite(res_grid$oob_err), , drop = FALSE]
    if (nrow(res_grid) == 0) {
      return(list(
        model = NULL, ntree = NA, preds = factor(NA),
        p_m = matrix(NA, nrow = 0, ncol = 0),
        eval_metrics = list(), Accuracy = NA, Kappa = NA,
        AUC = NA, F1_Macro = NA, F1_Weighted = NA, MCC = NA
      ))
    }
    
    best_row <- res_grid[which.min(res_grid$oob_err), ]
    best_trees <- best_row$ntree
    best_mtry  <- best_row$mtry
    
    set.seed(123)
    best_rf <- ranger::ranger(
      dependent.variable.name = ".Class",
      data         = train_df,
      num.trees    = best_trees,
      mtry         = best_mtry,
      splitrule    = "gini",
      importance   = "impurity",
      probability  = TRUE,
      oob.error    = TRUE,
      num.threads  = num_threads
    )
    
    caret_model <- list(
      finalModel = best_rf,
      bestTune   = data.frame(mtry = best_mtry, splitrule = "gini", min.node.size = 1)
    )
    final_rf <- best_rf
  }
  
  # ---- Predict & score ----
  pred_out <- ranger_predict_probs(final_rf, X_test)
  preds    <- pred_out$preds
  prob_mat <- pred_out$probs
  
  if (is.null(prob_mat)) {
    cls <- levels(y_train)
    prob_mat <- matrix(0, nrow = length(preds), ncol = length(cls))
    colnames(prob_mat) <- cls
    rownames(prob_mat) <- rownames(X_test)
    prob_mat[cbind(seq_along(preds), match(as.character(preds), cls))] <- 1
  }
  
  ev <- eval_metrics(true_labels = y_test,
                     prob_matrix = prob_mat,
                     pred_labels = preds)
  
  list(
    model        = caret_model,
    ntree        = best_trees,
    preds        = preds,
    p_m          = prob_mat,
    eval_metrics = ev,
    Accuracy     = ev$Accuracy,
    Kappa        = ev$Kappa,
    AUC          = ev$AUC,
    F1_Macro     = ev$F1_Macro,
    F1_Weighted  = ev$F1_Weighted,
    MCC          = ev$MCC_Multiclass
  )
}

# Recursive Feature Addition -------------
recursive_feature_addition <- function(X_train, y_train,
                                       X_test,  y_test,
                                       ntree_candidates,
                                       metric = "Accuracy",
                                       parallel = TRUE, n_cores = NULL,
                                       num_threads_model = 1) {
  
  if (is.null(n_cores)) {
    n_cores <- max(1, parallel::detectCores() - 1)
  }
  
  # 1) Fit full model to get initial importance
  full_res <- tune_rf_subset(
    X_train, y_train,
    X_test,  y_test,
    ntree_candidates, metric,
    parallel = parallel,
    n_cores  = n_cores,
    num_threads = num_threads_model
  )
  
  imp_vec <- full_res$model$finalModel$variable.importance
  init_feat <- names(imp_vec)[which.max(imp_vec)]
  
  best_feats <- init_feat
  remaining  <- setdiff(colnames(X_train), best_feats)
  
  # 2) Baseline performance
  base_res <- tune_rf_subset(
    X_train[, best_feats, drop = FALSE], y_train,
    X_test [, best_feats, drop = FALSE], y_test,
    ntree_candidates, metric,
    parallel = FALSE,          # avoid nested parallel
    num_threads = num_threads_model
  )
  best_mcc <- base_res$MCC
  
  # Parallel backend for feature-try loop
  created_cluster <- FALSE
  if (parallel) {
    if (!foreach::getDoParRegistered() || foreach::getDoParWorkers() < 2) {
      cl <- parallel::makeCluster(n_cores)
      doParallel::registerDoParallel(cl)
      created_cluster <- TRUE
    }
  }
  
  on.exit({
    if (parallel && created_cluster) {
      try(parallel::stopCluster(cl), silent = TRUE)
      foreach::registerDoSEQ()
    }
  }, add = TRUE)
  
  # 3) Iterative greedy addition
  while (length(remaining) > 0) {
    
    # Evaluate all candidates in parallel processing to optimize computational time
    if (parallel) {
      cand_res <- foreach::foreach(feat = remaining,
                                   .combine = rbind,
                                   # FIX IS HERE: Added .export
                                   .export = c("tune_rf_subset", "eval_metrics"), 
                                   .packages = c("ranger", "caret", "pROC", "dplyr")) %dopar% {
                                     
                                     trial_feats <- c(best_feats, feat)
                                     
                                     # Now the worker node can find this function:
                                     res <- tune_rf_subset(
                                       X_train[, trial_feats, drop = FALSE], y_train,
                                       X_test [, trial_feats, drop = FALSE], y_test,
                                       ntree_candidates, metric,
                                       parallel = FALSE,      # critical
                                       num_threads = 1        # critical inside workers
                                     )
                                     data.frame(feat = feat, MCC = res$MCC, stringsAsFactors = FALSE)
                                   }
    } else {
      cand_res <- do.call(rbind, lapply(remaining, function(feat) {
        trial_feats <- c(best_feats, feat)
        res <- tune_rf_subset(
          X_train[, trial_feats, drop = FALSE], y_train,
          X_test [, trial_feats, drop = FALSE], y_test,
          ntree_candidates, metric,
          parallel = FALSE,
          num_threads = num_threads_model
        )
        data.frame(feat = feat, MCC = res$MCC, stringsAsFactors = FALSE)
      }))
    }
    
    cand_res <- cand_res[is.finite(cand_res$MCC), , drop = FALSE]
    if (nrow(cand_res) == 0) break
    
    # Pick best improvement
    best_cand <- cand_res[which.max(cand_res$MCC), , drop = FALSE]
    
    if (best_cand$MCC > best_mcc) {
      best_mcc  <- best_cand$MCC
      best_feats <- c(best_feats, best_cand$feat)
      remaining  <- setdiff(remaining, best_cand$feat)
    } else {
      break
    }
  }
  
  # 4) Final model
  final_res <- tune_rf_subset(
    X_train[, best_feats, drop = FALSE], y_train,
    X_test [, best_feats, drop = FALSE], y_test,
    ntree_candidates, metric,
    parallel = FALSE,
    num_threads = num_threads_model
  )
  
  list(
    best_mcc         = final_res$MCC,
    best_acc         = final_res$Accuracy,
    best_features_rf = best_feats,
    final_model      = final_res$model,
    prob_matrix      = final_res$p_m,
    predictions      = final_res$preds,
    eval_metrics     = final_res$eval_metrics
  )
}

#─────────────────────────────────────────────────────────────────────────────
library(caret)
library(pROC)

eval_metrics <- function(true_labels, prob_matrix, pred_labels){
  # 0) Ensure factor levels align
  classes <- union(levels(true_labels), levels(pred_labels))
  y_true <- factor(true_labels, levels = classes)
  y_pred <- factor(pred_labels,  levels = classes)
  
  # 1) Build the confusion‐matrix table
  cm <- table(Actual = y_true, Predicted = y_pred)
  N  <- sum(cm)                       # total samples
  diag_cm <- diag(cm)
  
  # 2) Accuracy
  accuracy <- sum(diag_cm) / N
  
  # 3) Cohen’s Kappa (via caret)
  kappa <- tryCatch({
    cm_obj <- confusionMatrix(y_pred, y_true)
    as.numeric(cm_obj$overall["Kappa"])
  }, error = function(e) NA)
  
  # 4) One‐vs‐all multiclass AUC
  auc <- tryCatch({
    roc_obj <- pROC::multiclass.roc(response  = true_labels,
                                    predictor = prob_matrix)
    as.numeric(roc_obj$auc)
  }, error = function(e) NA)
  
  # 5) Per‐class precision & recall
  col_sums <- colSums(cm)   # predicted totals for each class
  row_sums <- rowSums(cm)   # actual   totals for each class
  
  precision_per_class <- ifelse(col_sums > 0,
                                diag_cm / col_sums,
                                0)        # no preds → p=0
  recall_per_class    <- ifelse(row_sums > 0,
                                diag_cm / row_sums,
                                0)        # no actual → r=0
  
  # 6) Per‐class F1 & weighted F1
  f1_per_class <- ifelse(
    (precision_per_class + recall_per_class) > 0,
    2 * precision_per_class * recall_per_class /
      (precision_per_class + recall_per_class),
    0
  )
  support <- row_sums   # how many true examples of each class
  f1_weighted <- sum(f1_per_class * support) / sum(support)
  
  # 7) Multiclass MCC (Gorodkin’s formula)
  t_i <- row_sums
  p_j <- col_sums
  c_val <- sum(diag_cm)         # sum of true positives
  num   <- N * c_val - sum(t_i * p_j)
  den   <- sqrt((N^2 - sum(t_i^2)) * (N^2 - sum(p_j^2)))
  mcc   <- ifelse(den > 0, num / den, 0)
  
  return(list(
    Accuracy         = accuracy,
    Kappa            = kappa,
    AUC              = auc,
    F1_Weighted      = f1_weighted,
    MCC_Multiclass   = mcc
  ))
}


####################################################
##########        CONFUSION MATRIX       ###########
####################################################
conf_mat_plot <- function(y_test, preds, conf_mat_title, accuracy) {
  # 1) Compute raw counts table: rows = treated labels, cols = predicted pristine labels
  all_classes <- unique(c(levels(preds), levels(y_test)))
  y_f <- factor(y_test,  levels = all_classes)
  p_f <- factor(preds,  levels = all_classes)
  
  cm_tab <- table(Actual = y_f, Predicted = p_f)
  
  # 2) Turn it into a data.frame for ggplot
  cm_df <- as.data.frame(cm_tab, stringsAsFactors = FALSE) %>%
    dplyr::rename(Freq = Freq)
  
  # 3) Add total-per-row and percent-per-cell
  cm_df$Total <- NA
  cm_df$Percent <- NA
  for (label in unique(cm_df$Actual)){
    cm_df[which(cm_df$Actual == label),]$Total <- sum(cm_df[which(cm_df$Actual == label),]$Freq)
  }
  
  cm_df <- cm_df %>%
    dplyr::mutate(
      Percent = round(Freq / Total * 100, 2)
    ) %>% 
    dplyr::mutate(Percent = ifelse(is.nan(Percent), 0, Percent))
  
  
  cm_df <- cm_df %>%
    dplyr::mutate(
      Label = ifelse(Predicted == Actual,
                     paste0(round(Percent,1),"%"),
                     ""
      )
    )
  
  # 5) Plot
  conf_mat <- ggplot(cm_df, aes(x = Predicted, y = Actual, fill = Percent)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "white", high = "steelblue") + #, limits = c(0,1)) +
    geom_text(aes(label = Percent), size = 6, show.legend = FALSE) +
    # explicitly drop any size guide, keep only the fill (Recall) legend
    guides(size = "none") +
    labs(
      title = paste0("Confusion Matrix of ENV → SB Mapping | ", conf_mat_title),
      subtitle = paste0(sprintf("Overall linking accuracy: %.2f%%\n", accuracy)),
      x = "Predicted Class (Store-Bought)",
      y = "Actual Class (Environmental)",
      fill = "Percentage of\nClassification (%)" # "Recall %"
    ) +
    theme_minimal(base_size = 22) +
    theme(
      axis.text.x  = element_text(angle = 45, hjust = 1),
      plot.title   = element_text(face = "bold", hjust = 0.5),
      plot.subtitle= element_text(hjust = 0.5)
    )
  return(list(conf_mat = conf_mat, cm_df = cm_df))
}
```

### RF classifier
```{r , echo=FALSE, warning = FALSE, message=FALSE}
run_rf_analysis_manuscript1 <- function(data,
                                        type_col,
                                        remove_cols,
                                        train_proportion,
                                        use_source_split = FALSE,
                                        use_store_vs_environmental_split = FALSE,
                                        use_polymer = FALSE,
                                        do_pairwise_test = FALSE,
                                        do_top_importance_selection = FALSE,
                                        top_importance_counts = c(100, 50, 25, 10),
                                        data_name,
                                        seed = 123,
                                        ntree_candidates = c(100, 500, 1000, 2500),
                                        metric = "Accuracy",
                                        do_rfe = FALSE,
                                        do_rfa = FALSE,
                                        min_sample_number,
                                        do_impute_norm_screen = TRUE,
                                        group_for_significance,
                                        K_outer_max = 5,
                                        ntree_screen = 500) {
  
  set.seed(seed)
  
  # ---- 1) Clean classes ----
  data[[type_col]] <- as.factor(data[[type_col]])
  levels(data[[type_col]]) <- make.names(levels(data[[type_col]]), unique = TRUE)
  
  class_counts <- table(data[[type_col]])
  keep_classes <- names(class_counts[class_counts >= min_sample_number])
  data <- data[data[[type_col]] %in% keep_classes, ]
  data[[type_col]] <- droplevels(data[[type_col]])
  
  rownames(data) <- NULL
  
  data <- data %>% 
    mutate(sample_id = paste0(Plastic_type, "_", technique)) %>% 
    mutate(sample_id = ave(sample_id, sample_id, FUN = function(x) {
      if (length(x) > 1) paste0(x, "_rep", seq_along(x)) else x
    })) %>% 
    tibble::column_to_rownames(var = "sample_id")
  
  # ---- 2) Build OUTER folds first ----
  outer_folds <- list()
  
  if(use_store_vs_environmental_split){
    store_bought_set  <- data[data$Source == "Store-Bought", ]
    environmental_set <- data[data$Source == "Environmental", ]
    
    # Outer folds on STORE-Bought only (training stability)
    sb_y <- store_bought_set[[type_col]]
    min_sb <- min(table(as.character(sb_y)))
    K <- max(2, min(K_outer_max, min_sb))
    
    sb_test_folds <- caret::createFolds(sb_y, k = K, returnTrain = FALSE)
    
    for(f in seq_len(K)){
      test_idx_sb  <- sb_test_folds[[f]]
      train_idx_sb <- setdiff(seq_len(nrow(store_bought_set)), test_idx_sb)
      
      outer_folds[[f]] <- list(
        train_data = store_bought_set[train_idx_sb, , drop = FALSE],
        test_data  = environmental_set  # fixed test
      )
    }
    
  } else if(use_source_split){
    store_bought_set  <- data[data$Source == "Store-Bought", ]
    environmental_set <- data[data$Source == "Environmental", ]
    
    env_y <- environmental_set[[type_col]]
    min_env <- min(table(as.character(env_y)))
    K <- max(2, min(K_outer_max, min_env))
    
    env_test_folds <- caret::createFolds(env_y, k = K, returnTrain = FALSE)
    
    for(f in seq_len(K)){
      test_idx_env  <- env_test_folds[[f]]
      train_idx_env <- setdiff(seq_len(nrow(environmental_set)), test_idx_env)
      
      outer_folds[[f]] <- list(
        train_data = rbind(store_bought_set,
                           environmental_set[train_idx_env, , drop = FALSE]),
        test_data  = environmental_set[test_idx_env, , drop = FALSE]
      )
    }
    
  } else {
    y <- data[[type_col]]
    min_all <- min(table(y))
    K <- max(2, min(K_outer_max, min_all))
    cat(paste0("min_all is: ", min_all, "_Number of fold: ", K))
    
    test_folds <- caret::createFolds(y, k = K, returnTrain = FALSE)
    
    for(f in seq_len(K)){
      test_idx  <- test_folds[[f]]
      train_idx <- setdiff(seq_len(nrow(data)), test_idx)
      
      outer_folds[[f]] <- list(
        train_data = data[train_idx, , drop = FALSE],
        test_data  = data[test_idx,  , drop = FALSE]
      )
    }
  }
  
  # ---- 3) Fit/evaluate per OUTER fold ----
  final_final_res_list <- vector("list", length = length(outer_folds))

  for(f in seq_along(outer_folds)){
    
    train_data <- outer_folds[[f]]$train_data
    test_data  <- outer_folds[[f]]$test_data
    
    train_data[[type_col]] <- droplevels(train_data[[type_col]])
    test_data[[type_col]]  <- droplevels(test_data[[type_col]])
    
    # --- 3a) Impute+Norm screening on TRAIN only (OOB MCC) ---
    if(do_impute_norm_screen){
      t0 <- Sys.time()
      
      best_res_imp_norm <- find_best_impute_normalize(
        df_train     = train_data,
        type_col     = type_col,
        remove_cols  = remove_cols,
        ntree_screen = ntree_screen
      )
      
      best_imp  <- best_res_imp_norm$best_combo$Imputation
      best_norm <- best_res_imp_norm$best_combo$Normalization
      
      numeric_cols <- names(which(sapply(train_data, is.numeric)))
      numeric_cols <- setdiff(numeric_cols, c(remove_cols, type_col))
      
      X_train_orig <- train_data[, numeric_cols, drop = FALSE]
      X_test_orig  <- test_data[,  numeric_cols, drop = FALSE]
      
      pp_final <- impute_norm_train_apply(X_train_orig, X_test_orig, best_imp, best_norm)
      
      train_data[, numeric_cols] <- pp_final$train
      test_data[,  numeric_cols] <- pp_final$test
      
      t1 <- Sys.time()
      time_imp_norm <- as.numeric(difftime(t1, t0, units = "secs"))
      
    } else {
      best_res_imp_norm <- NULL
      best_imp <- NA_character_
      best_norm <- NA_character_
      time_imp_norm <- NA_real_
    }
    
    # --- 3c) Build X/y ---
    X_train <- train_data %>% dplyr::select(-c(all_of(remove_cols), all_of(type_col)))
    y_train <- train_data[[type_col]]
    X_test  <- test_data  %>% dplyr::select(-c(all_of(remove_cols), all_of(type_col)))
    y_test  <- test_data[[type_col]]
    
    # --- 3d) Full features ---
    cat("\n### Train/test with Full original features ###\n")
    start_time_full_features <- Sys.time()
    
    all_feats <- tune_rf_subset(X_train, y_train, X_test, y_test,
                                ntree_candidates, metric)
    
    best_model_all_feats <- all_feats$model
    mcc_full <- all_feats$MCC
    eval_metrics_all_feats <- all_feats$eval_metrics
    # --- Generate Heatmap (Full Features) ---
    prob_matrix_all_feats <- all_feats$p_m
    conf_mat <- conf_mat_plot(y_test, all_feats$preds,
                              conf_mat_title = paste0("All Features - ", data_name),
                              accuracy = mcc_full)
    conf_mat_all_feats <- conf_mat$conf_mat
    cm_df_all_feats <- conf_mat$cm_df
    end_time_full_features   <- Sys.time()
    time_all_feats       <- as.numeric(difftime(end_time_full_features,
                                                start_time_full_features,
                                                units="secs"))

    # --- 3e) Pairwise features (if any) ---
    if(do_pairwise_test){
      cat("\n### Train/test with Pair-wise features ###\n")
      start_time_sig <- Sys.time()
      pairwise_res <- pairwise_significance_tests(
        input_df = train_data,
        group_col = group_for_significance,
        start_col_index = 6
      )
      sig_feats <- pairwise_res$corrected
      
      if(length(sig_feats) > 0){
        # Case A: Features found - Run RF
        tmp_sig <- tune_rf_subset(
          X_train[, sig_feats, drop = FALSE], y_train,
          X_test[,  sig_feats, drop = FALSE], y_test,
          ntree_candidates, metric
        )
        mcc_sig <- tmp_sig$MCC
        eval_metrics_sig <- tmp_sig$eval_metrics
        
        prob_matrix_sig <- tmp_sig$p_m
        conf_mat <- conf_mat_plot(y_test, tmp_sig$preds,
                                  conf_mat_title = paste0("Pair-wise Significance-Based Features - ", data_name),
                                  accuracy = mcc_sig)
        conf_mat_sig <- conf_mat$conf_mat
        cm_df_sig <- conf_mat$cm_df
        
        # Save predictions to a common variable
        preds_sig <- tmp_sig$preds
        
      } else {
        # Case B: No features found - Initialize empty placeholders
        mcc_sig <- NA_real_
        eval_metrics_sig <- NULL
        prob_matrix_sig <- NULL
        conf_mat_sig <- NULL
        cm_df_sig <- NULL
        preds_sig <- NULL # Crucial: prevents "object 'tmp_sig' not found" later
      }
      end_time_sig   <- Sys.time()
      time_sig       <- as.numeric(difftime(end_time_sig, start_time_sig, units="secs"))
    }
    else {
      # If do_pairwise_test is FALSE entirely
      pairwise_sig_features <- character(0)
      sig_feats <- intersect(pairwise_sig_features, colnames(X_train))
    }
    
    # --- 3f) RFA / Top-N, etc. (your existing blocks can follow) ---
    # Recursive Feature Addition
    if(do_rfa){
      cat("\n### Train/test with RFA features ###\n")
      start_time_rfa <- Sys.time()
      
      rfa_res   <- recursive_feature_addition(
        X_train, y_train,
        X_test,  y_test,
        ntree_candidates = ntree_candidates,
        metric           = metric
      )
      
      rfa_selected_feats <- rfa_res$best_features_rf
      mcc_rfa                 <- rfa_res$best_mcc
      final_rf_rfa            <- rfa_res$final_model$finalModel
      
      prob_matrix_rfa <- rfa_res$prob_matrix
      
      conf_mat <- conf_mat_plot(y_test, rfa_res$predictions,
                                conf_mat_title = paste0("RFA Features - ", type_col),
                                accuracy = mcc_rfa)
      conf_mat_rfa <- conf_mat$conf_mat
      cm_df_rfa <- conf_mat$cm_df
      
      eval_metrics_rfa <- rfa_res$eval_metrics
      
      end_time_rfa   <- Sys.time()
      time_rfa       <- as.numeric(difftime(end_time_rfa, start_time_rfa, units="secs"))
    }
    
    # 10) Classification with Top-N Features by Importance (if requested)
    if(do_top_importance_selection){
      cat("\n### Running Classification on Top-N Features by Importance ###\n")
      
      imp_vec <- best_model_all_feats$finalModel$variable.importance
      
      rf_importances <- data.frame(
        Feature    = names(imp_vec),
        Importance = as.numeric(imp_vec),
        stringsAsFactors = FALSE
      )
      
      rf_importances <- rf_importances[order(rf_importances$Importance, decreasing = TRUE), ]
      
      top_importance_results <- list()
      total_features <- nrow(rf_importances)
      
      for(top_n in top_importance_counts){
        start_time_top_100_50_25_10 <- base::Sys.time()
        
        if(total_features < top_n){
          selected_features <- rf_importances$Feature
          cat(sprintf("Only %d features available; using all.\n", total_features))
        } else {
          selected_features <- head(rf_importances$Feature, top_n)
        }
        
        tmp_top <- tune_rf_subset(
          X_train[, selected_features, drop = FALSE], y_train,
          X_test[, selected_features, drop = FALSE], y_test,
          ntree_candidates, metric
        )
        
        mcc_top         <- tmp_top$MCC
        final_rf_top <- tmp_top$model$finalModel
        
        prob_matrix_top <- tmp_top$p_m
        conf_mat <- conf_mat_plot(y_test, tmp_top$preds,
                                  conf_mat_title = paste0("Top ", length(selected_features), " Features - ", data_name),
                                  accuracy = mcc_top)
        conf_mat_top <- conf_mat$conf_mat
        cm_df_top <- conf_mat$cm_df
        
        eval_metrics_top <- tmp_top$eval_metrics
        
        end_time_top_100_50_25_10   <- Sys.time()
        time_top_100_50_25_10   <- as.numeric(difftime(end_time_top_100_50_25_10,
                                                       start_time_top_100_50_25_10,
                                                       units="secs"))
        
        top_importance_results[[paste0("top_", top_n)]] <- list(
          best_model = final_rf_top,
          selected_features = selected_features,
          preds_for_buidling_confmat_top = tmp_top$preds,
          prob_matrix_top = prob_matrix_top,
          conf_mat_top=conf_mat_top,
          eval_metrics_top = eval_metrics_top,
          cm_df_top = cm_df_top,
          time_top_100_50_25_10 = time_top_100_50_25_10
        )
      }
    }
    
    
    # --- 3g) Package fold output ---
    result_list <- list(
      fold_id = f,
      final_imp_norm_train = train_data,
      final_imp_norm_test  = test_data,
      best_model_all_feats = best_model_all_feats,
      final_rf_all_feats   = best_model_all_feats$finalModel,
      all_features_mcc     = mcc_full,
      eval_metrics_all_feats = eval_metrics_all_feats,
      prob_matrix = prob_matrix_all_feats, 
      time_all_features = time_all_feats, 
      conf_mat_all_feats = conf_mat_all_feats, 
      cm_df_all_feats = cm_df_all_feats, 
      preds_for_buidling_confmat_all = all_feats$preds
    )
    
    if(do_impute_norm_screen) {
      result_list$imp_norm_res_table <- best_res_imp_norm$results_table
      result_list$time_imp_norm <- time_imp_norm
      result_list$best_imputation     <- best_imp
      result_list$best_normalization    <- best_norm
    }
    
    if(do_pairwise_test){
      result_list$mcc_sig <- mcc_sig
      result_list$time_sig          <- time_sig
      result_list$prob_matrix_sig <- prob_matrix_sig
      result_list$eval_metrics_sig <- eval_metrics_sig
      result_list$sig_selected_feats <- sig_feats
      result_list$conf_mat_sig <- conf_mat_sig
      result_list$cm_df_sig <- cm_df_sig
      # CHANGED: Use the common variable 'preds_sig' defined above
      result_list$preds_for_buidling_confmat_sig <- preds_sig
    }
    
    if(do_rfa) {
      result_list$rfa_selected_feats <- rfa_selected_feats
      result_list$prob_matrix_rfa    <- prob_matrix_rfa
      result_list$mcc_rfa  <- mcc_rfa
      result_list$time_rfa                <- time_rfa
      result_list$eval_metrics_rfa <- eval_metrics_rfa
      result_list$conf_mat_rfa <- conf_mat_rfa
      result_list$cm_df_rfa <- cm_df_rfa
      result_list$preds_for_buidling_confmat_rfa <- rfa_res$predictions
    }
    
    if(do_top_importance_selection) {
      result_list$top_importance_results <- top_importance_results
    }
    
    # Save the result for this fold into the master list
    final_final_res_list[[f]] <- result_list
  }
  
  # -----------------------------------------------------------
  # NEW: Name the list elements as fold1, fold2, etc.
  # -----------------------------------------------------------
  names(final_final_res_list) <- paste0("fold", seq_along(final_final_res_list))
  
  return(final_final_res_list)
}
```


**UPDATE 12 DEC 2025: testing unsupervised imputation (RF impute and kNN imputation) to see if there is any difference in best imp norm between SB only train and SB_and_ENV train across 7 data combinations**

**UPDATE 13 Dec 2025: I edited the feature reduction to work more correct than before and i do feature reduction and then finding shared features between SB and ENV. => Then I investigate the imp x norm again to see if there is any crazy changes (for ex, does the small value imputation and TSN stills the best or not.)**

### Option 1: Use SB to predict ENV - Only select shared features between SB and ENV 

```{r class.source = 'fold-hide', echo=FALSE, warning = FALSE, message=FALSE}
###################################################################################
## Explanation of the parameter of run_rf_analysis()
# split environmental into 50:50 and combine with store_bought -> use_source_split = FALSE
# use store-bought as training and environmental as testing -> use_store_vs_environmental_split = TRUE
###################################################################################
# 1. Detect available cores
# Use detectCores() - 1 to leave one core free for your OS so your computer doesn't freeze.
num_cores <- parallel::detectCores(logical = FALSE) - 1  

# 2. SAFETY CHECK (Crucial for your specific error)
# Your previous error was likely due to running out of RAM. 
# Start with a SMALL number (e.g., 2 or 4) to test stability.
# If you have 32GB+ RAM, you can try higher, but for now, limit it:
# if(num_cores > 4) num_cores <- 4 

# 3. Create the cluster
cl <- parallel::makePSOCKcluster(num_cores)

# 4. Register the cluster
doParallel::registerDoParallel(cl)
cat("Parallel processing registered with", num_cores, "cores.\n")

t0 <- Sys.time()
data_combinations <- list(
  gc                 = gc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  hplc               = hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  icp                = icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  gc_hplc            = gc_hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  gc_icp             = gc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  hplc_icp           = hplc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  gc_hplc_icp        = gc_hplc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025
)

# Predefine the result list with names from data_combinations
run_rf_analysis_list_SB_train_13Dec25 <- vector("list", length(data_combinations))
names(run_rf_analysis_list_SB_train_13Dec25) <- c(
  "gc_USSB_train",
  "hplc_USSB_train",
  "icp_USSB_train",
  "gc_hplc_USSB_train",
  "gc_icp_USSB_train",
  "hplc_icp_USSB_train",
  "gc_hplc_icp_USSB_train"
  )

# Iterate through each data combination by index
for(i in seq_along(data_combinations)) {
  cat("Processing dataset:", names(data_combinations)[i], "\n")
  
  # Run the analysis using the corresponding data and pass the custom name for figure titles
  run_rf_analysis_list_SB_train_13Dec25[[names(run_rf_analysis_list_SB_train_13Dec25)[i]]] <- run_rf_analysis_manuscript1(
    data = data_combinations[[i]],
    type_col                = "Plastic_type", 
    remove_cols             = c("Source", "Polymer", "technique"),
    train_proportion = 0.8,
    use_store_vs_environmental_split = TRUE,
    data_name = names(data_combinations)[i], 
    group_for_significance = "Plastic_type", 
    do_pairwise_test = FALSE,             # Adjust parameters as needed
    do_top_importance_selection = FALSE,
    do_rfa = FALSE,
    do_impute_norm_screen = TRUE,
    seed = 123,
    min_sample_number = 1)
  
  # saveRDS(
  #   object = run_rf_analysis_list_SB_train_12Dec25[[names(run_rf_analysis_list_SB_train_12Dec25)[i]]], 
  #   file = paste0(getwd(), "RData_", names(run_rf_analysis_list_SB_train_12Dec25)[[i]], ".rds")
  # )
  
  # Force memory cleanup
  gc()
}

t1 <- Sys.time()
print(paste0("Total run time for all data combos are: ", round(as.numeric(difftime(t1, t0, units = "secs")),2), " seconds"))
stopCluster(cl)
registerDoSEQ() # Good practice to reset to sequential after you are done
```

### Option 2: Use Both SB and ENV to predict ENV

```{r class.source = 'fold-hide', echo=FALSE, warning = FALSE, message=FALSE}
# 1. Detect available cores
# Use detectCores() - 1 to leave one core free for your OS so your computer doesn't freeze.
num_cores <- parallel::detectCores(logical = FALSE) - 1  

# 2. SAFETY CHECK (Crucial for your specific error)
# Your previous error was likely due to running out of RAM. 
# Start with a SMALL number (e.g., 2 or 4) to test stability.
# If you have 32GB+ RAM, you can try higher, but for now, limit it:
# if(num_cores > 4) num_cores <- 4 

# 3. Create the cluster
cl <- parallel::makePSOCKcluster(num_cores)

# 4. Register the cluster
doParallel::registerDoParallel(cl)

cat("Parallel processing registered with", num_cores, "cores.\n")

t0 <- Sys.time()
data_combinations <- list(
  gc                 = gc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  hplc               = hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  icp                = icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  gc_hplc            = gc_hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  gc_icp             = gc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  hplc_icp           = hplc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  gc_hplc_icp        = gc_hplc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025
)

# Predefine the result list with names from data_combinations
run_rf_analysis_list_SB_and_ENV_train_13Dec25 <- vector("list", length(data_combinations))
names(run_rf_analysis_list_SB_and_ENV_train_13Dec25) <- c(
  "gc_SB_and_ENV_train",
  "hplc_SB_and_ENV_train",
  "icp_SB_and_ENV_train",
  "gc_hplc_SB_and_ENV_train",
  "gc_icp_SB_and_ENV_train",
  "hplc_icp_SB_and_ENV_train",
  "gc_hplc_icp_SB_and_ENV_tra"
  )

# Iterate through each data combination by index
for(i in seq_along(data_combinations)) {
  cat("Processing dataset:", names(data_combinations)[i], "\n")
  
  # Run the analysis using the corresponding data and pass the custom name for figure titles
  run_rf_analysis_list_SB_and_ENV_train_13Dec25[[names(run_rf_analysis_list_SB_and_ENV_train_13Dec25)[i]]] <- run_rf_analysis_manuscript1(
    data = data_combinations[[i]],
    data_name = names(data_combinations)[i], 
    use_source_split = TRUE,
    type_col                = "Plastic_type", 
    remove_cols             = c("Source", "Polymer", "technique"),
    group_for_significance = "Plastic_type",
    do_pairwise_test = FALSE,             # Adjust parameters as needed
    do_top_importance_selection = FALSE,
    do_rfa = FALSE,
    train_proportion = 0.8,
    do_impute_norm_screen = TRUE,
    seed = 123,
    min_sample_number = 1)
  
  # saveRDS(
  #   object = run_rf_analysis_list_SB_and_ENV_train_12Dec25[[names(run_rf_analysis_list_SB_and_ENV_train_12Dec25)[i]]], 
  #   file = paste0(getwd(), "RData_", names(run_rf_analysis_list_SB_and_ENV_train_12Dec25)[[i]], ".rds")
  # )
  
  # Force memory cleanup
  gc()
}
t1 <- Sys.time()
print(paste0("Total run time for all data combos are: ", round(as.numeric(difftime(t1, t0, units = "secs")),2), " seconds"))
stopCluster(cl)
registerDoSEQ() 
```