---
title: "Microplastic fingerprinting"
output: 
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/huyng/OneDrive - Toronto Metropolitan University/Microplastic/Microplastic-Fingerprinting")
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(out.width='750px', dpi=200)
```

## Documentation

This repo is accompanying the publication: "Chemical Fingerprints of New vs. Weathered Microplastics: A Machine Learning Approach"

## Loading Packages and Functions
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# #########################
##   Load Libraries    ##
#########################
library(ggplot2)
library(ranger)
library(foreach)
library(doParallel)
library(parallel)
library(vegan)
library(readxl)
library(tidyverse)
library(dplyr)
library(data.table)
library(writexl)
library(tidyr)
library(grid)
library(gridExtra)
library(stats)
library(FactoMineR)
library(factoextra)
library(ggforce)
library(cluster)
library(MASS)  # For mvrnorm
library(caret)
library(randomForest)
library(pheatmap)
library(viridis)
library(missForest)
library(pROC)
library(VIM)
library(softImpute)
library(purrr)
library(ggpubr)
library(missForest)
library(pROC)
library(VIM)
library(softImpute)
library(purrr)
library(ggpubr)
library(writexl)
library(RANN)   # specific for fast KNN
```

## STEP 1.1: Data import

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# ATDGCMS
setwd("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/ATDGCMS")

atdgcms_list <- list.files(pattern = '*.csv', full.names = TRUE) %>%
  .[!str_detect(., "2022")] %>%
  # remove USE samples with only 1 replication
  .[!str_detect(., "USE_01")] %>%
  .[!str_detect(., "USE_02")] %>%
  .[!str_detect(., "USE_05")] %>%
  .[!str_detect(., "USE_06")] %>%
  .[!str_detect(., "USE_09")] %>%
  .[!str_detect(., "USE_10")] %>%
  .[!str_detect(., "USE_11")] %>%
  .[!str_detect(., "USE_12")] %>%
  .[!str_detect(., "USE_13")] %>%
  .[!str_detect(., "USE_15")] %>%
  .[!str_detect(., "USE_16")] %>%
  .[!str_detect(., "USE_17")] %>%
  .[!str_detect(., "USE_18")] %>%
  .[!str_detect(., "USE_19")] %>%
  .[!str_detect(., "USE_20")] %>%
  .[!str_detect(., "USSB_01")] %>%
  .[!str_detect(., "USSB_08")] # %>%
  # .[!str_detect(., "USSB_09")] # %>%
  # .[str_detect(., "_USSB")] #  environmental samples

# Blank samples 
blank_list <- list.files(pattern = '*.csv', full.names = TRUE) %>%
  .[str_detect(., "2022-")]

# Import samples to list
# Function to read each file and add a "File" column
read_and_add_filename <- function(file) {
  df <- read.csv(file) %>%
    mutate(File = basename(file))  # Add a "File" column with the file name
  return(df)
}

# Apply the function to all files and combine them into a single dataframe
atdgcms_step1.1 <- lapply(atdgcms_list, read_and_add_filename) %>%
  bind_rows() %>%  # Combine all dataframes into one
  # dplyr::bind_rows(purrr::map(atdgcms_list, read.csv)) %>%
  dplyr::select(-c("Start", "End", "Width", "Base.Peak", "Cpd", "Label", "Height", "Ions")) %>%
  # # Change all ATDGCMS file name with underscore-separated to hyphen-separated
  dplyr::mutate(File = gsub("_", "-", File)) %>%
  dplyr::mutate(File = gsub(".csv", "", File)) %>%
  mutate(type = "Sample")

atdgcms_blank <- dplyr::bind_rows(purrr::map(blank_list, read.csv)) %>%
  dplyr::select(c("Area", "File", "m.z", "RT")) %>%
  mutate(type = "Blanks")

atdgcms_step1.1 <- rbind(atdgcms_step1.1, atdgcms_blank) %>% 
  arrange(RT) %>%
  # mutate(Replicate = ifelse(str_detect(File, "R1"), "R1",
  #                           ifelse(str_detect(File, "R2"), "R2", "R3"))) %>%
  mutate(simplified_file = map_chr(File, ~ paste(str_split_1(.x, pattern = "-")[c(4, 5)], collapse = "-")))

# HPLCTOFMS

###### Import Batch 0
setwd("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/HPLCTOFMS/EF_Non-target data_Batch 0")

hplc_batch0 <- list.files(pattern = '*.xls') %>%
  .[!str_detect(., "Blank")]

# Blank samples 
blank_hplc_batch0 <- list.files(pattern = '*.xls') %>%
  .[str_detect(., "Blank")]

# Import samples to list
hplc_batch0_step1.1 <- dplyr::bind_rows(purrr::map(hplc_batch0, read_xls, skip = 1), 
                                        purrr::map(blank_hplc_batch0, read_xls, skip = 1)) %>%
  dplyr::select(m.z = `m/z`, RT, Height, File) %>%
  mutate(File = gsub("_", "-", File),
         type = ifelse(str_detect(File, "Blank"), "Blanks", "Sample"),
         Day = case_when(str_detect(File, "Day0") ~ "0",
                         str_detect(File, "Day2") ~ "2",
                         str_detect(File, "Day15") ~ "15",
                         TRUE ~ "34"),
         Replicate = ifelse(str_detect(File, "R1"), "R1", "R2"),
         Batch_number = "0",
         simplified_file = map_chr(File, ~ paste(str_split_1(.x, "-")[1:2], collapse = "-")))

colnames(hplc_batch0_step1.1)[1] <- "m.z"

###### Import Batch 1
setwd("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/HPLCTOFMS/EF_Non-target data_Batch 1")

# Import sample data
hplc_batch1 <- list.files(pattern = '*.xls') %>%
  .[!str_detect(., "Blank")]
# # Import Blank data 
blank_hplc_batch1 <- list.files(pattern = '*.xls') %>%
  .[str_detect(., "Blank")]

# Put sample and blank data together and create additional description columns
hplc_batch1_step1.1 <- dplyr::bind_rows(purrr::map(hplc_batch1, read_xls, skip = 1), 
                                        purrr::map(blank_hplc_batch1, read_xls, skip = 1)) %>%
  dplyr::select(m.z = `m/z`, RT, Height, File) %>%
  mutate(File = gsub("_", "-", File),
         type = ifelse(str_detect(File, "Blank"), "Blanks", "Sample"),
         Day = case_when(str_detect(File, "Day0") ~ "0",
                         str_detect(File, "Day2") ~ "2",
                         str_detect(File, "Day15") ~ "15",
                         TRUE ~ "34"),
         Replicate = ifelse(str_detect(File, "R1"), "R1", "R2"),
         Batch_number = "1",
         simplified_file = map_chr(File, ~ paste(str_split_1(.x, "-")[2:3], collapse = "-")))

# Change column name
colnames(hplc_batch1_step1.1)[1] <- "m.z"

###### Import Batch 2
setwd("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/HPLCTOFMS/EF_Non-target data_Batch 2")

hplc_batch2 <- list.files(pattern = '*.xls') %>%
  .[!str_detect(., "Blank")]
# Blank samples 
blank_hplc_batch2 <- list.files(pattern = '*.xls') %>%
  .[str_detect(., "Blank")]

hplc_batch2_step1.1 <- dplyr::bind_rows(purrr::map(hplc_batch2, read_xls, skip = 1), 
                                        purrr::map(blank_hplc_batch2, read_xls, skip = 1)) %>%
  dplyr::select(m.z = `m/z`, RT, Height, File) %>%
  mutate(File = gsub("_", "-", File),
         type = ifelse(str_detect(File, "Blank"), "Blanks", "Sample"),
         Day = case_when(str_detect(File, "Day0") ~ "0",
                         str_detect(File, "Day2") ~ "2",
                         str_detect(File, "Day15") ~ "15",
                         TRUE ~ "34"),
         Replicate = ifelse(str_detect(File, "R1"), "R1", "R2"),
         Batch_number = "2",
         simplified_file = map_chr(File, ~ paste(str_split_1(.x, "-")[2:3], collapse = "-")))

colnames(hplc_batch2_step1.1)[1] <- "m.z"

all_hplc <- rbind(hplc_batch0_step1.1,
                  hplc_batch1_step1.1,
                  hplc_batch2_step1.1) %>% arrange(RT)

## Stable Isotope data
stable_isotope <- read_xlsx(path = paste0(getwd(), "/data/Isotope/Mikroplastik_Results_2ndRound_KCN-1.xlsx")) %>% 
  tidyr::pivot_longer(cols = 4:5, names_to = "replicate", values_to = "Value") %>% 
  dplyr::mutate(Value = abs(Value))
```

## STEP 1.2: Collapsing compounds based on RT1, RT2, Ion1 threshold --------------------------

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# ATDGCMS
atdgcms_grouped <- grouping_comp(atdgcms_step1.1,
                                 rtthres1 = 0.005,
                                 mzthres = 0.05,
                                 type = "ATDGCMS")

# HPLCTOFMS
# Apply different alignment to 2 parts of chromatogram via rtthres1 (RT <= 6 min) and rtthres2 (RT > 6 min)
hplc_batch_grouped <- grouping_comp(all_hplc,
                                    split_time = 6,
                                    rtthres1 = 0.4,
                                    rtthres2 = 0.28,
                                    mzthres = 0.0005,
                                    type = "HPLCTOFMS")
```

## Step 1.3: Remove all the benchmark compounds for HPLC (4) and GC (1 - Toluene)
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
#ATDGCMS
idx <- which(atdgcms_grouped$m.z >= 98 & 
               atdgcms_grouped$m.z <= 98.5 & 
               atdgcms_grouped$RT > 13.4 & 
               atdgcms_grouped$RT < 13.5)
atdgcms_grouped <- atdgcms_grouped[-idx,]

# HPLC
hplc_benchmark_mz <- c(156.0957, 198.1723, 246.2568, 305.1956, 342.1722)
# Search for compound in combined_df2_grouped that fall in the window of benchmark' molecular ions
# Batch 0
hplc_benchmark_idx <- c()
for (mz in hplc_benchmark_mz) {
  idx <- which(hplc_batch_grouped$m.z <= (mz + 0.0005) & hplc_batch_grouped$m.z >= (mz - 0.0005))
  # collect all these indices of benchmark compounds
  hplc_benchmark_idx <- c(hplc_benchmark_idx, idx)
}
hplc_batch_grouped <- hplc_batch_grouped[-hplc_benchmark_idx,]
```


## Step 2: Blank subtraction & Remove all detections with negative peak area

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# ATDGCMS --------------
# Create list to store temp dfs
process_atdgcms <- function(atdgcms_grouped) {
  result_list <- list()
  idx <- 1
  
  # Iterate through each unique feature
  for (feature in unique(atdgcms_grouped$Feature)) {
    temp <- atdgcms_grouped[atdgcms_grouped$Feature == feature, ]
    
    # If no blanks exist for the current feature, add it as is
    if (identical(which(temp$type == "Blanks"), integer(0))) {
      result_list[[idx]] <- temp
      idx <- idx + 1
      next
    } else {
      # Calculate the mean and sd for blanks
      blank_idx <- which(temp$type == "Blanks")
      blank_mean <- base::mean(temp[blank_idx,]$Area, na.rm = TRUE)
      blank_sd   <- stats::sd(temp[blank_idx,]$Area, na.rm = TRUE)
      
      # Exclude blanks from further processing
      temp <- temp[temp$type != "Blanks", ]
      
      # Adjust the 'Area' for each sample within the feature
      for (sample in unique(temp$File)) {
        sample_idx <- which(temp$File == sample)
        temp[sample_idx,]$Area <- temp[sample_idx,]$Area - blank_mean
        
        # Keep only those entries with adjusted Area > 0 and > 3 * blank_sd
        positive_idx <- which(temp$Area > 0 & temp$Area > 3 * blank_sd)
        temp <- temp[positive_idx, ]
      }
    }
    
    # Append processed data frame for this feature
    result_list[[idx]] <- temp
    idx <- idx + 1
  }
   
  output <- dplyr::bind_rows(result_list)
  return(output)
}

# For positive values => combine data again to 1 grand data frame
atdgcms_adjusted_step3 <- process_atdgcms(atdgcms_grouped)

# HPLCTOFMS  --------------
hplc_result_list <- list()
idx <- 1

# Iterate through each batch
for (batch in unique(hplc_batch_grouped$Batch_number)) {
  # Iterate through each unique feature within each batch
  for (feature in unique(hplc_batch_grouped$Feature)) {
    temp <- hplc_batch_grouped[which(hplc_batch_grouped$Feature == feature & hplc_batch_grouped$Batch_number == batch), ]
    
    # If no blanks exist for the current feature, add it as is
    if (identical(which(temp$type == "Blanks"), integer(0))) {
      hplc_result_list[[idx]] <- temp
      idx <- idx + 1
      next
    } else { # if blanks exist for the current feature in the current batch number
      # Calculate the mean and sd for blanks
      blank_idx <- which(temp$type == "Blanks")
      blank_mean <- base::mean(temp[blank_idx,]$Height, na.rm = TRUE)
      blank_sd   <- stats::sd(temp[blank_idx,]$Height, na.rm = TRUE)
      
      # Exclude blanks from further processing
      temp <- temp[temp$type != "Blanks", ]
      
      # Adjust the 'Height' for each sample within the feature
      for (sample in unique(temp$File)) {
        sample_idx <- which(temp$File == sample)
        temp[sample_idx,]$Height <- temp[sample_idx,]$Height - blank_mean
        
        # Keep only those entries with adjusted Height > 0 and > 3 * blank_sd
        positive_idx <- which(temp$Height > 0 & temp$Height > (3 * blank_sd))
        temp <- temp[positive_idx, ]
      }
    }
    
    # Append processed data frame for this feature
    hplc_result_list[[idx]] <- temp
    idx <- idx + 1
  }
}

hplc_adjusted_step3 <- dplyr::bind_rows(hplc_result_list)
```

# Add info on Source and rename columns
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# ATDGCMS
# idx_list_filter1 <- comp_filter_opt(atdgcms_adjusted_step3)
gc <- atdgcms_adjusted_step3 %>%
  dplyr::filter(!is.na(Feature)) %>%
  dplyr::select(File, Feature, m.z, RT, Area) %>%
  mutate(Source = ifelse(str_detect(File, "USE"), "Environmental", "Store-Bought")) %>%
  dplyr::rename(Values = Area)

gc$technique <- "GC"

# HPLCTOFMS
# idx_list_filter2 <- comp_filter_opt(hplc_adjusted_step3)
hplc <- hplc_adjusted_step3 %>% # [idx_list_filter2[[1]],] %>%
  # dplyr::filter(Day %notin% c("0", "2")) %>%
  dplyr::filter(!is.na(Feature)) %>%
  dplyr::select(File, Feature, 
                # Day, 
                m.z, RT, Height) %>%
  dplyr::mutate(Source = ifelse(str_detect(File, "USE"), "Environmental", "Store-Bought")) %>%
  dplyr::rename(Values = Height)

hplc$technique <- "HPLC"
```

## STEP 4: Trace metal data

#### Data prepping
- Problem with icpms_round1 versus. icpms_round2: round1 has Li [No gas] that round2 does not.
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# Trace metal data
icpms_round1 <- readxl::read_excel("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/ICPMS_Trace metal/icpms_round1_rawdata_removal USE-01-rep1 and USE-03 (only2observations).xlsx") %>% 
  column_to_rownames(., var = "File")

icpms_round2_batch1 <- readxl::read_excel("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/ICPMS_Trace metal/icpms_round2_batch1_rawdata.xlsx") %>% 
  column_to_rownames(., var = "File")

icpms_round2_batch2 <- readxl::read_excel("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/ICPMS_Trace metal/icpms_round2_batch2_rawdata.xlsx") %>% 
  column_to_rownames(., var = "File")

# 1) Select [H2] instead of [H2 HMI]----------------------
library(stringr)

remove_H2_HMI_modus <- function(icpms_df) {
  strings <- colnames(icpms_df)
  pattern <- "^(.*\\[ )([A-Za-z0-9 ]+)( \\])$"
  metal_namedata <- data.frame(
    original = strings,
    prefix = str_extract(strings, "^.*\\["),
    content = str_match(strings, pattern)[,3],
    stringsAsFactors = FALSE
  )
  
  metal_namedata <- metal_namedata %>% dplyr::filter(!str_detect(content, pattern = "H2 HMI"))
  df <- icpms_df %>% dplyr::select(metal_namedata$original)
  return(df)
}

icpms_round1 <- remove_H2_HMI_modus(icpms_round1)
icpms_round2_batch1 <- remove_H2_HMI_modus(icpms_round2_batch1)
icpms_round2_batch2 <- remove_H2_HMI_modus(icpms_round2_batch2)

# 2) Between the ones that use [N2O] as analyzing modus -> figure out within the same metal element, which one has the highest "Recovery [%]" or if Recovery [%] is missing then lowest "U(k=2)" ---------------------------------------

recovery_uncertainty_round1 <- readxl::read_excel("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/ICPMS_Trace metal/Trace metal data_Recovery_Uncertainty_round1.xlsx") %>%
  filter(`Element+Modus` %in% colnames(icpms_round1))

recovery_uncertainty_round2_batch1 <- readxl::read_excel("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/ICPMS_Trace metal/Trace metal data_Recovery_Uncertainty_round2_batch1.xlsx") %>%
  filter(`Element+Modus` %in% colnames(icpms_round2_batch1))

recovery_uncertainty_round2_batch2 <- readxl::read_excel("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/ICPMS_Trace metal/Trace metal data_Recovery_Uncertainty_round2_batch2.xlsx") %>%
  filter(`Element+Modus` %in% colnames(icpms_round2_batch2))

get_final_metal_names <- function(recovery_uncertainty) {
  
  # 1) Extract the first number
  first_numbers <- sub(" .*", "", recovery_uncertainty$`Element+Modus`)
  
  # 2) Define a function to find duplicate indices
  duplicate_indices <- function(vector) {
    unique_elements <- unique(vector)
    result <- list()
    
    for (element in unique_elements) {
      indices <- which(vector == element)
      if (length(indices) > 1) {
        result[[element]] <- indices
      }
    }
    
    return(result)
  }
  
  # Apply duplicate_indices to first_numbers
  res <- duplicate_indices(first_numbers)
  
  # 3) For duplicates, pick either the row with the highest "Recovery [%]"
  #    or if that is empty, the row with the lowest "U(k=2)"
  recovery_uncertainty_checked <- c()
  for (ele in seq_along(res)) {
    temp <- recovery_uncertainty[res[[ele]], ]
    idx1 <- temp[which.max(temp$`Recovery [%]`), ]
    
    # if no maximum Recovery row is found, pick the minimum U(k=2) row instead
    if (identical(idx1, character(0))) {
      chosen <- temp[which.min(temp$`U(k=2)`), ]$`Element+Modus`
    } else {
      chosen <- temp[which.max(temp$`Recovery [%]`), ]$`Element+Modus`
    }
    recovery_uncertainty_checked <- c(recovery_uncertainty_checked, chosen)
  }
  
  # 4) Combine the unique rows (non-duplicates) plus the chosen duplicates
  final_metal_names <- c(
    recovery_uncertainty[-unlist(res, use.names = FALSE), ]$`Element+Modus`,
    recovery_uncertainty_checked
  )
  
  return(final_metal_names)
}

final_metal_names_round1 <- get_final_metal_names(recovery_uncertainty_round1)
final_metal_names_round2_batch1 <- get_final_metal_names(recovery_uncertainty_round2_batch1)
final_metal_names_round2_batch2 <- get_final_metal_names(recovery_uncertainty_round2_batch2)

# subset dataframe for further analyses
final_icpms_round1 <- icpms_round1 %>% 
  dplyr::select(all_of(final_metal_names_round1))

final_icpms_round2_batch1 <- icpms_round2_batch1 %>% 
  dplyr::select(all_of(final_metal_names_round2_batch1))

final_icpms_round2_batch2 <- icpms_round2_batch2 %>% 
  dplyr::select(all_of(final_metal_names_round2_batch2))

common_cols <- Reduce(intersect, list(names(final_icpms_round1), 
                                      names(final_icpms_round2_batch1), 
                                      names(final_icpms_round2_batch2)))
combined_icpms_round1_round2 <- rbind(final_icpms_round1[common_cols], 
                                      final_icpms_round2_batch1[common_cols],
                                      final_icpms_round2_batch2[common_cols])
View(combined_icpms_round1_round2)

# 3) Final cleaning of name of metal elements (OPTIONAL) ----------------------------------
# Function to process the strings
process_string <- function(name_vector) {
  # Remove everything after the square bracket
  cleaned_vector <- str_extract(name_vector, "^[^\\[]+")
  # Trim any trailing whitespace
  cleaned_vector <- str_trim(cleaned_vector)
  # Use a regular expression to remove the numerical parts
  # Explanation of Regular Expression:
  # 
  # \\d+ matches one or more digits.
  # *(-> *\\d+)? * optionally matches the pattern -> digits with any number of spaces around it.
  # gsub function is used to find and replace the matching parts in the string with an empty string ("")
  cleaned_vector <- gsub("\\d+ *(-> *\\d+)? *", "", cleaned_vector)
  return(cleaned_vector)
}

# Apply the function to each element of the vector
# colnames(combined_icpms_round1_round2) <- sapply(colnames(combined_icpms_round1_round2), process_string)

# 4) Discard all metal that may appear in seawater - Sodium (Na), Magnesium (Mg), Potassium (K), Calcium (Ca), Sr (optional) --------------------------
strings_to_remove <- c("Na", "Mg", "K", "Ca") #, "Sr")

cols_to_remove <- unique(unlist(sapply(strings_to_remove, function(x) grep(x, colnames(combined_icpms_round1_round2)))))
# remove the seawater metal from combined rounds ICP data
combined_icpms_round1_round2 <- combined_icpms_round1_round2[, -cols_to_remove]

icp <- combined_icpms_round1_round2 %>%
  rownames_to_column(., var = "File")  %>% 
  pivot_longer(., cols = 2:ncol(.), names_to = "Feature", values_to = "Values")

icp$technique <- "ICP"
icp <- icp %>%
  mutate(Source = ifelse(str_detect(File, "USE"), "Environmental", "Store-Bought")) %>%
  mutate(Values = as.numeric(Values))
```

## Combine Label Info with raw data
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
p <- "C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/data/Sample Labelling_all data_GC+HPLC+ICP_30Sept2025.xlsx"

sampinfo <- readxl::read_excel(path = p)

# joining data frame with sampinfo df
icp <- left_join(x = icp ,
                 y = sampinfo %>% dplyr::select(c(File, Plastic_type, Subcategory, Polymer)), 
                 by = 'File') # %>% dplyr::filter(!is.na(Similar_type))

gc <- left_join(x = gc ,
                y = sampinfo %>% dplyr::select(c(File, Plastic_type, Subcategory, Polymer)), 
                by = 'File') # %>% dplyr::filter(!is.na(Similar_type))

hplc <- left_join(x = hplc ,
                  y = sampinfo %>% dplyr::select(c(File, Plastic_type, Subcategory, Polymer)), 
                  by = 'File') # %>% dplyr::filter(!is.na(Similar_type))

## Aggregate Day 15 and 34 by mean
# new_hplc <- hplc %>%
#   group_by(Feature, Subcategory, Polymer, technique) %>%
#   summarise(Mean_Values = mean(Values, na.rm = TRUE), .groups = "drop")

gc_hplc <- bind_rows(gc, hplc)

gc_icp <- bind_rows(icp, 
                    gc %>% dplyr::select(-c("m.z", "RT"))) # %>% dplyr::filter(!is.na(Similar_type))

hplc_icp <- bind_rows(icp, 
                      hplc %>% dplyr::select(-c("m.z", "RT"))) # %>% dplyr::filter(!is.na(Similar_type))

gc_hplc_icp <- bind_rows(icp, 
                         gc_hplc %>% dplyr::select(-c("m.z", "RT"))) # %>% dplyr::filter(!is.na(Similar_type))

```


## Data cleaning
### Summary of data before feature reduction
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
df_in <- gc_hplc_icp
# 1. Process Data ONCE (Consolidated Pivot)
# We calculate the mean for duplicates and pivot to wide format immediately
df_wide <- df_in %>% 
  dplyr::select(File, Plastic_type, Source, Polymer, Feature, technique, Values) %>%
  dplyr::group_by(File, Source, Polymer, Plastic_type, technique, Feature) %>%
  dplyr::summarise(Values = mean(Values, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = Feature, values_from = Values) %>%
  tibble::column_to_rownames(var = "File") 

# 2. Separate Numeric Data from Metadata
# Identify metadata columns (non-numeric) to exclude them from calculations
df_numeric <- df_wide %>% dplyr::select(where(is.numeric))

# --- PART 1 & 2: Basic Statistics ---

cat("Number of chemical features: ", ncol(df_numeric), "\n")

# Note: Using df_numeric for dimensions to avoid diluting the % with metadata columns
missing_pct <- sum(is.na(df_numeric)) / prod(dim(df_numeric)) * 100
cat("Percentage of missing values (%):", round(missing_pct, 2), "\n")

# --- PART 3: Correlations ---

# Step 1: Calculate NA counts and Mean Response directly from the numeric matrix
stats_df <- tibble::tibble(
  Feature = colnames(df_numeric),
  num_missing_values = colSums(is.na(df_numeric)),
  mean_response = colMeans(df_numeric, na.rm = TRUE)
)

# Step 2: Check for Metadata Columns (The Fix)
# We only calculate m/z and RT correlations if the columns actually exist
if (all(c("m.z", "RT") %in% colnames(df_in))) {
  
  feature_meta <- df_in %>%
    dplyr::group_by(Feature) %>%
    dplyr::summarise(
      m_z = mean(m.z, na.rm = TRUE),
      retention_time = mean(RT, na.rm = TRUE),
      .groups = "drop"
    )
  
  # Join and calculate all correlations
  analysis_df <- stats_df %>% dplyr::left_join(feature_meta, by = "Feature")
  
  cat("Pearson (Mean Response vs Missing):", round(cor(analysis_df$mean_response, analysis_df$num_missing_values, use = "pairwise"), 2), "\n")
  cat("Pearson (m/z vs Missing):          ", round(cor(analysis_df$m_z, analysis_df$num_missing_values, use = "pairwise"), 2), "\n")
  cat("Pearson (RT vs Missing):           ", round(cor(analysis_df$retention_time, analysis_df$num_missing_values, use = "pairwise"), 2), "\n")
  
} else {
  
  # Fallback if m.z/RT are missing
  cor_response <- cor(stats_df$mean_response, stats_df$num_missing_values, use = "pairwise.complete.obs")
  
  cat("Pearson (Mean Response vs Missing):", round(cor_response, 2), "\n")
  message("\nNOTE: 'm.z' and 'RT' columns were not found in the input file. Those correlations were skipped.")
}
```

### Feature reduction 1: Regular 80% rule - Remove feature if there are more than 80% MVs

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# Replace 'gc_hplc_icp' with your actual dataframe name
results <- shared_feature_filtering(gc, 
                                    feat_reduc_mode = "regular80")

# B. Extract Data for Calculation
df_final <- results$filtered_data
df_orig  <- results$original_data

# Define metadata columns to exclude from stats
meta_cols <- c("Plastic_type", "technique", "Source", "Polymer")

# Get list of feature names only
feats_final <- setdiff(names(df_final), meta_cols)
feats_orig  <- setdiff(names(df_orig), meta_cols)

# --- 1. Percentage of chemical features retained (%) ---
pct_features <- (length(feats_final) / length(feats_orig)) * 100
cat(paste0("1. Percentage of chemical features retained: ", round(pct_features, 2), "%\n"))

# --- 2. Percentage of missing values after feature reduction (%) ---
# We calculate this ONLY on the feature columns of the filtered dataset
if(length(results$kept_features) > 0) {
  vals_final <- df_final[, feats_final]
  pct_missing <- (sum(is.na(vals_final)) / prod(dim(vals_final))) * 100
  cat(paste0("2. Percentage of missing values after feature reduction: ", round(pct_missing, 2), "%\n"))
} else {
  vals_final <- df_final[, feats_final]
  cat(paste0("2. Percentage of missing values after feature reduction: ", 0, "%\n"))
}


# --- 3. Percentage of signal response after feature reduction (%) ---
# Sum of intensities in filtered data / Sum of intensities in original (zero-filled) data
signal_final <- sum(vals_final, na.rm = TRUE)
signal_orig  <- sum(df_orig[, feats_orig], na.rm = TRUE)
pct_signal   <- (signal_final / signal_orig) * 100

cat(paste0("3. Percentage of signal response after feature reduction: ", round(pct_signal, 2), "%\n"))
```

### Feature reduction 2: Modified 80% rule
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
results <- shared_feature_filtering(gc, 
                                    feat_reduc_mode = "modified80")

# B. Extract Data for Calculation
df_final <- results$filtered_data
df_orig  <- results$original_data

# Define metadata columns to exclude from stats
meta_cols <- c("Plastic_type", "technique", "Source", "Polymer")

# Get list of feature names only
feats_final <- setdiff(names(df_final), meta_cols)
feats_orig  <- setdiff(names(df_orig), meta_cols)

# --- 1. Percentage of chemical features retained (%) ---
pct_features <- (length(feats_final) / length(feats_orig)) * 100
cat(paste0("1. Percentage of chemical features retained: ", round(pct_features, 2), "%\n"))

# --- 2. Percentage of missing values after feature reduction (%) ---
# We calculate this ONLY on the feature columns of the filtered dataset
if(length(results$kept_features) > 0) {
  vals_final <- df_final[, feats_final]
  pct_missing <- (sum(is.na(vals_final)) / prod(dim(vals_final))) * 100
  cat(paste0("2. Percentage of missing values after feature reduction: ", round(pct_missing, 2), "%\n"))
} else {
  vals_final <- df_final[, feats_final]
  cat(paste0("2. Percentage of missing values after feature reduction: ", 0, "%\n"))
}

# --- 3. Percentage of signal response after feature reduction (%) ---
# Sum of intensities in filtered data / Sum of intensities in original (zero-filled) data
signal_final <- sum(vals_final, na.rm = TRUE)
signal_orig  <- sum(df_orig[, feats_orig], na.rm = TRUE)
pct_signal   <- (signal_final / signal_orig) * 100

cat(paste0("3. Percentage of signal response after feature reduction: ", round(pct_signal, 2), "%\n"))
```


### Feature reduction 3: In-house Comprehensive feature filtering 
```{r , echo=FALSE, warning = FALSE, message=FALSE}
results <- shared_feature_filtering(hplc, 
                                    feat_reduc_mode = "inhouse")

# B. Extract Data for Calculation
df_final <- results$filtered_data
df_orig  <- results$original_data

# Define metadata columns to exclude from stats
meta_cols <- c("Plastic_type", "technique", "Source", "Polymer")

# Get list of feature names only
feats_final <- setdiff(names(df_final), meta_cols)
feats_orig  <- setdiff(names(df_orig), meta_cols)

# --- 1. Percentage of chemical features retained (%) ---
pct_features <- (length(feats_final) / length(feats_orig)) * 100
cat(paste0("1. Percentage of chemical features retained: ", round(pct_features, 2), "%\n"))

# --- 2. Percentage of missing values after feature reduction (%) ---
# We calculate this ONLY on the feature columns of the filtered dataset
if(length(results$kept_features) > 0) {
  vals_final <- df_final[, feats_final]
  pct_missing <- (sum(is.na(vals_final)) / prod(dim(vals_final))) * 100
  cat(paste0("2. Percentage of missing values after feature reduction: ", round(pct_missing, 2), "%\n"))
} else {
  vals_final <- df_final[, feats_final]
  cat(paste0("2. Percentage of missing values after feature reduction: ", 0, "%\n"))
}

# --- 3. Percentage of signal response after feature reduction (%) ---
# Sum of intensities in filtered data / Sum of intensities in original (zero-filled) data
signal_final <- sum(vals_final, na.rm = TRUE)
signal_orig  <- sum(df_orig[, feats_orig], na.rm = TRUE)
pct_signal   <- (signal_final / signal_orig) * 100

cat(paste0("3. Percentage of signal response after feature reduction: ", round(pct_signal, 2), "%\n"))
```

### Data fusion
```{r, echo = FALSE, message = FALSE, warning = FALSE}
# do data cleaning - change variable name accordingly to data combinations
gc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025 <- shared_feature_filtering(gc, 
                                                                               feat_reduc_mode = "modified80")$filtered_data
# hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025 <- shared_feature_filtering(hplc, 
#                                                                                  feat_reduc_mode = "inhouse")$filtered_data
icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025 <- shared_feature_filtering(icp, 
                                                                                feat_reduc_mode = "modified80")$filtered_data

# Create data fusion combos
gc_hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025 <- data_fusion(gc_hplc,
                                                                       source_feats = list(gc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
                                                                                           hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025))


gc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025 <- data_fusion(gc_icp,
                                                                      source_feats = list(gc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
                                                                                          icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025))
hplc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025 <- data_fusion(hplc_icp,
                                                                        source_feats = list(hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
                                                                                            icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025))
gc_hplc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025 <- data_fusion(gc_hplc_icp,
                                                                           source_feats = list(gc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
                                                                                               hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
                                                                                               icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025))

# Extract original data with process_and_filter function for quality assurance report after feature reduction ---------
results <- process_and_filter(gc_icp, 
                              class_col = "Plastic_type", 
                              threshold = 0.8, 
                              filter_mode = "inhouse")

# B. Extract Data for Calculation
df_final <- gc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025
df_orig  <- results$original_data

# Define metadata columns to exclude from stats
meta_cols <- c("Plastic_type", "technique", "Source", "Polymer")

# Get list of feature names only
feats_final <- setdiff(names(df_final), meta_cols)
feats_orig  <- setdiff(names(df_orig), meta_cols)

# --- 1. Percentage of chemical features retained (%) ---
pct_features <- (length(feats_final) / length(feats_orig)) * 100
cat(paste0("1. Percentage of chemical features retained: ", round(pct_features, 2), "%\n"))

# --- 2. Percentage of missing values after feature reduction (%) ---
# We calculate this ONLY on the feature columns of the filtered dataset
if(!identical(setdiff(names(gc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025), meta_cols), character(0))) {
  vals_final <- df_final[, feats_final]
  pct_missing <- (sum(is.na(vals_final)) / prod(dim(vals_final))) * 100
  cat(paste0("2. Percentage of missing values after feature reduction: ", round(pct_missing, 2), "%\n"))
} else {
  vals_final <- df_final[, feats_final]
  cat(paste0("2. Percentage of missing values after feature reduction: ", 0, "%\n"))
}

# --- 3. Percentage of signal response after feature reduction (%) ---
# Sum of intensities in filtered data / Sum of intensities in original (zero-filled) data
signal_final <- sum(vals_final, na.rm = TRUE)
signal_orig  <- sum(df_orig[, feats_orig], na.rm = TRUE)
pct_signal   <- (signal_final / signal_orig) * 100

cat(paste0("3. Percentage of signal response after feature reduction: ", round(pct_signal, 2), "%\n"))
```

## Figure 2: HCA
```{r echo=FALSE, warning=FALSE, message=FALSE}
# Data prep for HCA ---------------------
df <- run_rf_analysis_list_SB_and_ENV_train_02Aug2025$hplc_SB_and_ENV_train$final_imp_norm_dat %>%
  dplyr::mutate(Plastic_type = gsub("..", " ", Plastic_type, fixed = TRUE)) %>%
  dplyr::mutate(Plastic_type = gsub(".", " ", Plastic_type, fixed = TRUE)) %>%
  dplyr::mutate(Plastic_type_Source = paste0(Plastic_type, "-", Source))

hc_df <- df %>% 
  dplyr::select(-c("Subcategory", "Source", "Plastic_type", "Polymer", "technique")) %>% #, "Plastic_type_Source")) %>%
  dplyr::select(where(is.numeric)) %>%
  as.data.frame()

# --- Clean: remove bad rows & handle non-finite values ---
# 1) Replace non-finite with 0 (or choose a more domain-specific imputation if appropriate)
hc_df[!is.finite(as.matrix(hc_df))] <- 0

# 2) Drop rows that are all zero (Bray–Curtis can yield NaN for empty rows)
keep <- rowSums(hc_df, na.rm = TRUE) > 0
hc_df_clean <- hc_df[keep, , drop = FALSE] #%>% 
  # Select the features from best performance feature selection method
  # dplyr::select(run_rf_analysis_list_SB_and_ENV_train_02Aug2025$icp_SB_and_ENV_train$rfa_selected_feats)
labels_clean <- df$Plastic_type_Source[keep] # df$Plastic_type_Source[keep] 

# 3) If any negatives remain, Bray–Curtis is invalid -> fall back to a metric that permits negatives
use_bray <- !any(hc_df_clean < 0, na.rm = TRUE)

if (use_bray) {
  d <- vegan::vegdist(hc_df_clean, method = "manhattan")
  link <- "average"
} else {
  # If your features were centered/scaled, Euclidean or Manhattan are appropriate.
  # (Alternatively, if these are compositions, consider Aitchison with a small pseudocount.)
  d <- dist(hc_df_clean, method = "robust.aitchison")
  link <- "average"
  message("Detected negative values; using Euclidean instead of Bray–Curtis.")
}

# --- Guard: ensure distance has no NA/NaN/Inf ---
if (any(!is.finite(d))) {
  stop("Distance matrix still has non-finite entries after cleaning. Re-check inputs.")
}

# --- Cluster & plot ---
hca_samp <- hclust(d, method = link)

cat("cophenetic correlation coefficient (CCC) from a hierarchical clustering (HCA) is:", cor(d, stats::cophenetic(hca_samp)), "\n")

# tiff(filename = paste0("Figure SI_HCA_HPLC_SB_and_ENV_train_", format(Sys.time(), "%d-%m-%Y"),".tif"), width = 42*600, height = 15*600, res = 600, compression = "lzw")

par(cex = 1)
plot(hca_samp,
     labels = labels_clean,
     hang   = -1,
     main   = "")
# dev.off()
```

# Random Forest
### Option 1: Use SB to predict ENV - Only select shared features between SB and ENV 
```{r class.source = 'fold-hide', echo=FALSE, warning = FALSE, message=FALSE}
###################################################################################
## Explanation of the parameter of run_rf_analysis()
# split environmental into 50:50 and combine with store_bought -> use_source_split = FALSE
# use store-bought as training and environmental as testing -> use_store_vs_environmental_split = TRUE
###################################################################################
# 1. Detect available cores
# Use detectCores() - 1 to leave one core free for your OS so your computer doesn't freeze.
num_cores <- parallel::detectCores(logical = FALSE) - 1  

# 2. SAFETY CHECK (Crucial for your specific error)
# Your previous error was likely due to running out of RAM. 
# Start with a SMALL number (e.g., 2 or 4) to test stability.
# If you have 32GB+ RAM, you can try higher, but for now, limit it:
# if(num_cores > 4) num_cores <- 4 

# 3. Create the cluster
cl <- parallel::makePSOCKcluster(num_cores)

# 4. Register the cluster
doParallel::registerDoParallel(cl)
cat("Parallel processing registered with", num_cores, "cores.\n")

t0 <- Sys.time()
data_combinations <- list(
  gc                 = gc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  # hplc               = hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  # icp                = icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  # gc_hplc            = gc_hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  gc_icp             = gc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025
  # hplc_icp           = hplc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  # gc_hplc_icp        = gc_hplc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025
)

# Predefine the result list with names from data_combinations
run_rf_analysis_list_SB_train_test_modified80_gcandgcicp <- vector("list", length(data_combinations))
names(run_rf_analysis_list_SB_train_test_modified80_gcandgcicp) <- c(
  "gc_USSB_train",
  # "hplc_USSB_train",
  # "icp_USSB_train",
  # "gc_hplc_USSB_train",
  "gc_icp_USSB_train"
  # "hplc_icp_USSB_train",
  # "gc_hplc_icp_USSB_train"
  )

# Iterate through each data combination by index
for(i in seq_along(data_combinations)) {
  cat("Processing dataset:", names(data_combinations)[i], "\n")
  
  # Run the analysis using the corresponding data and pass the custom name for figure titles
  run_rf_analysis_list_SB_train_test_modified80_gcandgcicp[[names(run_rf_analysis_list_SB_train_test_modified80_gcandgcicp)[i]]] <- run_rf_analysis_manuscript1(
    data = data_combinations[[i]],
    type_col                = "Plastic_type", 
    remove_cols             = c("Source", "Polymer", "technique"),
    train_proportion = 0.8,
    use_store_vs_environmental_split = TRUE,
    data_name = names(data_combinations)[i], 
    group_for_significance = "Plastic_type", 
    do_pairwise_test = TRUE,             # Adjust parameters as needed
    do_top_importance_selection = TRUE,
    do_rfa = TRUE,
    do_impute_norm_screen = TRUE,
    seed = 123,
    min_sample_number = 1)
  
  # saveRDS(
  #   object = run_rf_analysis_list_SB_train_12Dec25[[names(run_rf_analysis_list_SB_train_12Dec25)[i]]], 
  #   file = paste0(getwd(), "RData_", names(run_rf_analysis_list_SB_train_12Dec25)[[i]], ".rds")
  # )
  
  # Force memory cleanup
  gc()
}

t1 <- Sys.time()
print(paste0("Total run time for all data combos are: ", round(as.numeric(difftime(t1, t0, units = "secs")),2), " seconds"))
stopCluster(cl)
registerDoSEQ() # Good practice to reset to sequential after you are done
```

### Option 2: Use Both SB and ENV to predict ENV

```{r class.source = 'fold-hide', echo=FALSE, warning = FALSE, message=FALSE}
# 1. Detect available cores
# Use detectCores() - 1 to leave one core free for your OS so your computer doesn't freeze.
num_cores <- parallel::detectCores(logical = FALSE) - 1  

# 2. SAFETY CHECK (Crucial for your specific error)
# Your previous error was likely due to running out of RAM. 
# Start with a SMALL number (e.g., 2 or 4) to test stability.
# If you have 32GB+ RAM, you can try higher, but for now, limit it:
# if(num_cores > 4) num_cores <- 4 

# 3. Create the cluster
cl <- parallel::makePSOCKcluster(num_cores)

# 4. Register the cluster
doParallel::registerDoParallel(cl)

cat("Parallel processing registered with", num_cores, "cores.\n")

t0 <- Sys.time()
data_combinations <- list(
  gc                 = gc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  # hplc               = hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  # icp                = icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  # gc_hplc            = gc_hplc_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  gc_icp             = gc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025
  # hplc_icp           = hplc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025,
  # gc_hplc_icp        = gc_hplc_icp_combined_SB_ENV_shared_cols_ntr_clean_13Dec2025
)

# Predefine the result list with names from data_combinations
run_rf_analysis_list_SB_and_ENV_train_test_modified80_gcandgcicp <- vector("list", length(data_combinations))
names(run_rf_analysis_list_SB_and_ENV_train_test_modified80_gcandgcicp) <- c(
  "gc_SB_and_ENV_train",
  # "hplc_SB_and_ENV_train",
  # "icp_SB_and_ENV_train",
  # "gc_hplc_SB_and_ENV_train",
  "gc_icp_SB_and_ENV_train"
  # "hplc_icp_SB_and_ENV_train",
  # "gc_hplc_icp_SB_and_ENV_tra"
  )

# Iterate through each data combination by index
for(i in seq_along(data_combinations)) {
  cat("Processing dataset:", names(data_combinations)[i], "\n")
  
  # Run the analysis using the corresponding data and pass the custom name for figure titles
  run_rf_analysis_list_SB_and_ENV_train_test_modified80_gcandgcicp[[names(run_rf_analysis_list_SB_and_ENV_train_test_modified80_gcandgcicp)[i]]] <- run_rf_analysis_manuscript1(
    data = data_combinations[[i]],
    data_name = names(data_combinations)[i], 
    use_source_split = TRUE,
    type_col                = "Plastic_type", 
    remove_cols             = c("Source", "Polymer", "technique"),
    group_for_significance = "Plastic_type",
    do_pairwise_test = TRUE,             # Adjust parameters as needed
    do_top_importance_selection = TRUE,
    do_rfa = TRUE,
    train_proportion = 0.8,
    do_impute_norm_screen = TRUE,
    seed = 123,
    min_sample_number = 1)
  
  # saveRDS(
  #   object = run_rf_analysis_list_SB_and_ENV_train_12Dec25[[names(run_rf_analysis_list_SB_and_ENV_train_12Dec25)[i]]], 
  #   file = paste0(getwd(), "RData_", names(run_rf_analysis_list_SB_and_ENV_train_12Dec25)[[i]], ".rds")
  # )
  
  # Force memory cleanup
  gc()
}
t1 <- Sys.time()
print(paste0("Total run time for all data combos are: ", round(as.numeric(difftime(t1, t0, units = "secs")),2), " seconds"))
stopCluster(cl)
registerDoSEQ() 
```