---
title: "Microplastic fingerprinting"
output: 
  md_document:
    variant: markdown_github
---

# SPDX-License-Identifier: AGPL-3.0-only

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/huyng/OneDrive - Toronto Metropolitan University/Microplastic/Microplastic-Fingerprinting")
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(out.width='750px', dpi=200)
```

# Documentation

This code is generating Tables and Figures for the publication: "Chemical Fingerprints of New vs. Weathered Microplastics: A Machine Learning Approach"

# Loading Packages and Functions
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# if packages are not install then install
required_packages <- c("tidyverse", "readxl", "writexl", "data.table", "crayon", "grid", "gridExtra", "stats", "vegan", "cluster", "factoextra", "FactoMineR", "ggforce", "MASS", "caret", "randomForest", "ranger", "pheatmap", "viridis", "missForest", "pROC", "VIM", "softImpute", "ggpubr", "RANN", "foreach", "doParallel", "parallel")

new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) {
  install.packages(new_packages)
}

# #########################
##   Load Libraries    ##
#########################
library(tidyverse) # Loads dplyr, tidyr, purrr, ggplot2, readr, tibble
library(readxl)
library(writexl)
library(data.table)
library(crayon)
library(grid)
library(gridExtra)
library(stats)
library(vegan)
library(cluster)
library(factoextra)
library(FactoMineR)
library(ggforce)
library(MASS)
library(caret)
library(randomForest)
library(ranger)
library(pheatmap)
library(viridis)
library(missForest)
library(pROC)
library(VIM)
library(softImpute)
library(ggpubr)
library(RANN)
library(foreach)
library(doParallel)
library(parallel)

# Load Helper Functions
# Ensure the helper script is in the same folder as this Rmd
helper_file <- "Helper Functions for Computational Workflow.R"
if(file.exists(helper_file)) {
  source(helper_file)
} else {
  warning(paste("CRITICAL WARNING: Helper file '", helper_file, "' not found in working directory:", getwd()))
}

# --- OPTIMIZATION: Helper function for Stats Reporting in Step 6 ---
print_filtering_stats <- function(results, meta_cols = c("Plastic_type", "technique", "Source", "Polymer")) {
  
  df_final <- results$filtered_data
  df_orig  <- results$original_data
  
  # Get list of feature names only
  feats_final <- setdiff(names(df_final), meta_cols)
  feats_orig  <- setdiff(names(df_orig), meta_cols)
  
  # 1. Percentage of chemical features retained (%)
  pct_features <- (length(feats_final) / length(feats_orig)) * 100
  cat(paste0("1. Percentage of chemical features retained: ", round(pct_features, 2), "%\n"))
  
  # 2. Percentage of missing values after feature reduction (%)
  if(length(feats_final) > 0) {
    vals_final <- df_final[, feats_final, drop = FALSE]
    pct_missing <- (sum(is.na(vals_final)) / prod(dim(vals_final))) * 100
    cat(paste0("2. Percentage of missing values after feature reduction: ", round(pct_missing, 2), "%\n"))
  } else {
    cat("2. Percentage of missing values after feature reduction: 0% (No features left)\n")
    vals_final <- data.frame(0) # Placeholder for signal calc
  }
  
  # 3. Percentage of signal response after feature reduction (%)
  signal_final <- sum(vals_final, na.rm = TRUE)
  signal_orig  <- sum(df_orig[, feats_orig, drop = FALSE], na.rm = TRUE)
  pct_signal   <- (signal_final / signal_orig) * 100
  cat(paste0("3. Percentage of signal response after feature reduction: ", round(pct_signal, 2), "%\n"))
}
```

# Figure 2: HCA
```{r echo=FALSE, warning=FALSE, message=FALSE}
# Data prep for HCA
# Use result from Step 7 Option 2
target_obj <- rf_results_SB_ENV_train$gc
if(is.null(target_obj)) target_obj <- rf_results_SB_ENV_train[[1]]

# target_obj is a list of folds; use the first fold
fold_obj <- if (is.list(target_obj) && !is.null(target_obj$final_imp_norm_train)) {
  target_obj
} else {
  target_obj[[1]]
}
if (is.null(fold_obj)) stop("No folds found in rf_results_SB_ENV_train.")

df <- dplyr::bind_rows(
  fold_obj$final_imp_norm_train,
  fold_obj$final_imp_norm_test
) %>%
  dplyr::mutate(Plastic_type = gsub("..", " ", Plastic_type, fixed = TRUE)) %>%
  dplyr::mutate(Plastic_type = gsub(".", " ", Plastic_type, fixed = TRUE)) %>%
  dplyr::mutate(Plastic_type_Source = paste0(Plastic_type, "-", Source))

hc_df <- df %>% 
  dplyr::select(-any_of(c("Subcategory", "Source", "Plastic_type", "Polymer", "technique"))) %>% 
  dplyr::select(where(is.numeric)) %>%
  as.data.frame()

# --- Clean: remove bad rows & handle non-finite values ---
hc_df[!is.finite(as.matrix(hc_df))] <- 0
keep <- rowSums(hc_df, na.rm = TRUE) > 0
hc_df_clean <- hc_df[keep, , drop = FALSE] 
labels_clean <- df$Plastic_type_Source[keep] 

# --- Distance Calculation ---
use_bray <- !any(hc_df_clean < 0, na.rm = TRUE)

if (use_bray) {
  d <- vegan::vegdist(hc_df_clean, method = "manhattan")
  link <- "average"
} else {
  d <- dist(hc_df_clean, method = "euclidean")
  link <- "average"
  message("Detected negative values; using Euclidean instead of Bray-Curtis.")
}

if (any(!is.finite(d))) stop("Distance matrix still has non-finite entries.")

# --- Cluster & plot ---
hca_samp <- hclust(d, method = link)
cat("cophenetic correlation coefficient (CCC) from HCA is:", cor(d, stats::cophenetic(hca_samp)), "\n")

par(cex = 1)
plot(hca_samp, labels = labels_clean, hang = -1, main = "")
```

# Table 1 and Table S6. Best imputation and normalization combinations and full results of evaluation metrics for data imputation and normalization across 7 data combinations using store-bought and environmental samples as training dataset and test on environmental samples

```{r , echo=FALSE, warning = FALSE, message=FALSE}
# Here, as an example, we extract information of imputation and normalization from SB+ENV train
ML_result <- rf_results_SB_ENV_train

# Table 1 #########################
optimal_imp_norm <- list()
for (i in seq_along(ML_result)) {
  # Read data
  data_in <- ML_result[[i]]
  
  # Get a clean name for the dataset from the filename (removes .rds extension)
  dataset_name <- names(ML_result)[[i]]
  
  for (fold_id in seq_along(ML_result[[i]])) {
    optimal_imp_norm[[i]] <-  as.data.frame(best_imp = ML_result[[i]][[fold_id]]$best_imputation,
                                            best_norm = ML_result[[i]][[fold_id]]$best_normalization,
                                            Dataset = dataset_name,
                                            Training = "SB and ENV")# change to "SB only" when extracting SB only train results)
  }
}

# Examine the dataframe before exporting to Excel
View(bind_rows(optimal_imp_norm))

# Table S6 #########################
all_imp_norm_res_list <- list()
for (i in seq_along(ML_result)) {
  # Read data
  data_in <- ML_result[[i]]
  
  # Get a clean name for the dataset from the filename (removes .rds extension)
  dataset_name <- names(ML_result)[[i]]
  
  for (fold_id in seq_along(ML_result[[i]])) {
    all_imp_norm_res_list[[i]] <-  as.data.frame(ML_result[[i]][[fold_id]]$imp_norm_res_table) %>%
      dplyr::select(-KS_p) %>% 
      dplyr::mutate(Fold = fold$fold_id,
                    OOB_MCC = round(OOB_MCC, 2),
                    ClusterRes = rond(ClusterRes, 2),
                    CorrScore = round(CorrScore, 2)) %>%
      dplyr::arrange(combined_rank) %>% 
      dplyr::mutate(Dataset = dataset_name,
                    Training = "SB and ENV") # change to "SB only" when extracting SB only train results
  }
}

# Examine the dataframe before exporting to Excel
View(bind_rows(all_imp_norm_res_list))

# Export to Excel
write_xlsx(bind_rows(all_imp_norm_res_list), path = "Imputation and normalization results.xlsx")
rm(ML_result); gc() # Clear cache to reduce RAM consumption
```

# Table 2: Optimal feature-selection methods
```{r , echo=FALSE, warning = FALSE, message=FALSE}
# Here, as an example, we extract information of imputation and normalization from SB+ENV train
data_in <- rf_results_SB_ENV_train

# 1. Define a function to extract metrics from a SINGLE fold
extract_metrics <- function(x, fold_idx) {
  
  # Helper to grab "top_X" stats easily
  get_top <- function(k) {
    sub <- x$top_importance_results[[paste0("top_", k)]]
    list(mcc = sub$eval_metrics_top$MCC_Multiclass,
         time = sub$time_top_100_50_25_10,
         n_feat = length(sub$selected_features))
  }
  
  # Create the data frame for this specific fold
  tibble::tribble(
    ~method,         ~MCC_Multiclass,       ~Computational_time,          ~Number_feat_selected,
    "all feats",     x$all_features_mcc,    x$time_all_features,          x$final_rf_all_feats$num.independent.variables,
    "sig feat",      x$mcc_sig,             x$time_sig,                   length(x$sig_selected_feats),
    "rfa feats",     x$mcc_rfa,             x$time_rfa,                   length(x$rfa_selected_feats),
    "top 100 feats", get_top(100)$mcc,      get_top(100)$time,            get_top(100)$n_feat,
    "top 50 feats",  get_top(50)$mcc,       get_top(50)$time,             get_top(50)$n_feat,
    "top 25 feats",  get_top(25)$mcc,       get_top(25)$time,             get_top(25)$n_feat,
    "top 10 feats",  get_top(10)$mcc,       get_top(10)$time,             get_top(10)$n_feat
  ) %>%
    dplyr::mutate(Folds = fold_idx, .before = 1)
}

feat_select_summary <- list()
# Summarize feature selections
for (i in seq_along(data_in)) {
  feat_select_summary[[i]] <- purrr::imap_dfr(data_in[[i]], extract_metrics) %>%
    dplyr::group_by(method) %>%
    summarise(
      MCC_mean = round(mean(MCC_Multiclass, na.rm = TRUE), 2),
      MCC_sd = round(sd(MCC_Multiclass, na.rm = TRUE), 2),
      Computational_time_mean = round(mean(Computational_time, na.rm = TRUE), 2),
      Computational_time_sd = round(sd(Computational_time, na.rm = TRUE), 2),
      Number_feat_selected_mean = round(mean(Number_feat_selected, na.rm = TRUE), 2),
      Number_feat_selected_sd = round(sd(Number_feat_selected, na.rm = TRUE), 2)
    ) %>%
    dplyr::arrange(desc(MCC_mean)) %>%
    # Add the dataset name column so we know which file this came from
    dplyr::mutate(Dataset = names(data_in)[[i]], .before = 1)
}

# Final Binding ---
# Remove any NULL entries (in case of errors) and bind rows
final_summary_stats <- bind_rows(feat_select_summary)

View(final_summary_stats)
```

# Table S3. Descriptive statistics of retention time of five HPLC benchmark compounds.

```{r , echo=FALSE, warning = FALSE, message=FALSE}
bm_mz   <- c(156.0957, 198.1723, 246.2568, 305.1956, 342.1722)
bm_name <- c("Acetaminophen-d4 (m/z 156.0957)", "dTEP (m/z 198.1723)", "dTPrP (m/z 246.2568)",
             "13C-6PPD (m/z 305.1956)", "dTPP (m/z 342.1722)")

tol <- 0.0005

hplc_raw_data <- list(batch0 = hplc_batch0_step1.1,
                      batch1 = hplc_batch1_step1.1,
                      batch2 = hplc_batch2_step1.1)
benchmark_summary <- list()
# Build one long data frame with a facet label per target
for (i in seq_along(hplc_raw_data)) {
  benchmark_summary[[i]] <- map2_dfr(bm_mz, bm_name, ~
  hplc_raw_data[[i]] %>%
    filter(m.z >= (.x - tol), m.z <= (.x + tol)) %>%
    transmute(RT = as.numeric(RT), bm_name = .y)
  ) %>% 
    dplyr::mutate(Batch_number = names(hplc_raw_data)[[i]]) %>%
    dplyr::filter(!is.na(RT))
}

tableS3 <- bind_rows(benchmark_summary) %>% 
  # dplyr::mutate(Batch_number = names(hplc_raw_data)[[i]]) %>%
  dplyr::filter(!is.na(RT)) %>% 
  dplyr::group_by(bm_name) %>%
  dplyr::summarise(min_rt = summary(RT)[[1]],
                   first_quan_rt = summary(RT)[[2]],
                   median_rt = summary(RT)[[3]],
                   mean_rt = summary(RT)[[4]],
                   third_quan_rt = summary(RT)[[5]],
                   max_rt = summary(RT)[[6]],
                   max_minus_min = (summary(RT)[[6]] - summary(RT)[[1]]),
                   sd_rt = stats::sd(RT)
  )
```

# Table S4. Number of unique aligned compounds that are associated with HPLC benchmarks at two sets of RT windows were applied to split HPLC peaks into different RT sets based on inspection of histogram plots of RT distribution in all HPLC samples and distribution of RT of benchmark compounds

```{r , echo=FALSE, warning = FALSE, message=FALSE}
# Testing different combinations of RT window for 2 parts of the chromatogram ----------
combi <- tidyr::crossing(
  # c(0.1,0.15,0.2, 0.25, 0.3, 0.35, 
                           0.4, c(
                             # 0.02,0.03,0.04, 0.05, 0.06, 0.07, 0.08, 0.09, 0.1, 0.11, 0.12, 0.13, 0.14, 0.15, 0.16, 0.17, 0.18, 0.19, 0.2,
                             0.21, 0.22, 0.23, 0.24, 0.25, 0.26, 0.27, 0.28, 0.29, 0.3, 0.31, 0.32, 0.33, 0.34, 0.35, 0.36, 0.37, 0.38, 0.39, 0.4))
rt_test_list <- list()
for (row in 1:nrow(combi)) {
  df <- grouping_comp(combined_df2,
                      rtthres1 = as.numeric(combi[row,][1]),
                      rtthres2 = as.numeric(combi[row,][2]),
                      mzthres = 0.0005,
                      type = "HPLCTOFMS")
  
  rt_test_list[[paste0(as.numeric(combi[row,][1]), "_", as.numeric(combi[row,][2]))]] <- df
  
  rt_6 <- df %>% filter(RT <= 6)
  rt_higher6 <- df %>% filter(RT > 6)
  
  print(paste0("RT window ", as.numeric(combi[row,][1]), " No. unique compounds that belongs to Benchmark_1 Acetaminophen-d4 : ", 
               length(unique((rt_6 %>% 
                      filter(m.z >= (156.0957 - 0.0001) &
                               m.z <= (156.0957 + 0.0001)))$Feature))))
  
  print(paste0("RT window ", as.numeric(combi[row,][1]), " No. of unique compounds : ", length(unique(rt_6$Feature))))
  
  print(paste0("RT window ", as.numeric(combi[row,][2]), " No. unique compounds that belongs to Benchmark_2 dTEP : ", 
               length(unique((rt_higher6 %>% 
                                filter(m.z >= (198.1723 - 0.0001) &
                                         m.z <= (198.1723 + 0.0001)))$Feature))))
  print(paste0("RT window ", as.numeric(combi[row,][2]), " No. unique compounds that belongs to Benchmark_3 dTPrP : ", 
               length(unique((rt_higher6 %>% 
                      filter(m.z >= (246.2568 - 0.0001) &
                               m.z <= (246.2568 + 0.0001)))$Feature))))
  print(paste0("RT window ", as.numeric(combi[row,][2]), " No. unique compounds that belongs to Benchmark_4 13C-6PPD : ", 
               length(unique((rt_higher6 %>% 
                      filter(m.z >= (305.1956 - 0.0001) &
                               m.z <= (305.1956 + 0.0001)))$Feature))))
  print(paste0("RT window ", as.numeric(combi[row,][2]), " No. unique compounds that belongs to Benchmark_5 dTPP : ", 
               length(unique((rt_higher6 %>%
                      filter(m.z >= (342.1722 - 0.0001) &
                               m.z <= (342.1722 + 0.0001)))$Feature))))

  print(paste0("RT window ", as.numeric(combi[row,][2]), " No. of unique compounds : ", length(unique(rt_higher6$Feature))))
}
```

# Table S5. Summary of missing values present in all seven data combinations across three feature reduction approaches.

```{r , echo=FALSE, warning = FALSE, message=FALSE}
# As an example, here we examine "regular 80% rule" feature reduction
### BEFORE FEATURE REDUCTION
df_in <- gc

# 1. Process Data ONCE (Consolidated Pivot)
df_wide <- df_in %>% 
  dplyr::select(File, Plastic_type, Source, Polymer, Feature, technique, Values) %>%
  dplyr::group_by(File, Source, Polymer, Plastic_type, technique, Feature) %>%
  dplyr::summarise(Values = mean(Values, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = Feature, values_from = Values) %>%
  tibble::column_to_rownames(var = "File") 

df_numeric <- df_wide %>% dplyr::select(where(is.numeric))

# --- Basic Statistics ---
cat("Number of chemical features: ", ncol(df_numeric), "\n")
missing_pct <- sum(is.na(df_numeric)) / prod(dim(df_numeric)) * 100
cat("Percentage of missing values (%):", round(missing_pct, 2), "\n")

# --- Correlations ---
stats_df <- tibble::tibble(
  Feature = colnames(df_numeric),
  num_missing_values = colSums(is.na(df_numeric)),
  mean_response = colMeans(df_numeric, na.rm = TRUE)
)

if (all(c("m.z", "RT") %in% colnames(df_in))) {
  feature_meta <- df_in %>%
    dplyr::group_by(Feature) %>%
    dplyr::summarise(
      m_z = mean(m.z, na.rm = TRUE),
      retention_time = mean(RT, na.rm = TRUE),
      .groups = "drop"
    )
  analysis_df <- stats_df %>% dplyr::left_join(feature_meta, by = "Feature")
  cat("Pearson (Mean Response vs Missing):", round(cor(analysis_df$mean_response, analysis_df$num_missing_values, use = "pairwise"), 2), "\n")
  cat("Pearson (m/z vs Missing):          ", round(cor(analysis_df$m_z, analysis_df$num_missing_values, use = "pairwise"), 2), "\n")
  cat("Pearson (RT vs Missing):           ", round(cor(analysis_df$retention_time, analysis_df$num_missing_values, use = "pairwise"), 2), "\n")
} else {
  cat("Pearson (Mean Response vs Missing):", round(cor(stats_df$mean_response, stats_df$num_missing_values, use = "pairwise.complete.obs"), 2), "\n")
  message("NOTE: 'm.z' and 'RT' columns missing. Skipping specific correlations.")
}

### AFTER FEATURE REDUCTION
results <- shared_feature_filtering(gc, feat_reduc_mode = "regular80")
print_filtering_stats(results)
```

# Table S7. Summary of significant compounds identified via pair-wise tests between plastic categories from k-fold cross validation across seven data combinations

```{r , echo=FALSE, warning = FALSE, message=FALSE}
sig_feat_list_SB_ENV <- list()
k <- 1
for (i in seq_along(run_rf_analysis_list_SB_and_ENV_train_pairwise_17dec)) {
  current_data <- run_rf_analysis_list_SB_and_ENV_train_pairwise_17dec[[i]]
  for (fold_idx in seq_along(current_data)) {
    fold_content <- current_data[[fold_idx]]
    if (!is.null(fold_content$pvalue_corrected_df)) {
      sig_feat_list_SB_ENV[[k]] <- fold_content$pvalue_corrected_df %>% 
        dplyr::mutate(Dataset = names(run_rf_analysis_list_SB_and_ENV_train_pairwise_17dec)[[i]])
      
      # Increment the output counter
      k <- k + 1
    }
  }
}

# 2. Prepare the reference tables (Calculate means)
# Calculate means for GC data
gc_means <- gc %>%
  group_by(Feature) %>%
  summarise(
    mean_mz_gc = mean(m.z, na.rm = TRUE),
    mean_rt_gc = mean(RT, na.rm = TRUE)
  )

# Calculate means for HPLC data
hplc_means <- hplc %>%
  group_by(Feature) %>%
  summarise(
    mean_mz_hplc = mean(m.z, na.rm = TRUE),
    mean_rt_hplc = mean(RT, na.rm = TRUE)
  )

sig_feat_SB_ENVtrain <- bind_rows(sig_feat_list_SB_ENV) %>% 
  dplyr::select(-pval, -adjusted_pvalue) %>%
  # 2. Remove duplicates based on Dataset, Feature, and Categories
  # This ensures that for every Dataset, you only have unique Feature/Category combos
  dplyr::distinct(Dataset, Feature, Category_1, Category_2, .keep_all = TRUE) %>%
  # Join with the GC means
  left_join(gc_means, by = "Feature") %>%
  
  # Join with the HPLC means
  left_join(hplc_means, by = "Feature") %>%
  
  # Create the new columns based on the Feature name
  dplyr::mutate(
    `mean(m.z)` = case_when(
      str_detect(Feature, "ATDGCMS") ~ mean_mz_gc,
      str_detect(Feature, "HPLCTOFMS") ~ mean_mz_hplc,
      TRUE ~ NA_real_ # Sets to NaN/NA if neither is found
    ),
    `mean(RT)` = case_when(
      str_detect(Feature, "ATDGCMS") ~ mean_rt_gc,
      str_detect(Feature, "HPLCTOFMS") ~ mean_rt_hplc,
      TRUE ~ NA_real_
    )
  ) %>%
  
  # Remove the temporary helper columns
  dplyr::select(-mean_mz_gc, -mean_rt_gc, -mean_mz_hplc, -mean_rt_hplc)


write_xlsx(rbind(sig_feat_SB_ENVtrain, sig_feat_SBtrain), path = "Table S7_significant features SB train and SB_ENV train_old feature reduction 12Dec2025.xlsx")
```

# Table S8. Significant compounds identified via pair-wise tests between plastic categories across seven data combinations

```{r echo=FALSE, warning=FALSE, message=FALSE}
list_pairwise_df <- list()
for (i in 1:length(run_rf_analysis_list_SB_and_ENV_train_02Aug2025)) {
  pw <- pairwise_significance_tests(run_rf_analysis_list_SB_and_ENV_train_02Aug2025[[i]]$final_imp_norm_dat, 
                                    group_col = "Plastic_type", start_col_index = 6)
  list_pairwise_df[[i]] <- pw$uncorrected %>%
    mutate(adjusted_p = p.adjust(pval, method = "bonferroni")) %>%
    filter(adjusted_p < 0.05) %>%
    arrange(adjusted_p) %>%
    dplyr::mutate(Datasets = names(run_rf_analysis_list_SB_and_ENV_train_02Aug2025)[i])
}

pw_df <- bind_rows(list_pairwise_df)
write_xlsx(pw_df, path = "Table S8_Significant compounds identified via pair-wise tests between plastic categories.xlsx")
```

# Figure S1: Histogram illustrating the distribution of retention time (RT) of benchmark compound (Toluene-d8) that was injected in each run of ATD-GC-MS.

```{r echo=FALSE, warning=FALSE, message=FALSE}
# ATDGCMS data ------------------
# Filter data
bm_df_gcms <- atdgcms_step1.1 %>%
  filter(m.z >= 98, m.z <= 98.5, RT < 13.6) %>%
  mutate(RT = as.numeric(RT))

# Plot in ggplot2 style to match your multi-panel code
p <- ggplot(bm_df_gcms, aes(x = RT)) +
  geom_histogram(bins = 30) +
  labs(title = "Toluene-D8",
       x = "Retention time (seconds)",
       y = "Count") +
  theme_bw(base_size = 25) +
  theme(
    plot.margin = ggplot2::margin(t = 5, r = 40, b = 5, l = 10),  # more right padding
    plot.title  = element_text(hjust = 0.5)                       # center title
  )
p
# Save at 600 DPI, 20x10 inches (same output size as your PNG device)
ggsave(
  filename = sprintf("Figure S1_Histogram ATDGCMS benchmark Toluene D8_%s.png",
                     format(Sys.time(), "%Y-%m-%d")),
  plot = p,
  device = "png",
  dpi = 600,
  width = 20,
  height = 10,
  units = "in"
)
```

# Figure S2: Histogram illustrating distribution of retention time (RT) of five benchmark compounds that were injected in each run of HPLC-qTOF-MS. All histograms demonstrated uniform distribution with RT varied in the range of ± 0.4 (for Acetaminophen-d4) and ± 0.1 (for dTEP, dTPrP, 13C-6PPD and dTPP).

```{r echo=FALSE, warning=FALSE, message=FALSE}
# HPLC data ----------------------
bm_mz   <- c(156.0957, 198.1723, 246.2568, 305.1956, 342.1722)
bm_name <- c("Acetaminophen-d4 (m/z 156.0957)", "dTEP (m/z 198.1723)", "dTPrP (m/z 246.2568)",
             "13C-6PPD (m/z 305.1956)", "dTPP (m/z 342.1722)")

tol <- 0.0005

# Build one long data frame with a facet label per target
panel_df <- map2_dfr(bm_mz, bm_name, ~
  hplc_batch2_step1.1 %>%
    filter(m.z >= (.x - tol), m.z <= (.x + tol)) %>%
    transmute(RT = as.numeric(RT), panel = .y)
) %>% 
  dplyr::filter(!is.na(RT))

p <- ggplot(panel_df, aes(x = RT)) +
  geom_histogram(bins = 100) +
  facet_wrap(~ panel, ncol = 2, scales = "free") +     # free x AND y per facet
  scale_x_continuous(expand = expansion(mult = 0)) +   # no x padding: tight to min/max
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +  # tiny headroom on y
  labs(x = "Retention time (seconds)", y = "Count") +
  theme_bw(base_size = 25) +
  theme(
    plot.margin = ggplot2::margin(t = 5, r = 5, b = 5, l = 20)  # extra left padding
  )

p

ggsave(filename = sprintf("Figure S2_Histogram HPLCTOFMS benchmarks_%s.png",
                     format(Sys.time(), "%Y-%m-%d")), 
       plot = p, 
       device = "png",
       dpi = 600,        # Set DPI to 600
       width = 20,       # Width in inches (adjust as needed)
       height = 10,       # Height in inches (adjust as needed)
       units = "in")
```

# Figure S3 & S4: HCA clustering of HPLC-QToF-MS dataset (using both store-bought and environmental plastics) with plastic categories and their corresponding sources.

Please refer to step 8 HCA in the "Complete Workflow.Rmd" file on the repo to create HCA clustering figures

# Figure S5: Retention time and mass-to-charge of HPLC dataset’s original features and RFA selected features

```{r echo=FALSE, warning=FALSE, message=FALSE}
data_in <-  rf_results_SB_ENV_train$hplc

## Get all features 
all_feats_rt_mz <- hplc %>% 
  dplyr::filter(Feature %in% colnames(data_in$fold1$final_imp_norm_train[, 6:ncol(data_in$fold1$final_imp_norm_train)])) %>%
  dplyr::group_by(Feature) %>%
  dplyr::summarize(mean_RT = mean(RT),
                   mean_mz = mean(m.z))

all_feats_rt_mz_plot <- ggplot(data =all_feats_rt_mz) +
  geom_point(aes(x= mean_RT, y = mean_mz)) +
  scale_x_continuous(limits = c(0, 18)) +
  scale_y_continuous(limits = c(0, 800)) +
  labs(x = NULL,
       y = NULL,
       title = "All features - HPLC") +
  theme_light(base_size = 25) +
  theme(
    plot.title = element_text(hjust = 0.5)  # centers the title
  )
all_feats_rt_mz_plot

# Get RFA features
RFA_feats_rt_mz <- hplc %>% 
  dplyr::filter(Feature %in% unique(c(data_in$fold5$rfa_selected_feats, data_in$fold4$rfa_selected_feats, data_in$fold3$rfa_selected_feats))) %>%
  dplyr::group_by(Feature) %>%
  dplyr::summarize(mean_RT = mean(RT),
                   mean_mz = mean(m.z))

RFA_feats_rt_mz_plot <- ggplot(data =RFA_feats_rt_mz) +
  geom_point(aes(x= mean_RT, y = mean_mz)) +
  scale_x_continuous(limits = c(0, 18)) +
  scale_y_continuous(limits = c(0, 800)) +
  labs(x = NULL,
       y = NULL,
       title = "RFA features - HPLC") +
  theme_light(base_size = 25) +
  theme(
    plot.title = element_text(hjust = 0.5)  # centers the title
  )
RFA_feats_rt_mz_plot

y <- textGrob("Mass-to-charge", rot = 90, gp = gpar(fontsize = 20))
x <- textGrob("Retention time (minutes)", gp = gpar(fontsize = 20))

ggsave(filename = paste0("Figure S5_mz and RT of HPLC data before and after RFA feature selection_", format(Sys.time(), "%d-%m-%Y"),".png"), 
       plot = grid.arrange(all_feats_rt_mz_plot, 
             RFA_feats_rt_mz_plot, 
             bottom = x, left =y), 
       device = "png",
       dpi = 300,        # Set DPI to 600
       width = 15,       # Width in inches (adjust as needed)
       height = 10,       # Height in inches (adjust as needed)
       units = "in") 
```

# Figure S6 -> S12: Confusion matrices of training and testing on environmental plastics versus. training on store-bought plastics and testing on environmental plastics - 7 data combinations

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Here we use GC again as an example. You can change the fold number to view confusion matrices from different folds

sb_train_only <- rf_results_SB_ENV_train$gc_USSB_train$fold1$conf_mat_rfa
sb_and_env_train <- rf_results_SB_ENV_train$gc_SB_and_ENV_train$fold1$conf_mat_rfa

ggsave(filename = paste0("Confusion matrix_SB train only_", "GC", "_", format(Sys.time(), "%d-%m-%Y"), ".png"), 
       plot = sb_train_only, 
       device = "png",
       dpi = 300,        # Set DPI to 600
       width = 15,       # Width in inches (adjust as needed)
       height = 8,       # Height in inches (adjust as needed)
       units = "in")

ggsave(filename = paste0("Confusion matrix_SB and ENV train", "GC", "_", format(Sys.time(), "%d-%m-%Y"), ".png"), 
       plot = sb_and_env_train, 
       device = "png",
       dpi = 300,        # Set DPI to 600
       width = 15,       # Width in inches (adjust as needed)
       height = 8,       # Height in inches (adjust as needed)
       units = "in")
```