---
title: "Microplastic fingerprinting"
output: 
  md_document:
    variant: markdown_github
---

# SPDX-License-Identifier: AGPL-3.0-only

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/huyng/OneDrive - Toronto Metropolitan University/Microplastic/Microplastic-Fingerprinting")
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(out.width='750px', dpi=200)
```

# Documentation

This code is generating Tables and Figures for the publication: "Chemical Fingerprints of New vs. Weathered Microplastics: A Machine Learning Approach"

# Loading Packages and Functions
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# if packages are not install then install
required_packages <- c("tidyverse", "readxl", "writexl", "data.table", "crayon", "grid", "gridExtra", "stats", "vegan", "cluster", "factoextra", "FactoMineR", "ggforce", "MASS", "caret", "randomForest", "ranger", "pheatmap", "viridis", "missForest", "pROC", "VIM", "softImpute", "ggpubr", "RANN", "foreach", "doParallel", "parallel")

new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) {
  install.packages(new_packages)
}

# #########################
##   Load Libraries    ##
#########################
library(tidyverse) # Loads dplyr, tidyr, purrr, ggplot2, readr, tibble
library(readxl)
library(writexl)
library(data.table)
library(crayon)
library(grid)
library(gridExtra)
library(stats)
library(vegan)
library(cluster)
library(factoextra)
library(FactoMineR)
library(ggforce)
library(MASS)
library(caret)
library(randomForest)
library(ranger)
library(pheatmap)
library(viridis)
library(missForest)
library(pROC)
library(VIM)
library(softImpute)
library(ggpubr)
library(RANN)
library(foreach)
library(doParallel)
library(parallel)

# Load Helper Functions
# Ensure the helper script is in the same folder as this Rmd
helper_file <- "Helper Functions for Computational Workflow.R"
if(file.exists(helper_file)) {
  source(helper_file)
} else {
  warning(paste("CRITICAL WARNING: Helper file '", helper_file, "' not found in working directory:", getwd()))
}

# --- OPTIMIZATION: Helper function for Stats Reporting in Step 6 ---
print_filtering_stats <- function(results, meta_cols = c("Plastic_type", "technique", "Source", "Polymer")) {
  
  df_final <- results$filtered_data
  df_orig  <- results$original_data
  
  # Get list of feature names only
  feats_final <- setdiff(names(df_final), meta_cols)
  feats_orig  <- setdiff(names(df_orig), meta_cols)
  
  # 1. Percentage of chemical features retained (%)
  pct_features <- (length(feats_final) / length(feats_orig)) * 100
  cat(paste0("1. Percentage of chemical features retained: ", round(pct_features, 2), "%\n"))
  
  # 2. Percentage of missing values after feature reduction (%)
  if(length(feats_final) > 0) {
    vals_final <- df_final[, feats_final, drop = FALSE]
    pct_missing <- (sum(is.na(vals_final)) / prod(dim(vals_final))) * 100
    cat(paste0("2. Percentage of missing values after feature reduction: ", round(pct_missing, 2), "%\n"))
  } else {
    cat("2. Percentage of missing values after feature reduction: 0% (No features left)\n")
    vals_final <- data.frame(0) # Placeholder for signal calc
  }
  
  # 3. Percentage of signal response after feature reduction (%)
  signal_final <- sum(vals_final, na.rm = TRUE)
  signal_orig  <- sum(df_orig[, feats_orig, drop = FALSE], na.rm = TRUE)
  pct_signal   <- (signal_final / signal_orig) * 100
  cat(paste0("3. Percentage of signal response after feature reduction: ", round(pct_signal, 2), "%\n"))
}
```

# Figure 2: HCA
```{r echo=FALSE, warning=FALSE, message=FALSE}
# Data prep for HCA
# Use result from Step 7 Option 2
target_obj <- rf_results_SB_ENV_train$gc
if(is.null(target_obj)) target_obj <- rf_results_SB_ENV_train[[1]]

# target_obj is a list of folds; use the first fold
fold_obj <- if (is.list(target_obj) && !is.null(target_obj$final_imp_norm_train)) {
  target_obj
} else {
  target_obj[[1]]
}
if (is.null(fold_obj)) stop("No folds found in rf_results_SB_ENV_train.")

df <- dplyr::bind_rows(
  fold_obj$final_imp_norm_train,
  fold_obj$final_imp_norm_test
) %>%
  dplyr::mutate(Plastic_type = gsub("..", " ", Plastic_type, fixed = TRUE)) %>%
  dplyr::mutate(Plastic_type = gsub(".", " ", Plastic_type, fixed = TRUE)) %>%
  dplyr::mutate(Plastic_type_Source = paste0(Plastic_type, "-", Source))

hc_df <- df %>% 
  dplyr::select(-any_of(c("Subcategory", "Source", "Plastic_type", "Polymer", "technique"))) %>% 
  dplyr::select(where(is.numeric)) %>%
  as.data.frame()

# --- Clean: remove bad rows & handle non-finite values ---
hc_df[!is.finite(as.matrix(hc_df))] <- 0
keep <- rowSums(hc_df, na.rm = TRUE) > 0
hc_df_clean <- hc_df[keep, , drop = FALSE] 
labels_clean <- df$Plastic_type_Source[keep] 

# --- Distance Calculation ---
use_bray <- !any(hc_df_clean < 0, na.rm = TRUE)

if (use_bray) {
  d <- vegan::vegdist(hc_df_clean, method = "manhattan")
  link <- "average"
} else {
  d <- dist(hc_df_clean, method = "euclidean")
  link <- "average"
  message("Detected negative values; using Euclidean instead of Bray-Curtis.")
}

if (any(!is.finite(d))) stop("Distance matrix still has non-finite entries.")

# --- Cluster & plot ---
hca_samp <- hclust(d, method = link)
cat("cophenetic correlation coefficient (CCC) from HCA is:", cor(d, stats::cophenetic(hca_samp)), "\n")

par(cex = 1)
plot(hca_samp, labels = labels_clean, hang = -1, main = "")
```

# Table 1: Best imputation and normalization combinations for each of the seven analytical data combinations on two data splitting scenarios

```{r , echo=FALSE, warning = FALSE, message=FALSE}

```

# Table 2 and Table S6. Optimal feature-selection methods and full results of evaluation metrics for data imputation and normalization across 7 data combinations using store-bought and environmental samples as training dataset and test on environmental samples

data_in <- rf_results_SB_ENV_train$gc

result_gc_SBENV_train <- list()
i <- 1
for (fold in data_in) {
    result_gc_SBENV_train[[i]] <- as.data.frame(fold$imp_norm_res_table) %>%
        dplyr::mutate(Fold = fold$fold_id) %>%
        dplyr::arrange(combined_rank) %>% dplyr::select(-KS_p)
    i <- i + 1
}

# 3. Apply function to all data and combine
result_gc_SBENV_train <- bind_rows(result_gc_SBENV_train) %>% 
    dplyr::mutate(Dataset = "GC",
                  Training = "SB and ENV")
rm(data_in); gc()
```{r , echo=FALSE, warning = FALSE, message=FALSE}
# Here, as an example, we extract information of imputation and normalization from SB+ENV train
ML_result <- rf_results_SB_ENV_train

impxnorm_list <- list()
i <- 1
for (i in seq_along(ML_result)) {
  # Read data
  data_in <- ML_result[[i]]
  
  # Get a clean name for the dataset from the filename (removes .rds extension)
  dataset_name <- names(ML_result)[[i]]
  
  for (fold in ML_result[[i]]) {
    impxnorm_list[[i]] <-  as.data.frame(fold$imp_norm_res_table) %>%
      dplyr::select(-KS_p) %>% 
      dplyr::mutate(Fold = fold$fold_id,
                    OOB_MCC = round(OOB_MCC, 2),
                    ClusterRes = rond(ClusterRes, 2),
                    CorrScore = round(CorrScore, 2)) %>%
      dplyr::arrange(combined_rank) %>% 
      dplyr::mutate(Dataset = dataset_name)
    i <- i + 1
  }
}

write_xlsx(bind_rows(impxnorm_list), path = "TableS6_18Dec2025- Test RF and kNN imputation_old feature reduc 12Dec2025.xlsx")

# 3. Apply function to all data and combine
Result_icp_USSB_train <- bind_rows(Result_icp_USSB_train) %>% 
  dplyr::mutate(Dataset = "ICP",
                Training = "SB only")
rm(data_in); gc()

all_imp_norm_res_12Dec2025 <- rbind(Result_gc_hplc_icp_SB_and_ENV_train,
                                    Result_gc_hplc_icp_USSB_train,
                                    Result_gc_hplc_SB_and_ENV_train,
                                    Result_gc_hplc_USSB_train,
                                    Result_gc_icp_SB_and_ENV_train,
                                    Result_gc_icp_USSB_train,
                                    Result_gc_SB_and_ENV_train,
                                    Result_gc_USSB_train,
                                    Result_hplc_icp_SB_and_ENV_train,
                                    Result_hplc_icp_USSB_train,
                                    Result_hplc_SB_and_ENV_train,
                                    Result_hplc_USSB_train,
                                    Result_icp_SB_and_ENV_train,
                                    Result_icp_USSB_train)

base::saveRDS(all_imp_norm_res_12Dec2025, file = paste0(getwd(), "/ML results with 7 data combos using old feature reduction (12Dec2025)/", "Table S6_all_imp_norm_res_12Dec2025.rds"))
```

# Table S3. Descriptive statistics of retention time of five HPLC benchmark compounds.

# Table S4. Number of unique aligned compounds that are associated with HPLC benchmarks at two sets of RT windows were applied to split HPLC peaks into different RT sets based on inspection of histogram plots of RT distribution in all HPLC samples and distribution of RT of benchmark compounds

# Table S5. Summary of missing values present in all seven data combinations across three feature reduction approaches.


# Table S7. Summary of significant compounds identified via pair-wise tests between plastic categories from k-fold cross validation across seven data combinations
```{r , echo=FALSE, warning = FALSE, message=FALSE}
sig_feat_list_SB_ENV <- list()
k <- 1
for (i in seq_along(run_rf_analysis_list_SB_and_ENV_train_pairwise_17dec)) {
  current_data <- run_rf_analysis_list_SB_and_ENV_train_pairwise_17dec[[i]]
  for (fold_idx in seq_along(current_data)) {
    fold_content <- current_data[[fold_idx]]
    if (!is.null(fold_content$pvalue_corrected_df)) {
      sig_feat_list_SB_ENV[[k]] <- fold_content$pvalue_corrected_df %>% 
        dplyr::mutate(Dataset = names(run_rf_analysis_list_SB_and_ENV_train_pairwise_17dec)[[i]])
      
      # Increment the output counter
      k <- k + 1
    }
  }
}

# 2. Prepare the reference tables (Calculate means)
# Calculate means for GC data
gc_means <- gc %>%
  group_by(Feature) %>%
  summarise(
    mean_mz_gc = mean(m.z, na.rm = TRUE),
    mean_rt_gc = mean(RT, na.rm = TRUE)
  )

# Calculate means for HPLC data
hplc_means <- hplc %>%
  group_by(Feature) %>%
  summarise(
    mean_mz_hplc = mean(m.z, na.rm = TRUE),
    mean_rt_hplc = mean(RT, na.rm = TRUE)
  )

sig_feat_SB_ENVtrain <- bind_rows(sig_feat_list_SB_ENV) %>% 
  dplyr::select(-pval, -adjusted_pvalue) %>%
  # 2. Remove duplicates based on Dataset, Feature, and Categories
  # This ensures that for every Dataset, you only have unique Feature/Category combos
  dplyr::distinct(Dataset, Feature, Category_1, Category_2, .keep_all = TRUE) %>%
  # Join with the GC means
  left_join(gc_means, by = "Feature") %>%
  
  # Join with the HPLC means
  left_join(hplc_means, by = "Feature") %>%
  
  # Create the new columns based on the Feature name
  dplyr::mutate(
    `mean(m.z)` = case_when(
      str_detect(Feature, "ATDGCMS") ~ mean_mz_gc,
      str_detect(Feature, "HPLCTOFMS") ~ mean_mz_hplc,
      TRUE ~ NA_real_ # Sets to NaN/NA if neither is found
    ),
    `mean(RT)` = case_when(
      str_detect(Feature, "ATDGCMS") ~ mean_rt_gc,
      str_detect(Feature, "HPLCTOFMS") ~ mean_rt_hplc,
      TRUE ~ NA_real_
    )
  ) %>%
  
  # Remove the temporary helper columns
  dplyr::select(-mean_mz_gc, -mean_rt_gc, -mean_mz_hplc, -mean_rt_hplc)


write_xlsx(rbind(sig_feat_SB_ENVtrain, sig_feat_SBtrain), path = "Table S7_significant features SB train and SB_ENV train_old feature reduction 12Dec2025.xlsx")
```

# Table S8. Significant compounds identified via pair-wise tests between plastic categories across seven data combinations
```{r echo=FALSE, warning=FALSE, message=FALSE}
list_pairwise_df <- list()
for (i in 1:length(run_rf_analysis_list_SB_and_ENV_train_02Aug2025)) {
  pw <- pairwise_significance_tests(run_rf_analysis_list_SB_and_ENV_train_02Aug2025[[i]]$final_imp_norm_dat, 
                                    group_col = "Plastic_type", start_col_index = 6)
  list_pairwise_df[[i]] <- pw$uncorrected %>%
    mutate(adjusted_p = p.adjust(pval, method = "bonferroni")) %>%
    filter(adjusted_p < 0.05) %>%
    arrange(adjusted_p) %>%
    dplyr::mutate(Datasets = names(run_rf_analysis_list_SB_and_ENV_train_02Aug2025)[i])
}

pw_df <- bind_rows(list_pairwise_df)
write_xlsx(pw_df, path = "Table S8_Significant compounds identified via pair-wise tests between plastic categories.xlsx")
```

# Figure S1: Histogram illustrating the distribution of retention time (RT) of benchmark compound (Toluene-d8) that was injected in each run of ATD-GC-MS.

```{r echo=FALSE, warning=FALSE, message=FALSE}
```

# Figure S2: Histogram illustrating distribution of retention time (RT) of five benchmark compounds that were injected in each run of HPLC-qTOF-MS. All histograms demonstrated uniform distribution with RT varied in the range of ± 0.4 (for Acetaminophen-d4) and ± 0.1 (for dTEP, dTPrP, 13C-6PPD and dTPP).

```{r echo=FALSE, warning=FALSE, message=FALSE}
```

# Figure S3: HCA clustering of HPLC-QToF-MS dataset (using both store-bought and environmental plastics) with plastic categories and their corresponding sources.

```{r echo=FALSE, warning=FALSE, message=FALSE}
```

# Figure S4: HCA clustering of ICP-MS/MS dataset (using both store-bought and environmental plastics) with plastic categories and their corresponding sources.

```{r echo=FALSE, warning=FALSE, message=FALSE}
```

# Figure S5: Retention time and mass-to-charge of HPLC dataset’s original features and RFA selected features

```{r echo=FALSE, warning=FALSE, message=FALSE}
```

# Figure S6 -> S12: Confusion matrices of training and testing on environmental plastics versus. training on store-bought plastics and testing on environmental plastics - 7 data combinations

```{r echo=FALSE, warning=FALSE, message=FALSE}
sb_train_only <- run_rf_analysis_list_SB_train$gc_USSB_train$conf_mat_sig
sb_and_env_train <- run_rf_analysis_list_SB_and_ENV_train_02Aug2025$gc_SB_and_ENV_train$conf_mat_rfa

ggsave(filename = paste0("Figure 4_confusion matrix_SB train only_", "GC", "_", format(Sys.time(), "%d-%m-%Y"), ".png"), 
       plot = sb_train_only, 
       device = "png",
       dpi = 300,        # Set DPI to 600
       width = 15,       # Width in inches (adjust as needed)
       height = 8,       # Height in inches (adjust as needed)
       units = "in")

ggsave(filename = paste0("Confusion matrix_SB and ENV train", "GC", "_", format(Sys.time(), "%d-%m-%Y"), ".png"), 
       plot = sb_and_env_train, 
       device = "png",
       dpi = 300,        # Set DPI to 600
       width = 15,       # Width in inches (adjust as needed)
       height = 8,       # Height in inches (adjust as needed)
       units = "in")
```