---
title: "Microplastic fingerprinting"
output: 
  md_document:
    variant: markdown_github
---

# SPDX-License-Identifier: AGPL-3.0-only

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/huyng/OneDrive - Toronto Metropolitan University/Microplastic/Microplastic-Fingerprinting")
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(out.width='750px', dpi=200)
```

# Documentation

This code is generating Tables and Figures for the publication: "Chemical Fingerprints of New vs. Weathered Microplastics: A Machine Learning Approach"

# Loading Packages and Functions
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# if packages are not install then install
required_packages <- c("tidyverse", "readxl", "writexl", "data.table", "crayon", "grid", "gridExtra", "stats", "vegan", "cluster", "factoextra", "FactoMineR", "ggforce", "MASS", "caret", "randomForest", "ranger", "pheatmap", "viridis", "missForest", "pROC", "VIM", "softImpute", "ggpubr", "RANN", "foreach", "doParallel", "parallel")

new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) {
  install.packages(new_packages)
}

# #########################
##   Load Libraries    ##
#########################
library(tidyverse) # Loads dplyr, tidyr, purrr, ggplot2, readr, tibble
library(readxl)
library(writexl)
library(data.table)
library(crayon)
library(grid)
library(gridExtra)
library(stats)
library(vegan)
library(cluster)
library(factoextra)
library(FactoMineR)
library(ggforce)
library(MASS)
library(caret)
library(randomForest)
library(ranger)
library(pheatmap)
library(viridis)
library(missForest)
library(pROC)
library(VIM)
library(softImpute)
library(ggpubr)
library(RANN)
library(foreach)
library(doParallel)
library(parallel)

# Load Helper Functions
# Ensure the helper script is in the same folder as this Rmd
helper_file <- "Helper Functions for Computational Workflow.R"
if(file.exists(helper_file)) {
  source(helper_file)
} else {
  warning(paste("CRITICAL WARNING: Helper file '", helper_file, "' not found in working directory:", getwd()))
}

# --- OPTIMIZATION: Helper function for Stats Reporting in Step 6 ---
print_filtering_stats <- function(results, meta_cols = c("Plastic_type", "technique", "Source", "Polymer")) {
  
  df_final <- results$filtered_data
  df_orig  <- results$original_data
  
  # Get list of feature names only
  feats_final <- setdiff(names(df_final), meta_cols)
  feats_orig  <- setdiff(names(df_orig), meta_cols)
  
  # 1. Percentage of chemical features retained (%)
  pct_features <- (length(feats_final) / length(feats_orig)) * 100
  cat(paste0("1. Percentage of chemical features retained: ", round(pct_features, 2), "%\n"))
  
  # 2. Percentage of missing values after feature reduction (%)
  if(length(feats_final) > 0) {
    vals_final <- df_final[, feats_final, drop = FALSE]
    pct_missing <- (sum(is.na(vals_final)) / prod(dim(vals_final))) * 100
    cat(paste0("2. Percentage of missing values after feature reduction: ", round(pct_missing, 2), "%\n"))
  } else {
    cat("2. Percentage of missing values after feature reduction: 0% (No features left)\n")
    vals_final <- data.frame(0) # Placeholder for signal calc
  }
  
  # 3. Percentage of signal response after feature reduction (%)
  signal_final <- sum(vals_final, na.rm = TRUE)
  signal_orig  <- sum(df_orig[, feats_orig, drop = FALSE], na.rm = TRUE)
  pct_signal   <- (signal_final / signal_orig) * 100
  cat(paste0("3. Percentage of signal response after feature reduction: ", round(pct_signal, 2), "%\n"))
}
```

# Figure 2: HCA

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Data prep for HCA ---------------------
df <- run_rf_analysis_list_SB_and_ENV_train_02Aug2025$hplc_SB_and_ENV_train$final_imp_norm_dat %>%
  dplyr::mutate(Plastic_type = gsub("..", " ", Plastic_type, fixed = TRUE)) %>%
  dplyr::mutate(Plastic_type = gsub(".", " ", Plastic_type, fixed = TRUE)) %>%
  dplyr::mutate(Plastic_type_Source = paste0(Plastic_type, "-", Source))

hc_df <- df %>% 
  dplyr::select(-c("Subcategory", "Source", "Plastic_type", "Polymer", "technique")) %>% #, "Plastic_type_Source")) %>%
  dplyr::select(where(is.numeric)) %>%
  as.data.frame()

# --- Clean: remove bad rows & handle non-finite values ---
# 1) Replace non-finite with 0 (or choose a more domain-specific imputation if appropriate)
hc_df[!is.finite(as.matrix(hc_df))] <- 0

# 2) Drop rows that are all zero (Bray–Curtis can yield NaN for empty rows)
keep <- rowSums(hc_df, na.rm = TRUE) > 0
hc_df_clean <- hc_df[keep, , drop = FALSE] #%>% 
  # Select the features from best performance feature selection method
  # dplyr::select(run_rf_analysis_list_SB_and_ENV_train_02Aug2025$icp_SB_and_ENV_train$rfa_selected_feats)
labels_clean <- df$Plastic_type_Source[keep] # df$Plastic_type_Source[keep] 

# 3) If any negatives remain, Bray–Curtis is invalid -> fall back to a metric that permits negatives
use_bray <- !any(hc_df_clean < 0, na.rm = TRUE)

if (use_bray) {
  d <- vegan::vegdist(hc_df_clean, method = "manhattan")
  link <- "average"
} else {
  # If your features were centered/scaled, Euclidean or Manhattan are appropriate.
  # (Alternatively, if these are compositions, consider Aitchison with a small pseudocount.)
  d <- dist(hc_df_clean, method = "robust.aitchison")
  link <- "average"
  message("Detected negative values; using Euclidean instead of Bray–Curtis.")
}

# --- Guard: ensure distance has no NA/NaN/Inf ---
if (any(!is.finite(d))) {
  stop("Distance matrix still has non-finite entries after cleaning. Re-check inputs.")
}

# --- Cluster & plot ---
hca_samp <- hclust(d, method = link)

cat("cophenetic correlation coefficient (CCC) from a hierarchical clustering (HCA) is:", cor(d, stats::cophenetic(hca_samp)), "\n")

# tiff(filename = paste0("Figure SI_HCA_HPLC_SB_and_ENV_train_", format(Sys.time(), "%d-%m-%Y"),".tif"), width = 42*600, height = 15*600, res = 600, compression = "lzw")

par(cex = 1)
plot(hca_samp,
     labels = labels_clean,
     hang   = -1,
     main   = "")
# dev.off()
```

# Table S6: Correlation and covariance magnitude of features across 7 data combinations
```{r , echo=FALSE, warning = FALSE, message=FALSE}
## Before imp_norm   ===============================================
numeric_df <- gc_hplc_icp_combined_SB_ENV_shared_cols_ntr_clean %>% 
  dplyr::select(where(is.numeric))
cor_mat <- cor(numeric_df,
               use    = "pairwise.complete.obs",
               method = "pearson")
cov_mat <- cov(numeric_df,
               use    = "pairwise.complete.obs")
# extract off‐diagonals
cor_off <- cor_mat[upper.tri(cor_mat)]
cov_off <- cov_mat[upper.tri(cov_mat)]

# drop NAs before computing
cor_off <- cor_off[!is.na(cor_off)]
cov_off <- cov_off[!is.na(cov_off)]

# now summary statistics
avg_abs_cor  <- mean(abs(cor_off))
fro_cov_off  <- sqrt( sum(cov_off^2) )

round(avg_abs_cor,3)
round(fro_cov_off,3)

## After imp_norm  ===============================================
numeric_df <- run_rf_analysis_list_SB_and_ENV_train$gc_hplc_icp_SB_and_ENV_train$final_imp_norm_dat %>% 
  dplyr::select(where(is.numeric))

# Average absolute pairwise correlation
# 0 means no linear association anywhere.
# 1 means every pair is perfectly (positively or negatively) correlated.
# Pearson correlation matrix (pairwise complete observations)
cor_mat <- cor(numeric_df,
               use    = "pairwise.complete.obs",
               method = "pearson")
off_diag <- cor_mat[upper.tri(cor_mat)]
avg_abs_cor <- mean(abs(off_diag))

# Average absolute covariance
# 0 means every pair is (on average) uncorrelated—though note you could have positive and negative covariances that cancel each other out if you didn’t take the absolute value.
# Larger values mean more joint variability.
# Covariance matrix (pairwise complete observations)
cov_mat <- cov(numeric_df,
               use    = "pairwise.complete.obs")
# mean_abs_cor <- mean(abs(cor_mat[upper.tri(cor_mat)]))
# mean_abs_cor

cov_off <- cov_mat[upper.tri(cov_mat)]
fro_cov_off <- sqrt( sum(cov_off^2) )

round(avg_abs_cor,3)
round(fro_cov_off,3)
```

# Table S8. Significant compounds identified via pair-wise tests between plastic categories across seven data combinations
```{r echo=FALSE, warning=FALSE, message=FALSE}
list_pairwise_df <- list()
for (i in 1:length(run_rf_analysis_list_SB_and_ENV_train_02Aug2025)) {
  pw <- pairwise_significance_tests(run_rf_analysis_list_SB_and_ENV_train_02Aug2025[[i]]$final_imp_norm_dat, 
                                    group_col = "Plastic_type", start_col_index = 6)
  list_pairwise_df[[i]] <- pw$uncorrected %>%
    mutate(adjusted_p = p.adjust(pval, method = "bonferroni")) %>%
    filter(adjusted_p < 0.05) %>%
    arrange(adjusted_p) %>%
    dplyr::mutate(Datasets = names(run_rf_analysis_list_SB_and_ENV_train_02Aug2025)[i])
}

pw_df <- bind_rows(list_pairwise_df)
write_xlsx(pw_df, path = "Table S8_Significant compounds identified via pair-wise tests between plastic categories.xlsx")
```

# Figure 3: Bar plot Comparing Feature selection - MCC score - 7 data combinations
```{r echo=FALSE, warning=FALSE, message=FALSE}
df_list <- list()
dataset_names <- c("GC", "HPLC", "ICP", "GC-HPLC", "GC-ICP", "HPLC-ICP", "GC-HPLC-ICP")
for (i in 1:length(names(run_rf_analysis_list_SB_and_ENV_train_02Aug2025))) {
  names <- names(run_rf_analysis_list_SB_and_ENV_train_02Aug2025)[i]
  df_list[[i]] <- data.frame(time = round(c(dim(run_rf_analysis_list_SB_and_ENV_train_02Aug2025[[names]]$final_imp_norm_dat)[2],
                                          length(run_rf_analysis_list_SB_and_ENV_train_02Aug2025[[names]]$sig_selected_feats),
                                          length(run_rf_analysis_list_SB_and_ENV_train_02Aug2025[[names]]$rfa_selected_feats),
                                          length(run_rf_analysis_list_SB_and_ENV_train_02Aug2025[[names]]$top_importance_results$top_100$selected_features),
                                          length(run_rf_analysis_list_SB_and_ENV_train_02Aug2025[[names]]$top_importance_results$top_50$selected_features),
                                          length(run_rf_analysis_list_SB_and_ENV_train_02Aug2025[[names]]$top_importance_results$top_25$selected_features),
                                          length(run_rf_analysis_list_SB_and_ENV_train_02Aug2025[[names]]$top_importance_results$top_10$selected_features)), 3)) %>% dplyr::mutate(Dataset = dataset_names[i]) %>% dplyr::mutate(Selection = c("All features", "Pairwise", "RFA", "Top 100", "Top 50", "Top 25", "Top 10"))
}
View(bind_rows(df_list))
write_xlsx(bind_rows(df_list), path = paste0("Computational fingerprinting paper_feature selection comparison_train SB+ENV_04Aug2025_number selected features_", format(Sys.time(), "%d-%m-%Y"), ".xlsx"))

# PLot =============================
feature_selection <- read_xlsx(path = "C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/Computational fingerprinting paper_feature selection comparison_train SB+ENV_MCC score_time_feature number_04Aug2025.xlsx") %>%  
  dplyr::mutate(Dataset = factor(Dataset, levels = c("GC", "HPLC", "ICP", "GC-HPLC", "GC-ICP", "HPLC-ICP", "GC-HPLC-ICP")))

feature_selection_plot <- ggplot(data = feature_selection, aes(x = Dataset, y = MCC, fill = `Feature selection`)) + 
  geom_bar(position = position_dodge(width = 0.9), stat = "identity") +
  scale_fill_manual(values =  c('#8dd3c7','#ffffb3','#bebada','#fb8072','#80b1d3','#fdb462','#b3de69','#fccde5')) +  
  scale_y_continuous(breaks = seq(from = 0, to = 1, by =.1), expand = c(0,0), limits = c(0, 1)) +
  theme_bw(base_size = 20) +
  theme(axis.text.x = element_text(size =20), axis.text.y = element_text(size =20))
feature_selection_plot

ggsave(filename = paste0("Figure 3_Feature selection - MCC score - 7 data combinations_", format(Sys.time(), "%d-%m-%Y"), ".png"), 
       plot = feature_selection_plot, 
       device = "png",
       dpi = 300,        # Set DPI to 600
       width = 18,       # Width in inches (adjust as needed)
       height = 8,       # Height in inches (adjust as needed)
       units = "in")
```

# Figure SI#: Confusion matrices of training and testing on environmental plastics versus. training on store-bought plastics and testing on environmental plastics - 7 data combinations
```{r echo=FALSE, warning=FALSE, message=FALSE}
sb_train_only <- run_rf_analysis_list_SB_train$gc_USSB_train$conf_mat_sig
sb_and_env_train <- run_rf_analysis_list_SB_and_ENV_train_02Aug2025$gc_SB_and_ENV_train$conf_mat_rfa

ggsave(filename = paste0("Figure 4_confusion matrix_SB train only_", "GC", "_", format(Sys.time(), "%d-%m-%Y"), ".png"), 
       plot = sb_train_only, 
       device = "png",
       dpi = 300,        # Set DPI to 600
       width = 15,       # Width in inches (adjust as needed)
       height = 8,       # Height in inches (adjust as needed)
       units = "in")

ggsave(filename = paste0("Figure 4_confusion matrix_SB and ENV train", "GC", "_", format(Sys.time(), "%d-%m-%Y"), ".png"), 
       plot = sb_and_env_train, 
       device = "png",
       dpi = 300,        # Set DPI to 600
       width = 15,       # Width in inches (adjust as needed)
       height = 8,       # Height in inches (adjust as needed)
       units = "in")
```

# Figure SI #: Boxplot of MCC across feature selection methods
```{r , echo=FALSE, warning = FALSE, message=FALSE}
# 1. Define a function to extract metrics from a SINGLE fold
extract_metrics <- function(x, fold_idx) {
  
  # Helper to grab "top_X" stats easily
  get_top <- function(k) {
    sub <- x$top_importance_results[[paste0("top_", k)]]
    list(mcc = sub$eval_metrics_top$MCC_Multiclass,
         time = sub$time_top_100_50_25_10,
         n_feat = length(sub$selected_features))
  }
  
  # Create the data frame for this specific fold
  tibble::tribble(
    ~method,         ~MCC_Multiclass,       ~Computational_time,          ~Number_feat_selected,
    "all feats",     x$all_features_mcc,    x$time_all_features,          x$final_rf_all_feats$num.independent.variables,
    "sig feat",      x$mcc_sig,             x$time_sig,                   length(x$sig_selected_feats),
    "rfa feats",     x$mcc_rfa,             x$time_rfa,                   length(x$rfa_selected_feats),
    "top 100 feats", get_top(100)$mcc,      get_top(100)$time,            get_top(100)$n_feat,
    "top 50 feats",  get_top(50)$mcc,       get_top(50)$time,             get_top(50)$n_feat,
    "top 25 feats",  get_top(25)$mcc,       get_top(25)$time,             get_top(25)$n_feat,
    "top 10 feats",  get_top(10)$mcc,       get_top(10)$time,             get_top(10)$n_feat
  ) %>%
    dplyr::mutate(Folds = fold_idx, .before = 1)
}

# 2. Clean file path handling
# --- 2. Setup Directory and File List ---
target_dir <- file.path(getwd(), "ML results with 7 data combos using old feature reduction (12Dec2025)")

# Get all .rds files
rds_files <- list.files(target_dir, pattern = "\\.rds$", full.names = TRUE)

# --- 3. Iterate through files and create the List ---
summary_list <- lapply(rds_files, function(file_path) {
  
  message(paste("Processing:", basename(file_path)))
  
  # Read data
  data_in <- readRDS(file_path)
  
  # Get a clean name for the dataset from the filename (removes .rds extension)
  dataset_name <- tools::file_path_sans_ext(basename(file_path))
  
  # Calculate metrics for all folds in this file
  # using tryCatch to keep loop running even if one file is corrupted
  tryCatch({
    raw_results <- purrr::imap_dfr(data_in, extract_metrics)
    
    # Summarize logic
    stats_df <- raw_results %>%
      dplyr::group_by(method) %>%
      summarise(
        MCC_mean = round(mean(MCC_Multiclass, na.rm = TRUE), 2),
        MCC_sd = round(sd(MCC_Multiclass, na.rm = TRUE), 2),
        Computational_time_mean = round(mean(Computational_time, na.rm = TRUE), 2),
        Computational_time_sd = round(sd(Computational_time, na.rm = TRUE), 2),
        Number_feat_selected_mean = round(mean(Number_feat_selected, na.rm = TRUE), 2),
        Number_feat_selected_sd = round(sd(Number_feat_selected, na.rm = TRUE), 2)
      ) %>%
      dplyr::arrange(desc(MCC_mean)) %>%
      # Add the dataset name column so we know which file this came from
      dplyr::mutate(Dataset = dataset_name, .before = 1)
    
    return(stats_df)
    
  }, error = function(e) {
    message(paste("Error processing", dataset_name, ":", e$message))
    return(NULL) # Return NULL if error, will be removed later
  })
})

# --- 4. Final Binding ---
# Remove any NULL entries (in case of errors) and bind rows
final_summary_stats <- bind_rows(summary_list)

View(final_summary_stats)


# Exploring SI confusion matrices for the 7 data combinations
ggsave(filename = paste0("Figure S12_confusion matrix_SBandENVtrain_", "HPLC-ICP", "_", format(Sys.time(), "%d-%m-%Y"), ".png"), 
       plot = data_in$fold1$conf_mat_rfa, 
       device = "png",
       dpi = 300,        # Set DPI to 600
       width = 25,       # Width in inches (adjust as needed)
       height = 12,       # Height in inches (adjust as needed)
       units = "in")
rm(data_in); gc()

# boxplot of MCC by method
library(patchwork)
# 2. Define a function to generate the standardized boxplot
# This prevents copy-pasting the same complex theme code 3 times
create_boxplot <- function(data, y_col, y_label, show_x_axis = FALSE) {
  
  p <- ggplot(data, aes(x = method, y = .data[[y_col]])) +
    geom_boxplot() +
    
    # --- Your Custom Whisker Caps ---
    stat_summary(fun.data = function(x) {
      stats <- boxplot.stats(x)$stats
      data.frame(ymin = stats[1], ymax = stats[1])
    }, geom = "errorbar", width = 0.2, color = "black", linewidth = 1) +
    stat_summary(fun.data = function(x) {
      stats <- boxplot.stats(x)$stats
      data.frame(ymin = stats[5], ymax = stats[5])
    }, geom = "errorbar", width = 0.2, color = "black", linewidth = 1) +
    
    # --- Theme and Labels ---
    theme_bw(base_size = 22) + # Reduced slightly from 27 to fit 3 plots, adjust if needed
    labs(y = y_label, x = NULL) + 
    theme(
      axis.text.y = element_text(color = "black"),
      panel.grid.major.y = element_blank(),
      panel.grid.minor.y = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      axis.title.y = element_text(margin = ggplot2::margin(r = 10))
    )

  # Logic to hide X axis for the top two plots, but show it for the bottom one
  if (show_x_axis) {
    p <- p + 
      labs(x = "Feature selection methods") +
      theme(axis.text.x = element_text(color = "black", face = "bold", angle = 45, hjust = 1))
  } else {
    p <- p + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
  }
  
  return(p)
}

plot_data <- Result_gc_hplc_icp_SB_and_ENV_train
plot_data$method <- factor(plot_data$method, levels = unique(plot_data$method))
# 3. Create the 3 individual plots
# Note: scaling is handled automatically by ggplot for each specific variable
p_mcc  <- create_boxplot(plot_data, "MCC_Multiclass", "MCC score", show_x_axis = FALSE) + 
          scale_y_continuous(limits = c(0, 1)) # Explicit limit for MCC only

p_time <- create_boxplot(plot_data, "Computational_time", "Time (sec)", show_x_axis = FALSE)

p_feat <- create_boxplot(plot_data, "Number_feat_selected", "# Features", show_x_axis = TRUE)

# 4. Stack them using Patchwork
# The "/" symbol stacks plots vertically
combined_plot <- p_mcc / p_time / p_feat

ggsave(filename = paste0("Figure_SI_Stacked_Boxplots_", "run_rf_analysis_list_SB_and_ENV_train_test_modified80_gcandgcicp_16Dec2025", "_gc_icp_USSB_train", format(Sys.time(), "%d-%m-%Y"),".png"),
       plot = combined_plot,
       device = "png",
       dpi = 300,        # Set DPI to 600
       width = 12,       # Width in inches (adjust as needed)
       height = 12,       # Height in inches (adjust as needed)
       units = "in")
```

# Table S6. Results of evaluation metrics for data imputation and normalization across 7 data combinations using  store-bought and environmental samples as training dataset and test on environmental samples
```{r , echo=FALSE, warning = FALSE, message=FALSE}
target_dir <- file.path(getwd(), "TableS6_18Dec2025- Test RF and kNN imputation_old feature reduc 12Dec2025")

# Get all .rds files
rds_files <- list.files(target_dir, pattern = "\\.rds$", full.names = TRUE)
impxnorm_RFkNN_12Dec2025 <- list()
i <- 1
for (item in rds_files) {
  # Read data
  data_in <- readRDS(item)
  
  # Get a clean name for the dataset from the filename (removes .rds extension)
  dataset_name <- tools::file_path_sans_ext(basename(item))
  
  for (fold in data_in) {
    impxnorm_RFkNN_12Dec2025[[i]] <-  as.data.frame(fold$imp_norm_res_table) %>%
      dplyr::mutate(Fold = fold$fold_id) %>%
      dplyr::arrange(combined_rank) %>% 
      dplyr::select(-KS_p) %>% 
      dplyr::mutate(Dataset = dataset_name)
    i <- i + 1
  }
}

write_xlsx(bind_rows(impxnorm_RFkNN_12Dec2025), path = "TableS6_18Dec2025- Test RF and kNN imputation_old feature reduc 12Dec2025.xlsx")

# 3. Apply function to all data and combine
Result_icp_USSB_train <- bind_rows(Result_icp_USSB_train) %>% 
  dplyr::mutate(Dataset = "ICP",
                Training = "SB only")
rm(data_in); gc()

all_imp_norm_res_12Dec2025 <- rbind(Result_gc_hplc_icp_SB_and_ENV_train,
                                    Result_gc_hplc_icp_USSB_train,
                                    Result_gc_hplc_SB_and_ENV_train,
                                    Result_gc_hplc_USSB_train,
                                    Result_gc_icp_SB_and_ENV_train,
                                    Result_gc_icp_USSB_train,
                                    Result_gc_SB_and_ENV_train,
                                    Result_gc_USSB_train,
                                    Result_hplc_icp_SB_and_ENV_train,
                                    Result_hplc_icp_USSB_train,
                                    Result_hplc_SB_and_ENV_train,
                                    Result_hplc_USSB_train,
                                    Result_icp_SB_and_ENV_train,
                                    Result_icp_USSB_train)

base::saveRDS(all_imp_norm_res_12Dec2025, file = paste0(getwd(), "/ML results with 7 data combos using old feature reduction (12Dec2025)/", "Table S6_all_imp_norm_res_12Dec2025.rds"))
```

# Table S7. Summary of significant compounds identified via pair-wise tests between plastic categories from k-fold cross validation across seven data combinations
```{r , echo=FALSE, warning = FALSE, message=FALSE}
sig_feat_list_SB_ENV <- list()
k <- 1
for (i in seq_along(run_rf_analysis_list_SB_and_ENV_train_pairwise_17dec)) {
  current_data <- run_rf_analysis_list_SB_and_ENV_train_pairwise_17dec[[i]]
  for (fold_idx in seq_along(current_data)) {
    fold_content <- current_data[[fold_idx]]
    if (!is.null(fold_content$pvalue_corrected_df)) {
      sig_feat_list_SB_ENV[[k]] <- fold_content$pvalue_corrected_df %>% 
        dplyr::mutate(Dataset = names(run_rf_analysis_list_SB_and_ENV_train_pairwise_17dec)[[i]])
      
      # Increment the output counter
      k <- k + 1
    }
  }
}

# 2. Prepare the reference tables (Calculate means)
# Calculate means for GC data
gc_means <- gc %>%
  group_by(Feature) %>%
  summarise(
    mean_mz_gc = mean(m.z, na.rm = TRUE),
    mean_rt_gc = mean(RT, na.rm = TRUE)
  )

# Calculate means for HPLC data
hplc_means <- hplc %>%
  group_by(Feature) %>%
  summarise(
    mean_mz_hplc = mean(m.z, na.rm = TRUE),
    mean_rt_hplc = mean(RT, na.rm = TRUE)
  )

sig_feat_SB_ENVtrain <- bind_rows(sig_feat_list_SB_ENV) %>% 
  dplyr::select(-pval, -adjusted_pvalue) %>%
  # 2. Remove duplicates based on Dataset, Feature, and Categories
  # This ensures that for every Dataset, you only have unique Feature/Category combos
  dplyr::distinct(Dataset, Feature, Category_1, Category_2, .keep_all = TRUE) %>%
  # Join with the GC means
  left_join(gc_means, by = "Feature") %>%
  
  # Join with the HPLC means
  left_join(hplc_means, by = "Feature") %>%
  
  # Create the new columns based on the Feature name
  dplyr::mutate(
    `mean(m.z)` = case_when(
      str_detect(Feature, "ATDGCMS") ~ mean_mz_gc,
      str_detect(Feature, "HPLCTOFMS") ~ mean_mz_hplc,
      TRUE ~ NA_real_ # Sets to NaN/NA if neither is found
    ),
    `mean(RT)` = case_when(
      str_detect(Feature, "ATDGCMS") ~ mean_rt_gc,
      str_detect(Feature, "HPLCTOFMS") ~ mean_rt_hplc,
      TRUE ~ NA_real_
    )
  ) %>%
  
  # Remove the temporary helper columns
  dplyr::select(-mean_mz_gc, -mean_rt_gc, -mean_mz_hplc, -mean_rt_hplc)


write_xlsx(rbind(sig_feat_SB_ENVtrain, sig_feat_SBtrain), path = "Table S7_significant features SB train and SB_ENV train_old feature reduction 12Dec2025.xlsx")
```

# Table 2: Summary statistics of ML results across 7 data combinations from 12Dec2025
```{r , echo=FALSE, warning = FALSE, message=FALSE}
file_path <- file.path(getwd(),"ML results with 7 data combos using old feature reduction (12Dec2025)", "Result_icp_USSB_train.rds")
data_in <- readRDS(file_path)

Result_icp_USSB_train <- list()
i <- 1
for (fold in data_in) {
  Result_icp_USSB_train[[i]] <- as.data.frame(fold$imp_norm_res_table) %>%
    dplyr::mutate(Fold = fold$fold_id) %>%
    dplyr::arrange(combined_rank) %>% dplyr::select(-KS_p)
  i <- i + 1
}

# 3. Apply function to all data and combine
Result_icp_USSB_train <- bind_rows(Result_icp_USSB_train) %>% 
  dplyr::mutate(Dataset = "ICP",
                Training = "SB only")
rm(data_in); gc()
```
