---
title: "Microplastic fingerprinting"
output: 
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/huyng/OneDrive - Toronto Metropolitan University/Microplastic/Microplastic-Fingerprinting")
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(out.width='750px', dpi=200)
```

## Documentation

This repo is accompanying the publication: "Computational fingerprinting workflow for environmental source tracking of microplastic based on representative additives"

Demo of each ML algorithms is shown below.

#######################################################################################################
################### DATA PROCESSING SECTION ##################################
#######################################################################################################

You can include R code in the document as follows:

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# Loading Packages --------------------------------------------------------
library(ggplot2)
library(vegan)
library(readxl)
library(tidyverse)
library(dplyr)
library(data.table)
library(writexl)
library(tidyr)
library(grid)
library(gridExtra)
library(stats)
library(FactoMineR)
library(factoextra)
library(ggforce)


# Functions -------------------------------------------------------------------------------------------------------
'%notin%' <- Negate('%in%')

# Grouping compounds based on RT1, RT2, and Ion1 - Version 1
grouping_comp <- function(data, rt_cutoff, rtthres1, rtthres2, mzthres_ppm, type) {
  
  # create empty list, each sub-list is a compound group with following criteria:
  # rtthres: RT threshold window
  # mzthres: mz threshold window
  dat <- data[ , , drop = FALSE]  # Shallow copy to avoid modification of original data
  
  # Initialize the compound column filled with NA values
  dat$Feature <- NA
  i <- 1
  j <- 1 # Counter for HPLC data only
  
  # Checkpoint for 'type' parameter
  if (!(type %in% c("ATDGCMS", "HPLCTOFMS"))) {
    stop("Invalid type specified. Must be either 'ATDGCMS' or 'HPLCTOFMS'")
  }

  # Loop through each row to group compounds
  if (type %in% "HPLCTOFMS") {
      # Split HPLC data into RT <= rt_cutoff min and RT > rt_cutoff min
      dat1 <- dat %>% filter(RT <= rt_cutoff) %>% arrange(Name, m.z, RT)
      dat2 <- dat %>% filter(RT > rt_cutoff) %>% arrange(Name, m.z, RT)
      
      for (row in 1:nrow(dat1)) {
        # filter data by index, ALWAYS DO THIS INSTEAD OF CREATE SUBSET DATAFRAME
        comp1 <- dat1[row,]$Name
        rt <- base::as.numeric(dat1[row, "RT"])
        mz <- base::as.numeric(dat1[row, "m.z"])
        
        idx1 <- which(
          dat1$Name == comp1 &
          dat1$RT <= (rt + rtthres1) & dat1$RT >= (rt - rtthres1) &
            dat1$m.z <= (mz + (mz*mzthres_ppm/1000000)) & dat1$m.z >= (mz - (mz*mzthres_ppm/1000000)) &
            is.na(dat1$Feature)
        )
        
        if (length(idx1) > 0) {
          dat1[idx1, "Feature"] <- paste0("Compound_RT1_", i, ".", type)
          i <- i + 1
        }  
      }
      
      for (row in 1:nrow(dat2)) {
        # filter data by index, ALWAYS DO THIS INSTEAD OF CREATE SUBSET DATAFRAME
        comp2 <- dat2[row,]$Name
        rt <- base::as.numeric(dat2[row, "RT"])
        mz <- base::as.numeric(dat2[row, "m.z"])
        
        idx2 <- which(
          dat2$Name == comp2 &
          dat2$RT <= (rt + rtthres2) & dat2$RT >= (rt - rtthres2) &
            dat2$m.z <= (mz + (mz*mzthres_ppm/1000000)) & dat2$m.z >= (mz - (mz*mzthres_ppm/1000000)) &
            is.na(dat2$Feature)
        )
        
        if (length(idx2) > 0) {
          dat2[idx2, "Feature"] <- paste0("Compound_RT2_", j, ".", type)
          j <- j + 1
        }  
      }
      
      dat <- rbind(dat1, dat2)
      
  } else {
    for (row in 1:nrow(dat)) {
      
      # filter data by index, ALWAYS DO THIS INSTEAD OF CREATE SUBSET DATAFRAME
      rt <- base::as.numeric(dat[row, "RT"])
      mz <- base::as.numeric(dat[row, "m.z"])
      
      idx <- which(
        dat$RT <= (rt + rtthres1) & dat$RT >= (rt - rtthres1) &
          dat$m.z <= (mz + mzthres) & dat$m.z >= (mz - mzthres) &
          is.na(dat$Feature)
      )
      
      if (length(idx) > 0) {
        dat[idx, "Feature"] <- paste0("Compound_", i, ".", type)
        i <- i + 1
      }  
    }
  }
  
  return(dat)
}
```

### STEP 1: Data import
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# Reading in raw data as Excel files
data <- readxl::read_xlsx("C:/Users/huyng/OneDrive - Toronto Metropolitan University/Microplastic/Microplastic-Fingerprinting/Eric PET 9th September 2024/Huy_PET Leachate Raw df_09-09-24.xlsx") %>% 
  # Select which column you want to keep or remove from the raw data input
  dplyr::select(-c(CAS, Score, Hits, `Diff (Tgt, ppm)`, `Flags (Tgt)`))

# Convert column name of m.z to "m/z"
colnames(data)[8] <- "m.z"
# Optional step to better describe your sample
data$Sample_name <- paste0(data$Product, "_", data$Day)

data <- data %>% 
  # select(-c(Product)) %>% 
  relocate(Sample_name, .after = 2)

# Check for each row of the input data, if the values of all blank columns are missing then we delete that row
rows_to_remove <- rowSums(is.na(data[, 12:26])) == length(12:26)
data <- data[!rows_to_remove, ]

# (optional step) If you saw that the column Acetaminophen-d4 RT in your data does not have numerical type then your need to convert it to numeric.
# data$`Acetaminophen-d4 RT` <- as.numeric(unlist(data$`Acetaminophen-d4 RT`))
```

### Quality assurance prior to STEP 2: Examine variation in RT of 4 benchmark compounds in HPLC across all samples with histogram distribution of RT of 4 benchmark compounds 

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
bm_df <- data %>%
  group_by(Sample_name, Replicate, Day) %>%
  summarise(Acetaminophen_RT = mean(`Acetaminophen-d4 RT`),
            dTEP_RT = mean(`dTEP RT`),
            dTPrP_RT = mean(`dTPrP RT`),
            PPD_RT = mean(`13C-6PPD RT`),
            dTPP_RT = mean(`dTPP RT`))

# Plot scatter of RT distribution
par(mfrow = c(3, 2))

hist(bm_df$Acetaminophen_RT[bm_df$Acetaminophen_RT < 6], main = "Histogram of Acetaminophen_RT", xlab = "Values", ylab = "Frequency")
hist(bm_df$dTEP_RT, main = "Histogram of dTEP_RT", xlab = "Values", ylab = "Frequency")
hist(bm_df$dTPrP_RT, main = "Histogram of dTPrP_RT", xlab = "Values", ylab = "Frequency")
hist(bm_df$PPD_RT, main = "Histogram of 6PPD_RT", xlab = "Values", ylab = "Frequency")
hist(bm_df$dTPP_RT, main = "Histogram of dTPP_RT", xlab = "Values", ylab = "Frequency")
```

### STEP 2: Grouping compounds based on RT1, RT2, Ion1 threshold --------------------------
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# HPLCTOFMS
# Apply different alignment to 2 parts of chromatogram via rtthres1 (RT <= 6 min) and rtthres2 (RT > 6 min)
combined_df2_grouped <- grouping_comp(data,
                                      rt_cutoff = 6,
                                      rtthres1 = 0.15,
                                      rtthres2 = 0.1,
                                      mzthres_ppm = 10,
                                      type = "HPLCTOFMS")
```

### Step 3: Calculate Relative Peak Area (RPA) using closest eluting benchmarks

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# Feature cleaning layer 1: remove all unique aligned feature that only appears once because they are more likely to be random background chemical noise or artifact from instrumental analysis
non_unique <- c()
for (comp in unique(combined_df2_grouped$Feature)) {
  idx <- which(grepl(comp, combined_df2_grouped$Feature, fixed = TRUE))
  if (length(idx) < 2) {
    next
  } else {
    non_unique <- c(non_unique, idx)
  }
}

non_unique_df1 <- combined_df2_grouped[non_unique,] %>% 
  relocate(Feature, .before = 1)

# Feature cleaning layer 2 - (OPTIONAL step, can skip it if you don't want to apply this layer of feature reduction): For each aligned feature, if the feature occurs in 2 replicates of the same sample, keep those occurences and remove the rest 
row_to_keep <- c()
for (comp in unique(non_unique_df1$Feature)) {
  idx <- which(grepl(comp, non_unique_df1$Feature, fixed = TRUE))
  count <- c()
  for (i in 1:length(unique(non_unique_df1[idx,]$Sample_name))) {
    count <- c(count, length(unique(non_unique_df1[idx, ][which(non_unique_df1[idx,]$Sample_name == unique(non_unique_df1[idx,]$Sample_name)[i]), ]$Replicate)))
  }
  if (any(count > 1)) {
    samp_to_keep <- unique(non_unique_df1[idx,]$Sample_name)[which(count > 1)]
    row_to_keep <- c(row_to_keep, idx[which(non_unique_df1[idx,]$Sample_name %in% samp_to_keep)])
  }
}

non_unique_df2 <- non_unique_df1[row_to_keep,]

# Calculate. Relative Peak Area (RPA)
tempdf_list <- list()
i <- 1
# calculate based on the closest eluting benchmark
for (comp in unique(non_unique_df2$Feature)) {
  idx <- which(non_unique_df2$Feature == comp)
  tempdf <- non_unique_df2[idx,]
  
  # Calculate the difference in retention time between each unique feature with the benchmark compounds
  min_Acetaminophen <- abs(mean(non_unique_df2[idx,]$RT) - mean(non_unique_df2[idx,]$`Acetaminophen-d4 RT`))
  min_dTEP <- abs(mean(non_unique_df2[idx,]$RT) - mean(non_unique_df2[idx,]$`dTEP RT`))
  min_dTPrP <- abs(mean(non_unique_df2[idx,]$RT) - mean(non_unique_df2[idx,]$`dTPrP RT`))
  min_6PPD <- abs(mean(non_unique_df2[idx,]$RT) - mean(non_unique_df2[idx,]$`13C-6PPD RT`))
  min_dTPP <- abs(mean(non_unique_df2[idx,]$RT) - mean(non_unique_df2[idx,]$`dTPP RT`))
  all_min <- c(min_Acetaminophen, min_dTEP, min_dTPrP, min_6PPD, min_dTPP)
  
  # conditional statement to figure out which benchmark is the closest eluting benchmark and then calculate RPA by dividing Peak Area (raw) to the Area of the closest eluting benchmark compound
  if (which(all_min == min(all_min, na.rm = TRUE)) == 1) {
    tempdf$RPA <- tempdf$Area / tempdf$`Acetaminophen-d4 Area`
    tempdf_list[[i]] <- tempdf
    i <- i + 1
  } else if (which(all_min == min(all_min, na.rm = TRUE)) == 2) {
    tempdf$RPA <- tempdf$Area / tempdf$`dTEP Area`
    tempdf_list[[i]] <- tempdf
    i <- i + 1
  } else if (which(all_min == min(all_min, na.rm = TRUE)) == 3) {
    tempdf$RPA <- tempdf$Area / tempdf$`dTPrP Area`
    tempdf_list[[i]] <- tempdf
    i <- i + 1
  } else if (which(all_min == min(all_min, na.rm = TRUE)) == 4) {
    tempdf$RPA <- tempdf$Area / tempdf$`13C-6PPD Area`
    tempdf_list[[i]] <- tempdf
    i <- i + 1
  } else {
    tempdf$RPA <- tempdf$Area / tempdf$`dTPP Area`
    tempdf_list[[i]] <- tempdf
    i <- i + 1
  }
}

rpa_df <- bind_rows(tempdf_list)
```

### Step 4: (OPTIONAL step, depending on your experimental design, use it if you leach over multiple days and have different concentration of the leachates between each sampling dates) - Calculate concentration corrected RPA by Day

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
rpa_df$corrected_rpa <- NA

for (row in 1:nrow(rpa_df)) {
  if (rpa_df[row,]$Day == 0) {
    rpa_df[row,]$corrected_rpa <- rpa_df[row,]$RPA / rpa_df[row,]$`Conc Factor Day 0`
  } else if (rpa_df[row,]$Day == 2) {
    rpa_df[row,]$corrected_rpa <- rpa_df[row,]$RPA / rpa_df[row,]$`Conc Factor Day 2`
  } else if (rpa_df[row,]$Day == 15) {
    rpa_df[row,]$corrected_rpa <- rpa_df[row,]$RPA / rpa_df[row,]$`Conc Factor Day 15`
  } else {
    rpa_df[row,]$corrected_rpa <- rpa_df[row,]$RPA / rpa_df[row,]$`Conc Factor Day 34`
  }
}
```

### Step 5: Calculate the mean and standard deviation of RPA across replicates and blanks

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# If you ran step 4 
if (!is.na(rpa_df$corrected_rpa)) {
  avg_df <- rpa_df %>%
    group_by(Feature, Sample_name, Day, Batch, Product) %>%
    summarise(mean_corrected_rpa = mean(corrected_rpa),
              sd_corrected_rpa = sd(corrected_rpa),
              mean_RT = mean(RT)
    )
  
} else { # If you did not run step 4
  avg_df <- rpa_df %>%
    group_by(Feature, Sample_name, Day, Batch, Product) %>%
    summarise(mean_rpa = mean(RPA),
              sd_rpa = sd(RPA),
              mean_RT = mean(RT)
    )
}
```


### Step 6: Blank Subtraction per batch

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# If you ran step 4 - run the code chunk below --------
df_list <- list()
i <- 1
for (comp in unique(avg_df$Feature)) {
  tempdf1 <- avg_df[which(avg_df$Feature == comp),]
  # for (day in unique(tempdf1$Day)) { # Comment out if you don't have multiple leaching days
  # tempdf2 <- tempdf1[which(tempdf1$Day == day),] { # Comment out if you don't have multiple leaching days
  blank_mean_corrected_rpa <- tempdf2[which(tempdf2$Product == "Blank"),]$mean_corrected_rpa
  blank_sd_corrected_rpa <- tempdf2[which(tempdf2$Product == "Blank"),]$sd_corrected_rpa
  if (length(blank_mean_corrected_rpa) != 0) {
    tempdf2$blank_substracted <- tempdf2$mean_corrected_rpa - blank_mean_corrected_rpa
    negative_idx <- which(tempdf2$blank_substracted < 0)
    positive_idx <- which(tempdf2$blank_substracted > 0 & tempdf2$blank_substracted > 3*blank_sd_corrected_rpa) # Quick head-ups: you can potentially lost many features at this step because values after blank subtraction are < 3*blank_sd_corrected_rpa
    
    tempdf2 <- tempdf2[positive_idx,]
    tempdf2 <- tempdf2[-negative_idx,] # remove obs. with negative blank_substracted values
    df_list[[i]] <- tempdf2
    i <- i + 1
  } else {
    tempdf2$blank_substracted <- tempdf2$mean_corrected_rpa
    df_list[[i]] <- tempdf2
    i <- i + 1
  }
}

# If you did not run step 4 - run the code chunk below --------
df_list <- list()
i <- 1
for (comp in unique(avg_df$Feature)) {
  tempdf1 <- avg_df[which(avg_df$Feature == comp),]
  # for (day in unique(tempdf1$Day)) { # Comment out if you don't have multiple leaching days
  # tempdf2 <- tempdf1[which(tempdf1$Day == day),] { # Comment out if you don't have multiple leaching days
  blank_mean_rpa <- tempdf2[which(tempdf2$Product == "Blank"),]$mean_rpa
  blank_sd_rpa <- tempdf2[which(tempdf2$Product == "Blank"),]$sd_rpa
  if (length(blank_mean_rpa) != 0) {
    tempdf2$blank_substracted <- tempdf2$mean_rpa - blank_mean_rpa
    negative_idx <- which(tempdf2$blank_substracted < 0)
    positive_idx <- which(tempdf2$blank_substracted > 0 & tempdf2$blank_substracted > 3*blank_sd_rpa) # Quick head-ups: you can potentially lost many features at this step because values after blank subtraction are < 3*blank_sd_corrected_rpa
    
    tempdf2 <- tempdf2[positive_idx,]
    tempdf2 <- tempdf2[-negative_idx,] # remove obs. with negative blank_substracted values
    df_list[[i]] <- tempdf2
    i <- i + 1
  } else {
    tempdf2$blank_substracted <- tempdf2$mean_rpa
    df_list[[i]] <- tempdf2
    i <- i + 1
  }
}

bs_df <- bind_rows(df_list)

bs_df <- rbind(bs_df, avg_df %>% filter(str_detect(Product, pattern = "Blank")))

name <- c()
for (row in 1:nrow(bs_df)) {
  name <- c(name, unique(rpa_df[which(str_detect(rpa_df$Feature, pattern = bs_df[row,]$Feature)),]$Name))
}

bs_df$Name <- name
bs_df <- bs_df %>% relocate(Name, .after = 1)

# Exporting your final dataframe into Excel files for further usage
writexl::write_xlsx(bs_df, path = paste0("blank_substracted_df_", format(Sys.Date(), "%d-%b-%Y"), ".xlsx"))
writexl::write_xlsx(rpa_df, path = paste0("rpa_df_", format(Sys.Date(), "%d-%b-%Y"), ".xlsx"))
```

#######################################################################################################
################### EXPLORATORY DATA ANALYSES AND CLUSTERING SECTION ##################################
#######################################################################################################

# Data import (usually after data processing step above, some user will change the name of column in the exported rpa_df above and may add new column so I keep this as a flexible import of their own dataframe for data analysis and clustering step)
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
ori_data <- readxl::read_xlsx("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Microplastic-Fingerprinting-scripts and data/Eric PMT Weathering _Suspect Screening_01July2024/Updated_Leachates_df.xlsx")
```

### Step 7 (option 1): Compound as predictor variables - output will be df where rows are samples and columns are Features
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
compound_df <- ori_data %>%
  dplyr::select(Product, Feature, Day, Replicate, Batch, corrected_rpa) %>% 
  dplyr::group_by(Product, Feature, Day, Replicate, Batch) %>%
  dplyr::summarise(across(corrected_rpa, base::sum)) %>% 
  tidyr::pivot_wider(names_from =  Feature,
                     values_from = corrected_rpa) %>%
  relocate(Product, Feature, Day, Replicate, Batch, .after = 1)
```

### Step 7 (option 2): Polymer as predictor variables || only use when you have Polymer identified in your dataset, otherwise skip it
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
polymer_df <- ori_data %>%
  dplyr::select(Product, Feature, Day, Replicate, Batch, corrected_rpa) %>% 
  dplyr::group_by(Product, Feature, Day, Replicate, Batch) %>%
  dplyr::summarise(across(corrected_rpa, base::sum)) %>% 
  tidyr::pivot_wider(names_from =  Polymer,
                     values_from = corrected_rpa) %>%
  relocate(Product, Feature, Day, Replicate, Batch, .after = 1)
```

### Step 7 (option 3): Polymer+Compound as predictor variables || only use when you have Polymer identified in your dataset, otherwise skip it
```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
polymer_compound_df <- ori_data %>%
  dplyr::mutate(polymer_feature =  paste0(Polymer, "_", Feature)) %>%
  dplyr::select(Product, Feature, Day, Replicate, Batch, corrected_rpa) %>% 
  dplyr::group_by(Product, Feature, Day, Replicate, Batch) %>%
  dplyr::summarise(across(corrected_rpa, base::sum)) %>% 
  tidyr::pivot_wider(names_from =  polymer_feature,
                     values_from = corrected_rpa) %>%
  relocate(Product, Feature, Day, Replicate, Batch, .after = 1)
```

### Step 8: Filling missing values and percentage normalization

```{r , echo=FALSE, warning = FALSE, message=FALSE}
## Choose which dataframe output from one of the three options above at Step 7: Compound || Polymer || Polymer+Compound as predictor variables
df_0.001 <- copy(compound_df) 

# Missing value imputation with 0.001
for (r in 1:nrow(df_0.001)) {
  df_0.001[r, which(base::is.na(df_0.001[r,]))] <- 0.001
}

input <- df_0.001

# Percentage Normalization
percentage_normalized_df <- as.data.frame(t(apply(input[, 6:ncol(input)],
                                       MARGIN = 1, 
                                       function(row) {row/sum(row, na.rm = TRUE)})))

percentage_normalized_df <- percentage_normalized_df %>% 
  dplyr::mutate(Day = input$Day) %>%
  dplyr::mutate(Replicate = input$Replicate) %>%
  dplyr::mutate(Batch = input$Batch) %>%
  dplyr::mutate(Product = input$Product) %>%
  relocate(Produc, Day, Replicate, Batch, .before = 1)

# remove artifacts from Subset compound that only appear in 1 Source/Polymer/Product_Name
percentage_normalized_df_cleaned <- na.omit(percentage_normalized_df)
```

### Step 9: Pairwise test

**NOTE**: For each compound (aka. column) -> Need to subset any categories that have less than 3 samples into a different df for later inspection, because after filling in missing data, if two categories missing data for that compound entirely, they will be filled with exactly same values -> p-values will be NaN and have cannot compute exact p-value with ties.

```{r class.source = 'fold-hide', echo = FALSE, message = FALSE, warning = FALSE}
# Initiate empty df for storing wilcoxon test results
input <- percentage_normalized_df_cleaned

pairwise_df <- data.frame(Feature=character(), comparison_pair=character(), pval=integer())

# Subset of compounds with less than 3 data points per category
col_contain_category_with_less_than_three_obs <- c()

for (col in 6:ncol(input)) {
  
  # looping through the combinations of product categories 
  for (c in 1:ncol(utils::combn(unique(input$Product), 2))) {
    # extract the combinations of product category pairs
    p_1 <- utils::combn(unique(input$Product), 2)[1,c]
    p_2 <- utils::combn(unique(input$Product), 2)[2,c]
    
    # If either one of the comparing category have less than 3 data points => move on to the next comparison pair
    if (sum(!is.na(as.numeric(unlist(input[which(input$Product == p_1), col])))) < 3 |
        sum(!is.na(as.numeric(unlist(input[which(input$Product == p_2), col])))) < 3) {
      # print(col)
      col_contain_category_with_less_than_three_obs <- c(col_contain_category_with_less_than_three_obs, col)
      next
    } else {
      # if (input[which(input$Product_Name == p_1), col] | input[which(input$Product_Name == p_2), col])
    # print(col)
    #   print(paste0(p_1, "_", p_2))
      # calculating the p-value between each product category pair 
      pval <- stats::wilcox.test(input[which(input$Product == p_1), col], 
                                 input[which(input$Product == p_2), col])$p.value
      # assigning row information
      pairwise_df[nrow(pairwise_df) + 1,] <- c(colnames(input)[col], 
                                               paste0(p_1, " & ", p_2), 
                                               pval)
    }
  }
}

sig_comp <- list()
j <- 1
pvaluecorrect <- c("holm", "hochberg", "hommel", "BH", "BY")
for (m in pvaluecorrect) {
  pairwise_df$adjusted_pvalue <- stats::p.adjust(pairwise_df$pval, method = m)
  print(pairwise_df %>%
          filter(., adjusted_pvalue < 0.05) %>%
          arrange(adjusted_pvalue))
  
  sig_comp[[j]] <- pairwise_df %>%
          filter(., adjusted_pvalue < 0.05) %>%
          arrange(adjusted_pvalue)
  j <- j + 1
}

not_corrected_pval <- pairwise_df %>%
  filter(., pval < 0.05) %>% 
  arrange(pval)



rt <- c()
mz <- c()
comp <- c()
for (row in 1:nrow(sig_comp[[1]])) {
  # print(row)
  rt <- c(rt, mean(ori_data[which(ori_data$Feature == sig_comp[[1]][row,]$Feature),]$Mean_RT))
  mz <- c(mz, mean(ori_data[which(ori_data$Feature == sig_comp[[1]][row,]$Feature),]$`m/z`))
  comp <- c(comp, unique(ori_data[which(ori_data$Feature == sig_comp[[1]][row,]$Feature),]$Name))
  
}

sig_comp[[1]]$rt <- rt
sig_comp[[1]]$mz <- mz
sig_comp[[1]]$Compound <- comp

View(sig_comp[[1]])
```


## Step 10: Clustering Analyses

### Step 10.1: PCA
```{r , echo=FALSE, warning = FALSE, message=FALSE}
pca_dat <- percentage_normalized_df_cleaned 

res.pca <- FactoMineR::PCA(
  pca_dat[, -c(1,2,3,4,5)],
  scale.unit = FALSE,
  graph = FALSE)

# Scree plot
fviz_screeplot(res.pca, ncp=10)

# Biplot
factoextra::fviz_pca_biplot(res.pca,  
                            geom = c("point", "text"),
                            label = "none",
                            # invisible = "var", 
                            repel = TRUE,
                            labelsize = 5, 
                            habillage = factor(pca_dat$Polymer),
                            addEllipses = TRUE,
                            ellipse.level = 0.95,
                            ggtheme = ggplot2::theme_minimal(base_size = 15),
                            title = "POLYMER",
                            # xlim = c(-200, 1000),
                            # ylim = c(-100, 200)
                            ) + 
  guides(fill=guide_legend(ncol=5)) +
  # if has error "Too few points to calculate an ellipse"
  # ggforce::geom_mark_ellipse(aes(fill = Groups,
  #                                color = Groups),
  #                            label.buffer = unit(40, 'mm')) +
  theme(legend.position = 'bottom',
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank()
        )
```


###  Step 10.2: t-SNE
```{r, echo=FALSE, warning = FALSE, message=FALSE}
# REFERENCES VISUALIZATION: 
# https://plotly.com/r/t-sne-and-umap-projections/
# https://distill.pub/2016/misread-tsne/

library(tsne)
library(plotly)

clus_dat <- percentage_normalized_df_cleaned 
features <- clus_dat[, -c(1,2,3,4,5)]

tsne <- tsne(features,
             initial_dims = 3, 
             k = 3, 
             perplexity = 20, # Hyperparameter: perplexity (optimal number of neighbors) < number of samples
             max_iter = 1000
             )

pdb <- cbind(data.frame(tsne), clus_dat$Polymer)
options(warn = -1)
tsne_plot <- plot_ly(data = pdb ,x =  ~X1, y = ~X2, z = ~X3, 
               color = ~clus_dat$Polymer) %>% 
  add_markers(size = 8) %>%
  layout( 
    xaxis = list(
      zerolinecolor = "#ffff",
      zerolinewidth = 2,
      gridcolor='#ffff'), 
    yaxis = list(
      zerolinecolor = "#ffff",
      zerolinewidth = 2,
      gridcolor='#ffff'),
    scene =list(bgcolor = "#e5ecf6"))

tsne_plot
```

###  Step 10.3: UMAP clustering 

```{r, echo=FALSE, warning = FALSE, message=FALSE}
library(umap)
library(plotly)
clus_dat <- percentage_normalized_df_cleaned 
features <- clus_dat[, -c(1,2,3,4,5)]

umap <- umap(features, n_components = 3,
             method = 'naive', metric= "manhattan",
             alpha = 0.0001, gamma = 0.0001)

layout <- cbind(data.frame(umap[["layout"]]), clus_dat$Source)
umap_plot <- plotly::plot_ly(layout, x = ~X1, y = ~X2, z = ~X3, 
                color = ~clus_dat$Source) %>% 
  add_markers() %>%
  layout(scene = list(xaxis = list(title = 'x-axis'), 
                      yaxis = list(title = 'y-axis'), 
                      zaxis = list(title = 'z-axis'))) 
umap_plot
```

###  Step 10.4: HCA

```{r echo=FALSE, warning=FALSE, message=FALSE}
hc_df <- percentage_normalized_df_cleaned[, -c(1,2,3,4,5)]

## Dissimilarity Indices Calculated by vegan::vegdist()
hca_samp <- stats::hclust(vegan::vegdist(hc_df,
                                         method = "robust.aitchison")) # Since our data is continous -> canberra // manhattan // aitchison // robust.aitchison

par(cex = 0.75)
plot(hca_samp,
     labels = percentage_normalized_df_cleaned$Product,
     hang = -1,
     main = "")
```

###  Step 10.5: NMDS 
**References:** 
https://eddatascienceees.github.io/tutorial-rayrr13/
https://jonlefcheck.net/2012/10/24/nmds-tutorial-in-r/

```{r echo=FALSE, warning=FALSE, message=FALSE}
data <- percentage_normalized_df_cleaned 

# Calculate the dissimilarity matrix using Bray-Curtis distance
dissimilarity_matrix <- vegdist(data %>% dplyr::select(6:ncol(data)), method = "bray", na.rm = TRUE)

# Perform NMDS
nmds_result <- metaMDS(dissimilarity_matrix, k = 2, trymax = 100)

# Extract NMDS scores
nmds_scores <- as.data.frame(vegan::scores(nmds_result))

# Add metadata (Product_Name, Source, Day) to the NMDS scores
nmds_scores <- cbind(nmds_scores, data %>% dplyr::select(Generalized_Product_name, Source
                                                  # Product_Name, Day
                                                  ))

# Define a vector of shapes
shapes <- c(0:16) #, 21:24, 25:32, 33:40, 41:48, 49:56, 57:63)

# Ensure the shapes vector has unique shapes
# shapes <- shapes[1:16]

# Generate a discrete color palette with 61 colors
# palette <- colorspace::qualitative_hcl(61, palette = "Dark 3")

# Plot the NMDS results using ggplot2
nmds_plot <- ggplot(nmds_scores, aes(x = NMDS1, y = NMDS2)) +
  geom_point(aes(
    color = Source,
    shape = Source, fill = Source), size = 3) +
  labs(#title = paste0("NMDS-", "Source"),
       x = "NMDS1", y = "NMDS2") +
  theme_minimal(base_size = 22) +
  theme(legend.position = "right") +
  # xlim(-1.25, 1.25) +
  # ylim(-1.5, 1.5) +
  scale_color_manual(name = "Source", values = c("firebrick3", "dodgerblue3")) #+
  # scale_shape_manual(name = "Source", values = shapes)  # +
  # ggforce::geom_mark_ellipse(aes(fill = Source, color = Source),
  #                            label.buffer = unit(20, 'mm'))

# Report Stress values
stressplot(nmds_result)

ggsave(filename = paste0("C:/Users/huyng/OneDrive - Toronto Metropolitan University (TMU)/Microplastic/Eric PMT weathering Paper/", "NMDS_eric_no-eclipse_", format(Sys.time(), "%d-%m-%Y"),".png"), 
       plot = nmds_plot, 
       device = "png",
       dpi = 600,        # Set DPI to 600
       width = 15,       # Width in inches (adjust as needed)
       height = 8,       # Height in inches (adjust as needed)
       units = "in")
```

### Step 11: PERMANOVA 
Reference: 
https://nathanielwoodward.com/posts/a-cool-thing-called-permanova/
https://onlinelibrary.wiley.com/doi/full/10.1002/9781118445112.stat07841

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Compound as predictor variables
permanova_source_compounddf <- percentage_normalized_df_cleaned %>% 
  dplyr::select(-c(Generalized_Product_name, Day, Product_Name))

permanova_source <- vegan::adonis2(vegan::vegdist(permanova_source_compounddf[,2:ncol(permanova_source_compounddf)], 
                                                  method = "bray")~Source, 
                                   permutations=5000, data = permanova_source_compounddf)


permanova_product_name_compounddf <- percentage_normalized_df_cleaned %>% 
  dplyr::select(-c(Source, Day, Product_Name))
permanova_product_name <- vegan::adonis2(vegan::vegdist(permanova_product_name_compounddf[,2:ncol(permanova_product_name_compounddf)], 
                                                        method = "bray")~Generalized_Product_name, 
                                         permutations=5000, data = permanova_product_name_compounddf)

# Compound+Polymer as predictor variables
permanova_source_CompoundPolymerdf <- percentage_normalized_df_cleaned %>% 
  dplyr::select(-c(Generalized_Product_name, Day, Product_Name))

permanova_source <- vegan::adonis2(vegan::vegdist(permanova_source_CompoundPolymerdf[,2:ncol(permanova_source_CompoundPolymerdf)], 
                                                  method = "bray")~Source, 
                                   permutations=5000, data = permanova_source_CompoundPolymerdf)


permanova_product_name_CompoundPolymerdf <- percentage_normalized_df_cleaned %>% 
  dplyr::select(-c(Source, Day, Product_Name))
permanova_product_name <- vegan::adonis2(vegan::vegdist(permanova_product_name_CompoundPolymerdf[,2:ncol(permanova_product_name_CompoundPolymerdf)], 
                                                        method = "bray")~Generalized_Product_name, 
                                         permutations=5000, data = permanova_product_name_CompoundPolymerdf)
```
